### 2016-01-19 inputs/abstract.tex
adding the latex folder

\begin{abstract}
Genotyping-by-sequencing (GBS) is a rapid and cost-effective genome-wide 
genotyping technique applicable whether a reference genome is available 
or not. Due to the cost-coverage trade-off, however, GBS typically produces 
large amounts of missing marker genotypes, whose imputation becomes therefore 
both challenging and critical to ensure reliability of later analyses.\\
In this work, the performance of four general imputation methods (K-Nearest 
Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
and one genotype-specific method (``Beagle'') were measured 
on GBS data from alfalfa (\emph{Medicago sativa}, autotetraploid, heterozygous, 
without reference genome) and 
rice (\emph{Oryza sativa}, with reference genome, 100\% homozygous). Benchmarks 
consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
proportions (up to 20\%) of known genotypes “masked” for imputation. The 
relative performance was measured as the total proportion of correctly imputed 
genotypes, globally and within each genotype class (two homozygotes in rice, 
two homozygotes and one heterozygote in alfalfa). Imputation efficiency was 
measured with computation times.\\
In general imputation accuracy was robust to increasing 
missing rates, and consistently higher in rice than in alfalfa. Accuracy 
was as high as 90-100\% for the major (most frequent) homozygous genotype, but dropped 
to 80-90\% (rice) and below 30\% (alfalfa) on the minor homozygous genotype. 
Beagle was the best performing method, both accuracy- 
and time-wise, when using a reference genome; when no reference 
genome was available, KNNI and RFI gave the highest accuracies, with KNNI 
being much faster.

\keywords{SNP, Genotyping by Sequencing (GBS), K-Nearest Neighbors Imputation 
(KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation 
(SVDI), Beagle, alfalfa, rice, imputation, reference genome}
\end{abstract}

### 2016-01-19 inputs/acknowledgements.tex
adding the latex folder

\begin{acknowledgements}
The creation of the rice data set was financially supported by the Marie Curie European
Reintegration Grant ``NEUTRADAPT''.
The creation of the alfalfa data sets was funded by the projects ‘Genomic selection in alfalfa (GENALFA)’ funded by the Italian Ministry of Foreign Affairs and International Cooperation in the framework of the Italy-USA scientific cooperation program, the Italian share of the FP7-ArimNet project ‘Resilient, water- and energy-efficient forage and feed crops for Mediterranean agricultural systems (REFORMA)’ funded by the Italian Ministry of Agricultural and Forestry Policies.
\end{acknowledgements}

### 2016-01-19 inputs/conclusions.tex
adding the latex folder

\section{Conclusions}
\label{sec:conclusions}
Imputation of missing genotypes can be an effective technique also for GBS data.
The most accurate imputation methods resulted in a total amount of wrongly imputed genotypes near to zero in rice and slightly over 10\% in alfalfa. 
The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa and, most importantly, affected the relative efficiency of the methods. While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome, such as in alfalfa. In particular, RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency.
The imputation accuracy tended to be relatively robust over increasing missing rates. However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''. In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\
Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.

### 2016-01-19 inputs/discussion.tex
adding the latex folder

\section{Discussion}
\label{sec:discussion}

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-chip data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they usually contain. On our data sets we found missing rates varying from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem tends to increase. Moreover, with such a high level of missingness, even imputation errors as low as 1\% may have negative impact on successive analysis, amplifying the importance of finding the most accurate imputation method.\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
%problem with url
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
%problem with url
\cite{he_learning_2009}.
The total classification accuracy can in fact be very high (close to 1) even with large error rates in the minority class. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
Some algorithms failed almost completely the imputation of the least common genotypes, while other algorithms proved to work well. In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes. MNI performed reasonably well in the alfalfa heterozygous class, with accuracy close to 0.6, contributed to its moderately good results (total accuracy around 0.8).

\subsection{Missing rate and the ``curse of dimensionality''}
\label{sec:missing_rate_curse_dimensionality}
In general, imputation methods were relatively robust to increasing missing rates. Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates. The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
KNNI's relatively pronounced drop in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed. 
On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$. With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases. This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighbourhood. RFI - which only partially relies on local structure of the data - suffered marginally from high missingness. SVD and MNI, which build on more general properties of the data, and Beagle - whose learning process is fundamentally different and specific for genotype imputation - were scarcely affected by missing rates.

\subsection{The role of the crop species: rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
%problems with this citation: url contains hashes
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to three main factors: 
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; 
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);
iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.

\subsection{Reference genome assembly and ordered vs unordered markers}
\label{sec:reference_genome_ordered_vs_unordered}
When a reference genome is available, markers positions can be used in the imputation process. All methods specifically developed for the imputation of missing genotypes have been designed for markers ordered along the genome sequence, and make explicit or implicit use of position information.
For alfalfathere is no reference genome sequenced yet. The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and could in principle be used. However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning the SNPs on a genome of a different species (and with a different ploidity) would introduce a further level of complexity, and the effect of alignments errors would be hardly assessable. Moreover, the non-aligning SNPs would be discarded, not on the basis of data quality, but of local synteny. We therefore decided to not use the genome of \emph{M. Truncatula} in the present analysis.\\
In rice, where a reference genome is available, we could assess the effect of using ordered vs. unordered markers for imputation. Among the imputation methods used in this study, Beagle is the only one to make use of marker information, and was therefore tested with ordered and randomly reshuffled markers. Interestingly, with ordered markers Beagle shows astounding resilience to high missingness, with imputation accuracy in the minority class $>$ 0.95 even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 0.75 (10\% allowed, 1\% artificial missing rates) to almost 0.5 (70\% allowed, 20\% artificial missing rates). When tested with alfalfa (with no marker position available) Beagle largerly overestimated by the majority class, performing poorly
This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.\\
Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provided interesting comparisons. In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\%to 80.9\% with 1\% to 20\% missing genotypes. Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range and finding Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI. 
Finally, Swarts et al.~\cite{swarts_novel_2014} used ordered GBS maize data to compare Beagle with the FSFHap and FILLIN algorithms they developed. All these algorithms rely on a Hidden Markov Model (HMM) to detect recombination break-points between haplotypes. They mostly obtained high imputation accuracies in the three genotype classes, providing further evidence of the added value of having a reference genome assembly.

\subsection{Size of the problem}
\label{sec:size_of_problem}
The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods. While problems related to the computational requirements can usually be tackled by the constant improvement in bioinformatics facilities, current trends in genotyping suggest that imputation methods and computational strategies well suited for large data sets are to be preferred. The non-linearity of computational efficiency generated, for most examined imputation algorithms, a total computation time for the complete rice data set which was greater than the sum of the computation times required by the individual chromosomes. Again, RFI provided the most blatant illustration of this behavior, with computation times for the complete rice data set going up to 937 hours (about 40 days), while the sum of the computation times for all the single chromosomes in the same experimental conditions added to 88.3 hours.

### 2016-01-19 inputs/introduction.tex
adding the latex folder

\section{Introduction}
\label{intro}

Imputation of missing alleles and genotypes is usually a preliminary step for
a wide range of genetic analyses: in fact, most models and software for 
population genetics, genomic selection (GS) and genome-wide association 
studies (GWAS) can not handle missing data and require complete datasets. 
SNP-chips represent one of the leading genotyping technologies, and produce 
datasets that, after markers and individuals with low call-rate are edited out, 
usually still contain a small proportion of uncalled genotypes (e.g., $<$5\%) 
randomly distributed along the genome. Genotyping-by-sequencing (GBS) is a 
relatively recent genotyping technique alternative to SNP-chips that can 
be used to produce SNP genotype data~\cite{elshire_robust_2011}.
GBS is platform-independent and is conveniently 
used in species that lack commercial SNP-chips or even lack a reference 
genome sequence. GBS data typically present a much larger proportion of 
sporadic missing genotypes: e.g., $\sim$52\% average in maize \cite{crossa_genomic_2013},
or up to 80\% for a fraction of SNPs in wheat \cite{poland_genomic_2012}. This 
is due to both intrinsic properties of the technology - e.g., low coverage - 
and to its application to species without a mature genome assembly \cite{glaubitz_tassel-gbs:_2014}.\\
There are several methods that are routinely used for the imputation of 
missing genotypes. Some rely solely on genotypic information, some make 
use also of genealogies, most of them are based on reconstructing haplotypes 
to be used in predictive models \cite{nicolazzi_software_2015}. Most methods for 
genotype imputation have been developed for SNP-chip data, and typically 
yield very high imputation accuracy: for instance, $>$95\% correctly imputed 
genotypes were reported in maize \cite{hickey_factors_2012} and cattle \cite{vanraden_genomic_2011}. 
Such methods can be applied to GBS data, too (see for instance 
\cite{huang_efficient_2014} and \cite{swarts_novel_2014} 
for applications in rice and maize). When a reference 
genome is available, SNP loci can be aligned against its sequence and 
are therefore ordered. However, GBS data may be generated 
also for species lacking a reference genome assembly. In this case the 
GBS output is a series of ``floating'' loci not linked to any chromosome: 
alignment is not possible and SNPs are therefore unordered. Unordered 
data add additional complexity to the task of imputing missing genotypes: 
exploitation of linkage disequilibrium (LD) and haplotype reconstruction 
are less straightforward, consecutive loci may lie on different chromosomes
or scaffolds, and overall data are less homogeneous. Rutkoski et al.~\cite{rutkoski_imputation_2013}
considered the application of general data imputation methods 
to wheat, maize and barley populations, obtaining best case scenarios accuracies 
as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).\\
Compared to SNP-chip data, GBS data are more challenging: many more 
missing genotypes, high variability between runs, intrinsically noisy data, 
reads from different loci that can overlap. For all these reasons, imputation 
of missing genotypes with GBS data is still a relatively immature technique, 
not yet amenable to produce highly standardised and repeatable results. 
There is therefore scientific and practical interest in gaining further 
insights into how genotype imputation with GBS data works, and in 
developing the best methods and strategies to accurately and efficiently 
impute such missing data. This would be relevant not only for the scientific 
community but also for the breeding industry, because of its direct 
consequences on genomic applications.\\
In this paper, we imputed missing genotypes from GBS data in two agronomically
important crop species: the forage crop alfalfa (\emph{Medicago sativa}), for 
which a reference genome is not available yet, and the 
cereal crop rice (\emph{Oryza sativa}) whose reference genome has been 
assembled \cite{international_rice_genome_sequencing_project_map-based_2005} and
recently updated \cite{kawahara_improvement_2013}. We compared the imputation 
efficiency of four general data-imputation methods (Mean Value Imputation, 
K-Nearest Neighbour Imputation, Singular Value Decomposition Imputation 
and Random Forest Imputation) with the performance of a method specific 
for genotype imputation (localized haplotype clustering imputation, 
implemented in the software package Beagle). The relative efficiency of 
the different imputation methods was assessed in terms of the 
accuracy achieved and the computation time required.\\

### 2016-01-19 inputs/materials_and_methods.tex
adding the latex folder

\section{Materials and methods}

\subsection{Plant material}
\label{sec:plant_material}

Samples from two autotetraploid alfalfa (\emph{Medicago sativa L.}) and one diploid rice (\emph{Oryza sativa L.}) populations were available (see Table~\ref{tab:descriptive_statistics}).\\
Alfalfa populations included elite germplasm from the Po Valley (Alfalfa-PV, \cite{annicchiarico_alfalfa_2015}) and elite germplasm adapted to Mediterranean-climate environments (Alfalfa-Med, \cite{annicchiarico_adaptation_2011}). For Alfalfa-PV, 124 parent genotypes were chosen by stratified mass selection for dry matter yield over three harvests among 4,480 densely planted genotypes grown in Lodi (northern Italy) under field conditions in 2002 and 2003. The Alfalfa-Med population comprised a sample of 154 parent genotypes derived from two cycles of free intercrossing among three outstanding populations in a multi-environment study: (i) the drought-tolerant Sardinian landrace Mamuntanas; (ii) the salt-tolerant Moroccan landrace Erfoud 1; (iii) the Australian variety SARDI 10, widely adapted across moisture-favorable and drought-prone environments. \\
The Rice dataset included 437 plants belonging to 391 accessions from the Rice Germplasm Collection maintained at CRA-Rice Research Unit (Vercelli, Italy). The sampled collection comprised accessions from the five main sub-populations of \emph{O. sativa} (274 temperate japonica, 108 tropical japonica, 28 indica, 16 aus and 11 aromatic). All accessions were purified through single seed descent, and were genotypically essentially 100\% homozygous.

\subsection{Genotyping by sequencing}
\label{sec:overview}

The 715 samples from alfalfa (Alfalfa-Med and Alfalfa-PV) and rice (Rice) populations were genotyped using the genotyping-by-sequencing (GBS) approach \cite{elshire_robust_2011}. Different GBS protocols were used to genotype the alfalfa and rice samples, since genotyping was carried out under different projects.\\
In alfalfa, DNA was isolated from fresh leaf tissues by the Wizard~\textregistered~Genomic DNA Purification Kit (Promega, A1125) and quantified with a Quant-iT PicoGreen dsDNA assay kit (Life Technologies, P7589). Two libraries were constructed for the two populations, where 100 ng of each DNA was digested with ApeKI (NEB, R0643L) and then ligated to a unique barcoded adapter and a common adapter. For the reference population Alfalfa-Med, 5 nmoles each of the primers and NEB 2X Taq Master Mix (NEB Cat \# M0270S) were included in the PCR reaction according to \cite{elshire_robust_2011} original protocol, while for Alfalfa-PV the KAPA library amplification readymix (Kapa Biosystems Cat \# KK2611) was used instead. Each library was sequenced in two lanes on Illumina HiSeq 2000 at the Genomic Sequencing and Analysis Facility at the University of Texas at Austin, Texas, USA. For both populations post sequencing analysis and SNP calling was carried out using Tassel UNEAK pipeline \cite{lu_switchgrass_2013}. A further quality filter, implemented through ad-hoc Python scripts, removed heterozygous loci with less than 4 and homozygous loci with less than 11 aligned reads. Alfalfa is an autotetraploid plant species and can therefore present three different heterozygous genotypes: AAAB, AABB and ABBB. Following \cite{li_saturated_2014} and \cite{li_genomic_2015}, these were all considered diploid heterozygotes (i.e. AB), while the two homozygous genotypes (AAAA and BBBB) were considered diploid homozygotes (i.e., AA or BB).\\
In rice, DNA was isolated from three-weeks old leaves using the DNeasy Plant Mini Kit (QIAGEN) with a TECAN Freedom EVO150 liquid handling robot (TECAN Group Ltd, Switzerland). DNA digestion was performed on 100 ng samples in 96-well plates using ApeKI, which was shown to cut every 1 kb on average in a in-silico digestion of the Nipponbare reference genome, and 96-plex libraries constructed according to the GBS protocol. The libraries were loaded into Genome Analyzer II (Illumina, Inc., San Diego, CA) for sequencing. Raw sequence data filtering, sequence alignment to the rice reference genome (Os-Nipponbare-Reference-IRGSP-1.0, \cite{kawahara_improvement_2013}), and SNP calling from GBS genotyping were carried out using the Tassel GBS pipeline \cite{glaubitz_tassel-gbs:_2014}.\\
In total, 32\,706, 40\,734 and 166\,418 SNP markers were called by GBS in Alfalfa-PV, Alfalfa-Med and Rice, respectively (see Table~\ref{tab:descriptive_statistics}). In both datasets SNP variants were renamed so that AA represented the most common homozygote, and BB the least common homozygote.

\input{inputs/table_descriptive_statistics}

\subsection{Imputation methods}
\label{sec:imputation_methods}
We considered five imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), and Localized Haplotype Clustering Imputation (``Beagle''). For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively. 
\paragraph{MNI}
\label{par:MNI}
: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, which was then discretized to the closest value in \{0,1,2\}. 
\paragraph{KNNI}
\label{par:KNNI}
: in K-Nearest Neighbors Imputation, missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001}. This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data. K=4 was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
\paragraph{SVDI}
\label{par:SVDI}
: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (m markers, n individuals):

\begin{equation}
\label{eq:SVDI_general}
M = U\Sigma V^{T}
\end{equation}

where U is a $m \times m$ unitary matrix (i.e. $UU^{T}=I$), $\Sigma$ is a $m \times n$ rectangular diagonal matrix of singular values and $V^{T}$ is the $n \times n$ conjugate transpose of the unitary matrix V. Matrix U and V contain the eigenvectors of $M^{T}M$ and $MM^{T}$, respectively, and the corresponding non-zero singular values in $\Sigma$ are equivalent to the square-root of the eigenvalues of $M^{T}M$ and $MM^{T}$. The first k eigenvectors - ordered by decreasing eigenvalue - capture most of the information in the marker genotype matrix, and were used to generate linear combinations (principal components) of the original m markers for the imputation of missing data points. The imputation procedure comprised: (i) initial imputation of missing genotypes using MNI, since SVD can only be performed on complete matrices; (ii) SVD to select the most informative k eigenvectors of the marker genotype matrix; (iii) these k eigenvectors were used as predictors in a linear regression model for marker genotypes:

\begin{equation}
\label{eq:SVDI_regression}
Y = \hat{U}\beta +\varepsilon 
\end{equation}

with Y as vector of genotypes at marker j; $\hat{U}$ matrix of k eigenvectors; $\beta$ vector of regression coefficients; $\varepsilon$ random errors terms; (iv) all eigenvectors (matrix U) and $\beta$ were used to estimate missing values at marker j. The procedure was repeated (steps (ii) to (iv)) until convergence. The final imputed data points were then discretized to the closer genotype in \{0,1,2\}. Additional details on SVDI can be found in \cite{troyanskaya_missing_2001}

\paragraph{RFI}
\label{par:RFI}
: in Random Forest (RF) imputation, missing genotypes at marker j were imputed by means of RF multiple regression trees \cite{breiman_random_2001} where all markers other than j were used for the prediction. At each marker j, 100 RF regression trees $\Theta_{1} \ldots \Theta_{100}$ were grown from a bootstrapped sample of the individuals in Y and a random subset of $\sqrt{m-1}$ markers. Missing genotypes were imputed averaging predictions over the 100 RF trees

\begin{equation}
\label{eq:RFI_regression}
\hat{Y} = \frac{1}{100}\sum_{i=1}^{100} h(x,\Theta_i)
\end{equation}

where $h(x,\Theta_i)$ is a function of the $i_{th}$ tree and subset of markers. RFI was repeated until convergence or for maximum 10 iterations. After regression, the imputed data were then discretized to the closer genotype in \{0,1,2\}.

\paragraph{Beagle}
\label{par:Beagle}
: ``localized haplotype clustering imputation'' is a method implemented in the software ``Beagle'' \cite{browning_rapid_2007}. Originally developed for human genetics, Beagle has since found wide application also in animal and plant genetics. Beagle has become so popular that the name of the software is commonly used  to refer metonymically to the method it implements, making the two hardly distinguishable. Beagle infers haplotypes and imputes sporadic missing alleles both with known and unknown phase, using a localized haplotype cluster model; this is a class of directed acyclic graphs which empirically models haplotype frequencies on a local scale and therefore adapts to local structure in the data. Beagle makes use of a hidden Markov model to find the most likely haplotype pair for each individual, given the genotype data for that individual and the graphical haplotype frequency model. The method works iteratively using an expectation-maximization (EM) approach. The imputed missing data, probabilities of missing genotypes and inferred haplotypes are calculated from the model that is fitted in the last iteration. For the imputation of missing genotypes in the completely homozygous rice accessions, a likelihood file was defined to specifically exclude the imputation of non-possible heterozygous genotypes.

\subsection{Imputation procedure}
\label{sec:imputation_procedure}
In order to assess the imputation performance of the different methods, we introduced increasing proportions of artificial missing genotypes in the data sets. These were then imputed with the various algorithms, measuring the resulting imputation accuracy and computation time. Unlike imputation experiments with SNP-chips (e.g., \cite{hickey_factors_2012}), GBS data intrinsically present a much larger proportion of missing genotypes, and a certain amount of naturally missing genotypes has to be allowed along with artificially introduced ones (see for instance \cite{rutkoski_imputation_2013}). We extracted four datasets from the overall data that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker. For each of these datasets, we randomly introduced an additional 1\%, 5\%, 10\% or 20\% missing genotypes, on which imputation accuracy could be measured. Table~\ref{tab:missing_rates} reports the number of markers and average missing rate (before introducing artificial missing genotypes) for the four call-rate thresholds. This procedure produced 16 combinations for each population: 4 call rate datasets $\times$ 4 levels of artificial missing markers. On each combination, we applied the 5 described imputation methods. Rice data were analysed as a whole and by each of the 12 chromosomes separately. \\
For rice data only, the position of the SNPs on the genome was known. It was therefore possible to test the imputation performance with and without this information. Among the proposed imputation methods, only Beagle can use SNP position in the imputation process. We therefore applied Beagle both to SNP ordered per chromosome and along the entire genome and to SNP that were randomly reshuffled (so as to break the linkage between them). \\
All of the above led to a total of 1408 (1200 + 208 for reshuffled rice data with Beagle) imputation analyses.

\input{inputs/table_missing_rates}

\subsection{Imputation accuracy and efficiency assessment}
\label{sec:imputation_accuracy_and_efficiency_assessment}
We used the artificial missing genotypes to measure the performance of the five imputation methods. The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele). Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class. The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of missing data (proportion of correct imputations):

\begin{equation}
\label{eq:accuracy}
accuracy = \frac{1}{n} \sum_{i=1}^{n}{I(y_i = \hat{y_i})}
\end{equation}

where $I()$ is an indicator function that returns 1 if the imputed ($\hat{y}$) and true ($y$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa (three genotype classes plus total accuracy) and three for rice (two genotype classes plus total accuracy).\\
Imputation methods were also compared in terms of computation efficiency, measured as the total computation time needed to complete the analyses. The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution. In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.

\subsection{Software}
\label{sec:software}
Data handling, editing and preparation, summary of results and plots were all performed using the open-source environment for statistical programming R \cite{r_core_team_r:_2014}. All imputation methods, except Beagle, were implemented in R: MNI directly in base R; KNNI using the function \emph{knncatimpute()} from the Scrime package \cite{schwender_scrime:_2013}; SVDI with the R package ``bcv'' \cite{perry_bcv:_2009}; RFI using the ``MissForest'' \cite{stekhoven_missforestnon-parametric_2012} R package (with parameters: ntree=100, maxiter=10, parallelize='variables'). The ``Beagle'' localized haplotype clustering imputation method was implemented with the Beagle software version 3.3.2 \cite{browning_rapid_2007}.

### 2016-01-19 inputs/results.tex
adding the latex folder

\section{Results}
\label{sec:results}

\subsection{Genotypes}
\label{sec:genotypes}  
GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}). The overall missing rate was 0.666, 0.596 and 0.534 in the three datasets. Supplementary Figure S1 shows the density distribution of missing rate per markers and per individuals in the complete datasets. The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%). Missing data point counts ranged from a minimum of 3\,878 in Alfalfa-PV with 10\% allowed and 1\% artificial missing genotypes, to a maximum of 5\,580\,616 in Rice with 70\% allowed and 20\% artificial missing genotypes. Over all datasets and allowed/artificial missing genotypes thresholds, the amount of missing data points to be imputed averaged 249\,405.\\
Minor Allele Frequency (MAF) was 0.172, 0.170 and 0.140 in Alfalfa-PV, Alfalfa-Med and Rice, respectively. The three datasets were therefore unbalanced with respect to genotype classes: Rice had the lowest MAF, while Alfalfa-PV and Alfalfa-Med had the smallest minor classes (minor homozygotes 3.5\% and 3.2\% of the total genotypes). Figure~\ref{fig:genotype_classes} reports the proportion of genotypes in each class in the complete datasets (100\%) and as a function of the maximum allowed missing rate per marker. The relative proportion of genotype classes remained approximately stable in rice. In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes. This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous loci. The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.

\begin{figure}
\includegraphics[width=0.95\textwidth]{Fig01_class_breakdown_all.pdf}
\caption[Proportions of genotype classes]{​Proportion of genotype classes for the different missing rate thresholds (0.1, 0.2, 0.4, 0.7 and 1 -the complete dataset), in Alfalfa-PV, Alfalfa-Med and Rice (all chromosomes and chromosome 1 only). Red bars represents the most common homozygous (AA), blue bars the least common homozygous (BB) and green bars the heterozygous (AB).}
\label{fig:genotype_classes}
\end{figure}

\subsection{Imputation accuracy}
\label{sec:imputation_accuracy}

In rice, the average imputation accuracy over all genotypic classes was close to 100\% with most methods and still above 90\% at maximum proportions of allowed (70\%) and artificial (20\%) missing genotypes (Figure~\ref{fig:accuracy_rice}). The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\% (Figure~\ref{fig:accuracy_alfalfa}). The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 90.8\% and 99.3\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 4.1\% and 73.4\% in alfalfa and rice, respectively). Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}). Heterozygotes had an intermediate imputation accuracy ranging from 0\% with MNI to 70\% with KNNI. Results for each individual alfalfa population, the 12 individual rice chromosomes, and the whole rice dataset are similar to those discussed here (see Supplementary Figures S2 to S16). \\
Missing data did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate. KNNI showed a decreasing imputation accuracy with increasing missing data, both in alfalfa and rice.\\
In rice, Beagle outperformed all other imputation methods, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}). When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle. In general we observed for all thresholds and in both genotype classes the following ranking of methods for imputation accuracy: RFI $>$ KNNI $>$ SVDI.
In particular, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI. As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
In alfalfa KNNI showed the highest imputation accuracy over all classes and in the heterozygous and minor homozygous genotype classes (Figure~\ref{fig:accuracy_alfalfa}). RFI and SVDI were second ranking. They did not differ much from MNI in the overall accuracy and in the major homozygous class, but gave higher imputation accuracy in the heterozygous and minor homozygous classes. Beagle was also applied to alfalfa, where the relative position of markers in the genome is not known. The overall imputation accuracy with Beagle was about 67\%, with accuracies of almost 100\% in the major homozygous class but practically 0\% in the heterozygous and minor homozygous classes.

\begin{figure}
\includegraphics[width=0.95\textwidth]{Fig02_Rice-chrom_avg.pdf}
\caption[Rice imputation accuracies]{
imputation accuracies overall, for the major homozygous genotype (AA), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged over the 12 rice chromosomes. Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black).}
\label{fig:accuracy_rice}
\end{figure}

\begin{figure}
\includegraphics[width=0.95\textwidth]{Fig03_Alfalfa_avg.pdf}
\caption[Alfalfa imputation accuracies]{
imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged two alfalfa populations (Alfalfa-Med and Alfalfa-PV). Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black).}
\label{fig:accuracy_alfalfa}
\end{figure}

\subsection{Computation time}
\label{sec:computation_time}  
We measured the efficiency of the different imputation algorithms as the amount of time required to complete the imputation process. Only the implementation of RFI could leverage a multi-core/multi-thread environment, so that RFI computation times should be evaluated considering 10 CPUs used in parallel, while all other algorithms used only one CPU at a time.\\
Imputation efficiency (Figure~\ref{fig:time_analysis}) was assessed with respect to the gross dimension of the data set (i.e., number of markers $\times$ number of samples) and to the total number of missing genotypes to be imputed (e.g., 341\,097 missing genotypes for the Alfalfa-Med population with 40\% missing rate and 20\% artificial missing).
RFI required by far the longest computation times (in spite of 10-core/thread parallelization), which grew approximately exponentially ($O(e^N)$) with problem size, and easily required tens of hours for individual rice chromosomes, and hundred of hours (up to 937 hours) for the complete rice dataset. The second slowest algorithm was KNNI, with computation times growing approximately quadratically ($O(N^2)$) with problem complexity (the chosen KNNI implementation contains no heuristic method to prune the all vs. all distance calculation). KNNI was however significantly faster than RFI (on average 16.7 times faster), requiring a time ranging from 20.68 seconds (rice chromosome 7, 10\% allowed, 1\% artificial) to about 28 hours (rice complete dataset, 70\% allowed, 20\% artificial) to complete the imputation task.\\
All other algorithms were faster, with computation times growing linearly or logarithmically with problem size. Beagle and SVDI resulted in similar execution times. MNI was by far the fastest imputation algorithm, being scarcely affected by the size of the problem (computation times ranging from 0.23 to 31.29 seconds).

\begin{figure}
\includegraphics[width=0.95\textwidth]{Fig04_time_analysis.pdf}
\caption[Computation times]{computation time as a function of the amount of the total size of the imputed dataset (left) and of missing data (right) for the five imputation methods (MNI: orange, KNNI: red, SVDI: blue, RFI: green, Beagle:black). Plots include measures on Alfalfa-Med population, Alfalfa-PV population, and rice chromosomes 1 to 12. Complete rice datasets are omitted for readibility.}
\label{fig:time_analysis}
\end{figure}

### 2016-01-19 inputs/supplementary.tex
adding the latex folder

\section{Supplementary material}
\label{sec:supplementary_material}

\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig01_miss_per_marker.pdf}
\caption{Distribution of missings-per-marker rates in the three datasets. This is a
density plot, so all curves delimit the same unitary area.}
\end{figure}

\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig02_Alfalfa-Med.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for alfalfa population Alfalfa-Med.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig03_Alfalfa-PV.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for alfalfa population Alfalfa-PV.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig04_Rice-chrom-1.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 1 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig05_Rice-chrom-2.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 2 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig06_Rice-chrom-3.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 3 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig07_Rice-chrom-4.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 4 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig08_Rice-chrom-5.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 5 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig09_Rice-chrom-6.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 6 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig10_Rice-chrom-7.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 7 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig11_Rice-chrom-8.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 8 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig12_Rice-chrom-9.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 9 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig13_Rice-chrom-10.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 10 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig14_Rice-chrom-11.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 11 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig15_Rice-chrom-12.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for rice chromosome 12 data.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig16_Rice.pdf}\caption{
imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\%
additional missing values artificially introduced (x-axis) for whole rice data set.
Lines colors represent the five imputation algorithms: MNI
(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}

### 2016-01-19 inputs/table_descriptive_statistics.tex
adding the latex folder

\begin{table}
\centering
\caption[Descriptive statistics]{
Descriptive statistics of alfalfa (\emph{Medicago sativa}) 
and rice (\emph{Oryza sativa}) genotyping. MAF is average minor allele frequency; 
p(AA), p(AB) and p(BB) are the proportions of AA, AB and BB genotypes.}
\label{tab:descriptive_statistics}
\begin{tabular}{lccccccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Dataset} & \textbf{No. samples} & \textbf{No. SNPs} & \textbf{Missing Rate} & \textbf{MAF} & \textbf{p(AA)} & \textbf{p(AB)} & \textbf{p(BB)}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
Alfalfa-PV & 124 & 32\,706 & 0.666 & 0.1724 & 0.690 & 0.274 & 0.035\\
Alfalfa-Med & 154 & 40\,734 & 0.596 & 0.1702 & 0.691 & 0.276 & 0.032\\
Rice & 437 & 166\,418 & 0.534 & 0.140 & 0.860 & - & 0.140 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

### 2016-01-19 redo_pdf.sh
files needed to compile latex

@@ -1,0 +1,24 @@
+#!/bin/bash
+#DOING MAIN
+FILE=main
+
+#cleaning all LaTeX clutter
+rm -f $FILE.aux $FILE.dvi $FILE.log $FILE.idx $FILE.ind $FILE.toc $FILE.ilg $FILE.out $FILE.bbl $FILE.blg $FILE.nav $FILE.snm $FILE.ps $FILE.pdf
+
+#doing the usual latex routine
+pdflatex $FILE
+bibtex $FILE
+pdflatex $FILE
+pdflatex $FILE
+
+#DOING SUPPLEMENTARY
+FILE=suppl
+
+#cleaning all LaTeX clutter
+rm -f $FILE.aux $FILE.dvi $FILE.log $FILE.idx $FILE.ind $FILE.toc $FILE.ilg $FILE.out $FILE.bbl $FILE.blg $FILE.nav $FILE.snm $FILE.ps $FILE.pdf
+
+#doing the usual latex routine
+pdflatex $FILE
+bibtex $FILE
+pdflatex $FILE
+pdflatex $FILE

### 2016-01-25 inputs/materials_and_methods.tex
more concise material explanation, added a reference

@@
 Samples from two autotetraploid alfalfa (\emph{Medicago sativa L.}) and one diploid rice (\emph{Oryza sativa L.}) populations were available (see Table~\ref{tab:descriptive_statistics}).\\
-Alfalfa populations included elite germplasm from the Po Valley (Alfalfa-PV, \cite{annicchiarico_alfalfa_2015}) and elite germplasm adapted to Mediterranean-climate environments (Alfalfa-Med, \cite{annicchiarico_adaptation_2011}).
-For Alfalfa-PV, 124 parent genotypes were chosen by stratified mass selection for dry matter yield over three harvests among 4,480 densely planted genotypes grown in Lodi (northern Italy) under field conditions in 2002 and 2003.
-The Alfalfa-Med population comprised a sample of 154 parent genotypes derived from two cycles of free intercrossing among three outstanding populations in a multi-environment study: (i) the drought-tolerant Sardinian landrace Mamuntanas; (ii) the salt-tolerant Moroccan landrace Erfoud 1; (iii) the Australian variety SARDI 10, widely adapted across moisture-favorable and drought-prone environments.
+Alfalfa populations included elite germplasms from the Po Valley (Alfalfa-PV) and Mediterranean-climate environments (Alfalfa-Med).
+For Alfalfa-PV, 124 parent genotypes were chosen by stratified mass selection for dry matter yield over three harvests.
+For Alfalfa-Med, 154 parent genotypes derived from two cycles of free intercrossing among three outstanding populations in a multi-environment study.
+For further details see \cite{Annicchiarico2015}.
 \\

### 2016-01-25 inputs/introduction.tex
Modified introduction

@@
 K-Nearest Neighbour Imputation, Singular Value Decomposition Imputation 
-and Random Forest Imputation) with the performance of a method specific 
+and Random Forest Imputation) with the performance of two methods specific 
 for genotype imputation (localized haplotype clustering imputation, 
-implemented in the software package Beagle). The relative efficiency of 
+implemented in the software package Beagle, %and haplotype reconstruction around the recombination site
+and the FILLIN algorithm implemented in the programming suite Tassel). The relative efficiency of 
 the different imputation methods was assessed in terms of the 
-accuracy achieved and the computation time required.\\
+accuracy achieved and the computation time required.
+Not only the total imputation accuracy was measured, but also the accuracy of imputing the different genotype classes (major and minor homozygotes, heterozygotes).\\

### 2016-01-25 inputs/introduction.tex
intro-14:54

@@
 population genetics, genomic selection (GS) and genome-wide association 
-studies (GWAS) can not handle missing data and require complete datasets. 
+studies (GWAS) can not handle missing data and require complete datasets (e.g. \cite{hayes2009invited}). 
 SNP-chips represent one of the leading genotyping technologies, and produce 
@@
 to wheat, maize and barley populations, obtaining best case scenarios accuracies 
-as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).\\
+as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).
+Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward2013}).\\
 Compared to SNP-chip data, GBS data are more challenging: many more 

### 2016-01-25 inputs/materials_and_methods.tex
mainly supplementary, plus small changes all over

@@
 For both populations post sequencing analysis and SNP calling was carried out using Tassel UNEAK pipeline \cite{lu_switchgrass_2013}.
+
 A further quality filter, implemented through ad-hoc Python scripts, removed heterozygous loci with less than 4 and homozygous loci with less than 11 aligned reads.
+A similar filtering was performed in \cite{Rocher_validation_2015} using less restrictive thresholds.
+
+.
 Alfalfa is an autotetraploid plant species and can therefore present three different heterozygous genotypes: AAAB, AABB and ABBB.
@@
 AB), while the two homozygous genotypes (AAAA and BBBB) were considered diploid homozygotes (i.e., AA or BB).\\
+
 In rice, DNA was isolated from three-weeks old leaves using the DNeasy Plant Mini Kit (QIAGEN) with a TECAN Freedom EVO150 liquid handling robot (TECAN Group Ltd, Switzerland).

### 2016-01-25 inputs/supplementary.tex
other references
mainly supplementary, plus small changes all over

@@
 \label{sec:supplementary_material}
+
+\subsection{SNP calling thresholds}
+\label{sec:SNP_calling_thresholds}
+Genotype calling can incur in two types of error, namely a heterozygote wrongly identified as homozigote ($HET \rightarrow HOM$) and the reverse case ($HOM \rightarrow HET$). 
+Under the simplifying hypothesis of absence of sequencing errors and homogeneous sequencing quality, it is possible to estimate error rates for a tetraploid genome using the following formula:
+
+\begin{equation}
+\begin{cases}
+0.75^{N_h} + 0.25^{N_h} & HET_{3-1} \rightarrow HOM \\ 
+2 \times 0.5^{N_h}      & HET_{2-2} \rightarrow HOM \\ 
+0                       & HOM \rightarrow HET
+\end{cases}
+\end{equation}
+
+Where $N_h$ is the minimum number of reads required to label a stack of aligning reads as heterozygote.
+The discussed case of $N_h = 11$ brings to a total expected $HET \rightarrow HOM$ error rate of 4.22\%.\\
+The hypothesis mentioned above greatly simplyfies the real case (e.g.
+$HOM \rightarrow HET$ errors can not happen, since a homozigote will always produce 100\% of reads of the same type).
+When accounting sequencing errors and reads quality a term for $HOM \rightarrow HET$ appears, due to reads sequences as the wrong allele.
+In the presented study we thus applied a threshold of 4 required reads per stack to call a homozigote to minimize this kind of error.
+
+\subsection{Supplementary figures}
+\label{sec:supplementary_figures}
 

### 2016-01-25 inputs/materials_and_methods.tex
tetraploid explained

@@
 A similar filtering was performed in \cite{Rocher_validation_2015} using less restrictive thresholds.
-
-.
 Alfalfa is an autotetraploid plant species and can therefore present three different heterozygous genotypes: AAAB, AABB and ABBB.
-Following \cite{li_saturated_2014} and \cite{li_genomic_2015}, these were all considered diploid heterozygotes (i.e.
-AB), while the two homozygous genotypes (AAAA and BBBB) were considered diploid homozygotes (i.e., AA or BB).\\
+Sequencing depth in this study was not sufficient for accurate tetraploid allelic dosage, but following \cite{li_saturated_2014} and \cite{li_genomic_2015} reliable genotype calls based on diploid allelic dosage were obtained considering diploid heterozygotes (i.e.
+AB), while the two homozygous genotypes (AAAA and BBBB) were considered diploid homozygotes (i.e., AA or BB).
 

### 2016-01-25 inputs/introduction.tex
intro-16:45

@@
 randomly distributed along the genome. Genotyping-by-sequencing (GBS) is a 
-relatively recent genotyping technique alternative to SNP-chips that can 
-be used to produce SNP genotype data~\cite{elshire_robust_2011}.
+relatively recent technique which is considered a viable alternative to SNP array-based genotyping to produce SNP genotype data~\cite{elshire_robust_2011}.
 GBS is platform-independent and is conveniently 
@@
 genome sequence. GBS data typically present a much larger proportion of 
-sporadic missing genotypes: e.g., $\sim$52\% average in maize \cite{crossa_genomic_2013},
-or up to 80\% for a fraction of SNPs in wheat \cite{poland_genomic_2012}. This 
+sporadic missing genotypes: e.g., $\sim$52\% and $\sim 58\%$ average in two maize experimental populations \cite{crossa_genomic_2013},
+or ``up to 80\% missing data per marker'' in wheat \cite{poland_genomic_2012}. This 
 is due to both intrinsic properties of the technology - e.g., low coverage - 

### 2016-01-25 inputs/introduction.tex
einleitung

@@
 to be used in predictive models \cite{nicolazzi_software_2015}. Most methods for 
-genotype imputation have been developed for SNP-chip data, and typically 
+genotype imputation have been applied to SNP array data, and typically 
 yield very high imputation accuracy: for instance, $>$95\% correctly imputed 
@@
 cereal crop rice (\emph{Oryza sativa}) whose reference genome has been 
-assembled \cite{international_rice_genome_sequencing_project_map-based_2005} and
-recently updated \cite{kawahara_improvement_2013}. We compared the imputation 
+assembled in 2005 \cite{international_rice_genome_sequencing_project_map-based_2005}, and
+was recently updated \cite{kawahara_improvement_2013}. We compared the imputation 
 efficiency of four general data-imputation methods (Mean Value Imputation, 

### 2016-01-25 inputs/discussion.tex
nelson action points

@@
 \cite{he_learning_2009}.
-The total classification accuracy can in fact be very high (close to 1) even with large error rates in the minority class. 
+Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
 The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans).
@@
 
-\subsection{The role of the crop species: rice vs. alfalfa}
+\subsection{Role of crop species: rice vs. alfalfa}
 \label{sec:role_of_species}
@@
 \cite{the_1000_genomes_project_consortium_integrated_2012}
-for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to three main factors: 
-i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; 
-ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);
-iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
+for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: 
+i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
+ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);\\
+iii) alfalfa datasets did not contain population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulation.
+Population structure is implicitly used by algorithms as Beagle and KNNI that create clusters of similar genotypes.
+Thus, rice dataset population structure facilitated the inputation process;\\
+iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
 Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.
@@
 
-\subsection{Size of the problem}
+\subsection{Problem size}
 \label{sec:size_of_problem}
-The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources.
+Problem size, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources.
 RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed.

### 2016-01-25 inputs/discussion.tex
aenderungen-19:55

--- 2016-01-19 inputs/discussion.tex
@@
 
-\subsection{The role of the crop species: rice vs. alfalfa}
+\subsection{Imputation efficiency differences in rice vs. alfalfa}
 \label{sec:role_of_species}
@@
 
-\subsection{Size of the problem}
+\subsection{Size of the imputation problem}
 \label{sec:size_of_problem}
+Albeit this was not the main objective of the paper, and no formal effort to optimise the implementation of the various imputation methods was done, still the recorded computation times give us some interesting information.
 The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources.

### 2016-01-25 inputs/introduction.tex
aenderungen-19:55

@@
 as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).
-Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward2013}).\\
+Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward_2013_saturated}).\\
 Compared to SNP-chip data, GBS data are more challenging: many more 

### 2016-01-25 inputs/materials_and_methods.tex
aenderungen-19:55

@@
 \label{sec:imputation_methods}
-We considered five imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), and Localized Haplotype Clustering Imputation (``Beagle'').
+We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite \cite{swarts_novel_2014}.
 For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.
@@
 This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data.
-K=4 was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
+$K=4$ was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
 \paragraph{SVDI}
@@
 \label{eq:SVDI_regression}
-Y = \hat{U}\beta +\varepsilon 
+Y = U\beta +\varepsilon 
 \end{equation}
@@
 The imputed missing data, probabilities of missing genotypes and inferred haplotypes are calculated from the model that is fitted in the last iteration.
-For the imputation of missing genotypes in the completely homozygous rice accessions, a likelihood file was defined to specifically exclude the imputation of non-possible heterozygous genotypes.
+For the imputation of missing genotypes in the completely homozygous rice accessions, a likelihood file (with the prior likelihood of each of the three possible SNP genotypes) was defined to specifically exclude the imputation of non-possible heterozygous genotypes (details in \cite{beagle3_manual}).
+
+\paragraph{FILLIN}
+The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}.
+``FILLIN'' is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function.
+This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.
 
@@
 For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class.
-The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of missing data (proportion of correct imputations):
+The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of artificially missing data (proportion of correct imputations):
 
@@
 \label{eq:accuracy}
-accuracy = \frac{1}{n} \sum_{i=1}^{n}{I(y_i = \hat{y_i})}
+accuracy = \frac{1}{n} \sum_{i=1}^{n}{I(g_i = \hat{g_i})}
 \end{equation}
 
-where $I()$ is an indicator function that returns 1 if the imputed ($\hat{y}$) and true ($y$) genotypes are equal, 0 otherwise.
+where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise.
 We obtained four accuracy measures for alfalfa (three genotype classes plus total accuracy) and three for rice (two genotype classes plus total accuracy).\\
-Imputation methods were also compared in terms of computation efficiency, measured as the total computation time needed to complete the analyses.
+For each imputation method, also the total computation time needed to complete the analyses was measured, in order to get an idea of their relative performance.
 The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution.

### 2016-01-25 inputs/results.tex
nelson action points
aenderungen-19:55

@@
 The overall missing rate was 0.666, 0.596 and 0.534 in the three datasets.
-Supplementary Figure S1 shows the density distribution of missing rate per markers and per individuals in the complete datasets.
+Supplementary Figure S1 shows the density distribution of missing rate per markers in the complete datasets.
 The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%).
@@
 \label{sec:computation_time}  
-We measured the efficiency of the different imputation algorithms as the amount of time required to complete the imputation process.
+The amount of time required to complete the imputation process was recorded for each method.
 Only the implementation of RFI could leverage a multi-core/multi-thread environment, so that RFI computation times should be evaluated considering 10 CPUs used in parallel, while all other algorithms used only one CPU at a time.\\

### 2016-01-26 inputs/discussion.tex
modifiche 26 mattina 10:36

--- 2016-01-25 inputs/discussion.tex
@@
 
-\subsection{Role of crop species: rice vs. alfalfa}
+\subsection{Imputation efficiency: differences in rice vs. alfalfa}
 \label{sec:role_of_species}
@@
 
-\subsection{Problem size}
+\subsection{Size of the imputation problem}
 \label{sec:size_of_problem}
+Albeit this was not the main objective of the paper, and no formal effort to optimise the implementation of the various imputation methods was done, still the recorded computation times give us some interesting information.
+The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources.
+RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed.
+When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios.
+This was partially mitigated by parallelization (we used 10 CPUs in our experiments).
+All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios.
+Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
+
 Problem size, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources.
@@
 Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
+
 The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods.

### 2016-01-26 inputs/discussion.tex
modifiche 12:38

@@
 With larger amounts of missing data the complexity of the imputation problem tends to increase.
-Moreover, with such a high level of missingness, even imputation errors as low as 1\% may have negative impact on successive analysis, amplifying the importance of finding the most accurate imputation method.\\
+Imputation error can have a negative impact on successive analyses (see for instance \cite{rutkoski_imputation_2013}).
+Contrary to SNP array data, where sporadic missing genotypes are around $2$-$5\%$, with the large amounts of missing data typical of GBS experiments, even imputation errors as low as 1\% may be detrimental, amplifying the importance of finding the most accurate imputation method.\\
 The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points.
@@
 
-Problem size, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources.
-RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed.
-When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios.
-This was partially mitigated by parallelization (we used 10 CPUs in our experiments).
-All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios.
-Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
-
 The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods.

### 2016-01-26 inputs/materials_and_methods.tex
modifiche 26 mattina 10:36
modifiche 12:38

@@
 \label{eq:SVDI_regression}
-Y = U\beta +\varepsilon 
+Y = U^*\beta +\varepsilon 
 \end{equation}
 
-with Y as vector of genotypes at marker j; $\hat{U}$ matrix of k eigenvectors; $\beta$ vector of regression coefficients; $\varepsilon$ random errors terms; (iv) all eigenvectors (matrix U) and $\beta$ were used to estimate missing values at marker j.
+with Y as vector of genotypes at marker j; $U^*$ matrix of the first k eigenvectors; $\beta$ vector of regression coefficients; $\varepsilon$ random errors terms; (iv) all eigenvectors (matrix U) and $\beta$ were used to estimate missing values at marker j.
 The procedure was repeated (steps (ii) to (iv)) until convergence.
@@
 This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.
+
+[ADD command line!!]
 

### 2016-01-31 inputs/abstract.tex
phrasing

@@
 Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
-and one genotype-specific method (``Beagle'') were measured 
+and one genotype-specific method (``Beagle'' and \textcolor{red}{FILLIN}) were measured 
 on GBS data from alfalfa (\emph{Medicago sativa}, autotetraploid, heterozygous, 

### 2016-01-31 inputs/introduction.tex
phrasing

@@
 and the FILLIN algorithm implemented in the programming suite Tassel). The relative efficiency of 
-the different imputation methods was assessed in terms of the 
-accuracy achieved and the computation time required.
-Not only the total imputation accuracy was measured, but also the accuracy of imputing the different genotype classes (major and minor homozygotes, heterozygotes).\\
+the different imputation methods was assessed in terms of  
+accuracy and computation time. Accuracy was measured in terms
+of gross fraction of correctly imputed genotypes, and within
+genotype classes (major/minor homozites and heterozigotes, when present).

### 2016-01-31 inputs/introduction.tex
phrasing, references

@@
 as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).
-Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward_2013_saturated}).\\
+Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}).\\
 Compared to SNP-chip data, GBS data are more challenging: many more 

### 2016-01-31 inputs/materials_and_methods.tex
phrasing, references

@@
 
-A further quality filter, implemented through ad-hoc Python scripts, removed heterozygous loci with less than 4 and homozygous loci with less than 11 aligned reads.
-A similar filtering was performed in \cite{Rocher_validation_2015} using less restrictive thresholds.
 Alfalfa is an autotetraploid plant species and can therefore present three different heterozygous genotypes: AAAB, AABB and ABBB.
@@
 AB), while the two homozygous genotypes (AAAA and BBBB) were considered diploid homozygotes (i.e., AA or BB).
+A further quality filter, implemented through ad-hoc Python scripts, removed heterozygous loci with less than 4 and homozygous loci with less than 11 aligned reads.
+A similar filtering was performed in \cite{Rocher_validation_2015} using less restrictive thresholds.
 
@@
 \label{sec:imputation_methods}
-We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite \cite{swarts_novel_2014}.
+We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite.
 For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.

### 2016-01-31 inputs/materials_and_methods.tex
phrasing

@@
 \label{par:MNI}
-: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, which was then discretized to the closest value in \{0,1,2\}. 
+: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, then discretized to the closest value in \{0,1,2\}.
 \paragraph{KNNI}
@@
 The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}.
-``FILLIN'' is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function.
+FILLIN is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function.
 This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.
-
-[ADD command line!!]
 
@@
 These were then imputed with the various algorithms, measuring the resulting imputation accuracy and computation time.
-Unlike imputation experiments with SNP-chips (e.g., \cite{hickey_factors_2012}), GBS data intrinsically present a much larger proportion of missing genotypes, and a certain amount of naturally missing genotypes has to be allowed along with artificially introduced ones (see for instance \cite{rutkoski_imputation_2013}).
 We extracted four datasets from the overall data that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker.
@@
 This procedure produced 16 combinations for each population: 4 call rate datasets $\times$ 4 levels of artificial missing markers.
-On each combination, we applied the 5 described imputation methods.
+On each combination, we applied the 6 described imputation methods.
 Rice data were analysed as a whole and by each of the 12 chromosomes separately.
-\\
-For rice data only, the position of the SNPs on the genome was known.
-It was therefore possible to test the imputation performance with and without this information.
-Among the proposed imputation methods, only Beagle can use SNP position in the imputation process.
-We therefore applied Beagle both to SNP ordered per chromosome and along the entire genome and to SNP that were randomly reshuffled (so as to break the linkage between them).
-\\
-All of the above led to a total of 1408 (1200 + 208 for reshuffled rice data with Beagle) imputation analyses.
+In absence of alfalfa genome, the Medicago Truncatula genome was used as reference for those algorithms (Beagle and FILLIN) requiring markers positions.
+
+To further investigate the effect of presence (or absence) of genome information we generated ``reshuffled'' datasets where marker positions were randomly assigned inside each chromosome.
+This way, the linkage disequilibrium between markers is broken and the relative performance decrease can be directly assessed.
+The reshuffled datasets were tested only on Beagle and FILLIN.
 
@@
 \label{sec:imputation_accuracy_and_efficiency_assessment}
-We used the artificial missing genotypes to measure the performance of the five imputation methods.
+We used the artificial missing genotypes to measure the performance of the six imputation methods.
 The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele).
@@
 
-where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise.
-We obtained four accuracy measures for alfalfa (three genotype classes plus total accuracy) and three for rice (two genotype classes plus total accuracy).\\
-For each imputation method, also the total computation time needed to complete the analyses was measured, in order to get an idea of their relative performance.
+where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa and three for rice.\\
+For each imputation method, the total computation time needed to complete the analyses was measured as an indicator of their relative performance.
 The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution.
@@
 Data handling, editing and preparation, summary of results and plots were all performed using the open-source environment for statistical programming R \cite{r_core_team_r:_2014}.
-All imputation methods, except Beagle, were implemented in R: MNI directly in base R; KNNI using the function \emph{knncatimpute()} from the Scrime package \cite{schwender_scrime:_2013}; SVDI with the R package ``bcv'' \cite{perry_bcv:_2009}; RFI using the ``MissForest'' \cite{stekhoven_missforestnon-parametric_2012} R package (with parameters: ntree=100, maxiter=10, parallelize='variables').
+All imputation methods except Beagle and FILLIN were implemented in R: MNI directly in base R; KNNI using the function \emph{knncatimpute()} from the Scrime package \cite{schwender_scrime:_2013}; SVDI with the R package ``bcv'' \cite{perry_bcv:_2009}; RFI using the ``MissForest'' \cite{stekhoven_missforestnon-parametric_2012} R package (with parameters: ntree=100, maxiter=10, parallelize='variables').
 The ``Beagle'' localized haplotype clustering imputation method was implemented with the Beagle software version 3.3.2 \cite{browning_rapid_2007}.
+The FILLIN algorithm is implemented using the Tassel CLI plugin \emph{-FILLINFindHaplotypesPlugin} followed by \emph{-FILLINImputationPlugin}.
+The latter allowed accuracy measurements through \emph{-accuracy} and \emph{-propSitesMask} options.

### 2016-01-31 inputs/results.tex
modifiche 12:38
added internal notes on things to do

@@
 GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}).
-The overall missing rate was 0.666, 0.596 and 0.534 in the three datasets.
+The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets.
 Supplementary Figure S1 shows the density distribution of missing rate per markers in the complete datasets.
@@
 
+\textcolor{red}{Some numbers on alignment on reference genome.}
+
 \begin{figure}
@@
 
+\textcolor{red}{
+\begin{itemize}
+  \item FILLIN on Rice: explain uninputed fraction
+  \item FILLIN on Alfalfa: FAIL
+  \item all rice chromosome gave the same results, as in Suppl figures
+\end{itemize}
+}
 In rice, the average imputation accuracy over all genotypic classes was close to 100\% with most methods and still above 90\% at maximum proportions of allowed (70\%) and artificial (20\%) missing genotypes (Figure~\ref{fig:accuracy_rice}).

### 2016-01-31 reviews/answer_to_revieweres.txt
added answer draft

COMMENTS TO THE AUTHOR:

Reviewer #1: In the MS entitled "Marker imputation efficiency for Genotyping-By-Sequencing data of crop species with and without a reference genome" Authors used different tools to impute missing SNP data in Medicago sativa, and rice. The MS demonstrated very good approach to impute SNP data even independent of whole genome sequence. No doubt the efforts made by authors are valuable. I have some major concern with the MS presentation, flow of information and regarding the novelty. Here are my point-wise comments:

General Comments:
The four methods used for imputation are now very common. Even software tool like Tassel host imputations methods like FILLIN, FSFHap, Numerical, and Beagle. There are several methods and tools available for imputation and many studies demonstrating comparisons have been published.

ANSWER: the main interest in our work was to assess the imputation efficiency in two important plant species, rice (major crop species) and alfalfa (forage species). We wanted to evaluate the effects of increasing missing rates, of aligning markers along the genome, and of the size of the dataset. We were also interested in checking how different imputation methods responded to different conditions. The main points that we think are novel and of interest in our owrk are: i) the imputation of missing data from GBS experiments in alfalfa (M. sativa) and rice (O. sativa); ii) the breakdown of the imputation accuracy into the different genotype classes (major and minor homozygotes, heterozygotes); iii) the effect of aligning markers to the reference genome of the species or to that of a closely related species

More particularly, in species without availability of whole genome sequence methods for the imputation of SNP data have been published. For instance, Ward, et al. (Saturated linkage map construction in Rubus idaeus using genotyping by sequencing and genome-independent imputation. 2013. BMC genomics, 14(1), 2).

Even in Alfaalfa (Medicago sativa) Rocher et al. (Validation of Genotyping-By-Sequencing Analysis in Populations of Tetraploid Alfalfa by 454 Sequencing. 2015. PloS one, 10(6), e0131918.) demonstrated imputation at missing rate 0.5.

ANSWER: the mentioned papers are indeed relevant, and were cited in the introduction of the paper

Specific comments:

In title "crop species with and without" is miss leading since authors have used two species so better to be specific "medicago and rice".

ANSWER: the title was changed to "Marker imputation efficiency for Genotyping-By-Sequencing data in rice (Oryza sativa) and alfalfa (Medicago sativa)"

Abstract need to be more concise and informative

ANSWER:

Introduction:
-------------
Line 38-39, GWAS can not handle missing data- need citation and reframe the sentence. Better to verify the fact.

ANSWER: In the paper, we were actually saying that most models and software for GWAS etc can not handle missing data, which is quite true; GWAS and genomic predictions with missing genotypes is generally more difficult. We added a reference to a review paper by Ben Hayes et al, 2009, in animal genomics. Especially linear algebra and arithmetic methods methods suffer from missing data, whereas numerical methods like MCMC can better cope with them. In any case, using missing data, is not impossible: e.g. maximum likelihood (Handling missing data by maximum likelihood, by P.D. Allison, 2012; or Schafer, 1997: Analysis of incomplete multivariate data). We feel this is out of the scope of the present paper, though, and did not indulge in such speculations. We simply modified the sentence and added a reference.

Line 43-44 - alternative to SNP-chip- need to improve

ANSWER:  The sentence was modified to "Genotyping-by-sequencing (GBS) is a relatively recent technique which is considered a viable alternative to SNP array-based genotyping to produce SNP genotype data"

Line 48 - '52% or up to 52%' need to verify the fact

ANSWER: From Crossa et al: around 52% and 58% in two maize datasets; from Poland et al: quoting from the text "up to 80% missing data per marker" (Page 105)

Line 55 -57 - imputation method has been developed for SNP-Chips. I am not sure about the accuracy of statement

ANSWER: sentence modified to "Most methods for genotype imputation have been applied to SNP array data, and typically yield very high imputation accuracy"

Line 85- rice genome recently assemble - not true

ANSWER: it was assembled in 2005, and then recently updated (2013). We made this clearer in the text

Material methods:
-----------------
Since the MS specifically focused on imputation of SNP data, the details like salt tolerance or high dry matter yield is unnecessary. So better to concise portion from 96-114 in single sentence.

ANSWER:

Similarly details about imputation methods can be concise (page 7 to 12) and more details can be provided as supplementary.

ANSWER: details on FILLIN were added; for the other method, we tried to make it more coincise, when possible

Results:
--------
Need to avoid words like close to 100% better to be very precise.

ANSWER:

Maximum missing allowed and maximum percentage of missing actually present are two different things and it need to be very clear.

ANSWER: [do not know exactly what to do; leave for later on, once a new draft of the paper is complete]

Discussion
----------
Subtitles are not good - for instance "Role of crop species" and "Size of the problem".

ANSWER: some subtitles in the Discussion were changed

Previous publication demonstrating imputation in alfalfa, and other species without genome need to be discuss. [can be expanded]
Also it will be better to highlight importance of percent study while discussing previous paper comparing different imputation methods. [percent study??]

Fig 2. and 3 - Y axis need to be zoomed to make difference visible. [when possible; contrasts with reviewer 2]

ANSWER: Figures 2 and 3 were modified. One reviewer was asking to keep the y-axis constant over all comparisons, at the expense of visualizing differences between methods. The other reviewer asked to zoom on the y-axis to make such differences visible. We decide to follow the suggestion of the latter reviewer, and zommed on the y-axis; this was specified in the text and in the Figure captions, to help the reader not get confused.

I suggest to reframe the MS to make it pursuable  for readers of Molecular Breeding

ANSWER: we made an effort in that direction

Reviewer 2

Manuscript: Marker imputation efficiency for Genotyping-By- Sequencing data of crop species with and without a reference genome
Authors: Nelson Nazzicari, Filippo Biscarini, Paolo Cozzi, E. Charles Brummer, Paolo Annicchiarico

General comments
This manuscript deals with an important and emerging topic in plant (and animal) breeding as GBS is considered as a viable alternative to SNP array-based genotyping. These GBS data come with major challenges, also on the imputation of missing values. This manuscript describes the application of six algorithmic approaches to data on two plant crops that are on the edges of the wide range of genotype complexity in plants. Not surprisingly, the Beagle software performs very well (superior over all other methods) when map information is available, and much worse, when map information is absent. 
My main concerns on the paper are:

1. Why these two species? They have very different properties which makes it difficult to associate the differences in results to particular characteristics. For example, is ploidy level more critical than a reference genome?  It might have been more insightful when including an outbred diploid species with a reference genome and similar to the rice to assess imputation accuracy with and without reference genome information. [can we include other data?]

2. A missed opportunity to study the utilization of a reference genome of a related species, as is available for alfalfa

ANSWER: In alfalfa (Medicago sativa), where no reference genome is available yet, SNP markers were aligned to the genome of a closely related species, Medicago truncatula. Results have been added to the plots. [results in the text: Nelson? Io?]

3. The choice of the five imputation methods. Why these?

ANSWER:

4. Why performs Beagle still well with the unordered SNPs in the rice dataset. How severe was the re-ordering?

ANSWER: Beagle performs still relatively well in rice even when reshuffling (changing the order) the SNPs. However, imputation accuracy was 0.99 with Beagle, and 0.94-0.96 when reshuffling markers. In the minority class, it was 0.96-0.98 with ordered data, and 0.60-0.80 with reshuffled data. Reshuffling was completely random, in order to break the order of SNP along the genome. The fact that imputation was not complete mayhem, is probably due to the fact that reshuffling was performed within-chromosome, thereby preserving some LD structure. As a matter of fact, results from the entire rice dataset were worse: 0.92-0.94 total imputation accuracy and 0.43-0.64 accuracy in the minority class, with reshuffled data.

5. Too much attention for computational issues while this was not the main objective of the study (there was no effort to optimize implementation of the various methods, except for RF).

ANSWER: This section has been re-written, shortened where possible, and a statement, in the discussion, on the limitations of the information on computation time was added. [Need to add results for FILLIN --> Nelson made the plot]

6. No attention to results on individual rice chromosomes. In case there were no clear differences between chromosomes, then that should be stated more clearly.

ANSWER:

In addition, various specific comments are provided below that may be of help to further strengthen the manuscript.­

Specific comments
Line 52: It would be nice if you could provide a list of associated packages, AND explain why Beagle was selected out of these methods.

ANSWER

Line 78: There is no development of the best method in this study, nevertheless the results of this study may identify the essential properties of such a method. Was this discussed?

ANSWER:

Line 83: For alfalfa one could use the assembly of Medicago Trancatula – was this considered?

ANSWER:

Line 86: Why these four general methods?

ANSWER:

Line 109: How to interpret the 437 plants from 391 accessions as these accessions are supposed to be fully homozygous? Were there duplicates?

ANSWER: yes, they're duplicates

Line 134: Why were these values (4 and 11) were chosen? I do not see the rational of taking 4 for a heterozygote and 11 for homozygote - shouldn’t these be the reverse?

ANSWER:

Line 135-139: The reduction from 5 genotypic classes to 3 genotypic classes in alfalfa is a critical simplification step. There is no justification why this step was executed. My assumption is that this is because of the (software) methods that were used. The explanation and justification should be provided here. I am wondering whether the four general methods could be applied to the 5 genotypic classes – was this considered/studied?

ANSWER:

Line 166:  How to interpret the ‘K closest markers’ in case there is no map or assembly information (as here for alfalfa)?

ANSWER: the distance in KNN is not the physical distance between markers along the genome, but the distance between samples (plants) based on some function of the predictors (SNP genotypes, in this case). The Simple Matching Coefficient distance (Schwender, 2007) was used in this implementation of KNN.

Line 169: How was distance specified, i.e., in bp or cM?  Did you try both and if so, where there differences?

ANSWER: again, this is not the phyisical distance between markers, but the distance between observations based on a function of the SNPs (see above)

Line 173-181: I wonder whether this explanation is appropriate for this audience.

ANSWER: we made it a bit shorter, but not too much so; we think it is helpful for the reader to be reminded at least of the basic steps of a method

Line 175: Were the eigenvectors of M’M and MM’ merged to arrive at the ‘first k eigenvectors’?

ANSWER: the first k eigenvectors were determined based on the ordered corresponding eigenvalues. The matrix U* contained these k eigenvector

Line 178: Which value for k was used, was there a decision criteria used?

ANSWER: [che diciamo qui? Seguiamo Rutkoski (o che per lei)?]

Line 182: For the SVD, how critical is the initial imputation step by MNI?  Were other methods considered for this initial imputation? [I think this is just an initialization prior, and the least informative the better]

ANSWER: the initial imputation step in SVD is just aimed at providing a sort of initialization "prior", since singular value decomposition can not be performed with missing values in the matrixes. The least informative this "prior" (e.g. MNI), the better. No other methods to initialize the marker matrix for SVD were tried.

Line 185 – 198: Explain the ‘hat’ for U in equation 2. Also, why has Y no hat while there is a hat on the Y in equation 3.

ANSWER: the hat on U was a mistake and therefore removed; it is not a matrix of estimated values. However, we used the superscript U*, since this is the matrix of the first k eigenvectors of U, and therefore it is not equal to U. In equation [2], Y is the vector of observations (no hat); in equation [3] Y is the vector of predicted values (hat).

Line 217: Elaborate on the content of this likelihood file or provide a reference (to the manual of Beagle)?

ANSWER: more details were added to the text, plus a reference to the Beagle Manual

Line 229: What was the reason to not go beyond 20% artificially induced missing values?

ANSWER: we allowed maximum 70% missing genotypes per marker; in addition, we added up to 20% artificially missing genotypes. In this way, for some markers we may arrive to 90% missing data, which we considered to be the limit we didn't want to break in this study.

Line 254: It is not clear whether the ‘total number of missing data’ refers to the sum of ‘natural’ and ‘artificial’ missing data or only to the latter part. Please improve description.

ANSWER: it refers to the total number of artificially (introduced) missing genotypes. This was specified in the text.

Line 255: It may be confusing to use y in equation 4 as a discrete variable since Y is continuous in equations  [2] and [3]. It seems more intuitive to use g instead of y.

ANSWER: changed to g

Line 281/Table 1 & Table 2:  Why not merge the content of these two tables?  Transpose Table 1 to have the same columns. Then ADD the genotype frequencies for the four call-rate scenarios (10,20,40,70%).  In that way the table may reveal shifts in frequencies when imposing stronger thresholds, which may invoke further inferences on results of imputation accuracies.

ANSWER:

Line 282: Why not present these rates as percentages as that is probably easier to read and more consistent (with e.g., line 286).

ANSWER: this was done, and modified in the text

Line 284: Figure S1 does not contain information at the ‘per individual’ level

ANSWER

Line 304: How sensitive are the results to the imposed quality filters?

ANSWER:

Line 313/Figure 3: Suggest to maintain consistency in y-axis scale, i.e., the plots for Minor homozygous accuracy should range from 0 to 100% 
as well.

ANSWER: Reviewer 1 had the opposite suggestion, which we followed. See answer above.

Line 324: the flat response seems to hold for the studied range (0.01 – 0.20). Were higher values of missing rates studied?  It will be interesting to study whether there will be a critical point and whether these are different for the different genotypic classes.

ANSWER: [diplomatic answer to be added]

Line 347: Rather extensive reporting on computation time which seems not justified as the used methods were used as these were available to the authors, i.e.., there was not dedicated efforts to speed up the implemented methods.

ANSWER: This section has been re-written, shortened where possible, and a statement, in the discussion, on the limitations of the information on computation time was added. [Need to add results for FILLIN --> Nelson made the plot]

Line 379-381: Where was this negative impact shown?  In this study or other studies – please include references.

ANSWER: a reference to Rutkoski et al was added;we tried to explain better the difference between SNP array (~5% missing) and GBS data (~50% missing)

Line 401-407: This is still rather descriptive without any postulation whether any of the methods are less or more sensitive to MAF and/or balance.  The superior results of Beagle in rice are mostly driven by utilizing map/assembly information which is ignored by other methods.

ANSWER

Line 440: The reasoning on factor ii) is not clear – would the results for rice be less accurate when analysing the genome at once? Also, why would it simplify the imputation process?

ANSWER: First, the per-chromosome analysis simplifies calculations, and allow for straightforward parallelization. Additionally, the results with rice as a whole are less accurate when markers are reshuffle [add numbers/supplementary Table/Figure?]; this is because the LD structure is better preserved when reshuffling markers.

Line 443: Did you explore the population structure of the three datasets? Would presence/absence of (severe) population structure affect imputation accuracy?

Answer: i) no structure in alfalfa, presence of structure in rice (indica/tropical/temperate); ii) implicitely exploited by most imputation methods (e.g. KNN --> distance based on genetic relatedness); iii) not likely to affect much imputation accuracy; iv) this is not GWAS: the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed;

Line 450: It is too easy to state that there is additional complexity to make use of a reference genome of a related species. It would be of HIGH relevance to explore this as it may be applied in many species that lack a reference genome themselves.

ANSWER:

Line 535-540: Too general/vague and thus not clear why this paragraph is included in the conclusion section (or even in the manuscript)! [

ANSWER: To be checked when the new draft is ready.

Figure 1: Be consistent in using rates versus percentages, i.e., this Figure uses rates while on Line 311 you refer to percentages.

ANSWER : to be checked

### 2016-02-03 inputs/materials_and_methods.tex
added internal notes on things to do
updated mEm

@@
 For Alfalfa-PV, 124 parent genotypes were chosen by stratified mass selection for dry matter yield over three harvests.
-For Alfalfa-Med, 154 parent genotypes derived from two cycles of free intercrossing among three outstanding populations in a multi-environment study.
+For Alfalfa-Med, 154 parent genotypes were derived from two cycles of free intercrossing among three outstanding populations in a multi-environment study.
 For further details see \cite{Annicchiarico2015}.
 \\
-The Rice dataset included 437 plants belonging to 391 accessions from the Rice Germplasm Collection maintained at CRA-Rice Research Unit (Vercelli, Italy).
+The Rice dataset included 437 plants belonging to 391 accessions (46 duplicates) from the Rice Germplasm Collection maintained at CRA-Rice Research Unit (Vercelli, Italy).
 The sampled collection comprised accessions from the five main sub-populations of \emph{O.
@@
 \label{sec:imputation_methods}
-We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and the ``FILLIN'' algorithm from the ``Tassel'' programming suite.
-For all algorithms, we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.
+We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and \textcolor{red}{haplotype reconstruction around the recombination site}
+(``FILLIN'' from Tassel framework).
+For all algorithms we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.
 \paragraph{MNI}
 \label{par:MNI}
-: in Mean Value Imputation each missing data point was simply replaced by the mean of the non-missing values for that marker, then discretized to the closest value in \{0,1,2\}.
+: in Mean Value Imputation each missing data point was replaced by the mean of the non-missing values for that marker, then discretized to the closest value in \{0,1,2\}.
 \paragraph{KNNI}
 \label{par:KNNI}
-: in K-Nearest Neighbors Imputation, missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001}.
-This K-neighbourhood was defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data.
+: in K-Nearest Neighbors Imputation missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001} defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data.
 $K=4$ was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
@@
 Beagle has become so popular that the name of the software is commonly used to refer metonymically to the method it implements, making the two hardly distinguishable.
-Beagle infers haplotypes and imputes sporadic missing alleles both with known and unknown phase, using a localized haplotype cluster model; this is a class of directed acyclic graphs which empirically models haplotype frequencies on a local scale and therefore adapts to local structure in the data.
+Beagle infers haplotypes and imputes sporadic missing alleles both with known and unknown phase, using a localized haplotype cluster model.
+This is a class of directed acyclic graphs which empirically models haplotype frequencies on a local scale and therefore adapts to local structure in the data.
 Beagle makes use of a hidden Markov model to find the most likely haplotype pair for each individual, given the genotype data for that individual and the graphical haplotype frequency model.
@@
 The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}.
-FILLIN is based on haplotype reconstruction around recombination break points; haplotypes are then clustered per genotype similarity using the Hamming distance function.
+FILLIN is based on haplotype reconstruction around recombination break points.
+Haplotypes are clustered per genotype similarity using the Hamming distance function.
 This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes.
+If convergence criteria are not met genotypes are left uninputed.
+Among the considered algorithms, FILLIN is the only one that can thus produce partially imputed results.
 
@@
 Rice data were analysed as a whole and by each of the 12 chromosomes separately.
-In absence of alfalfa genome, the Medicago Truncatula genome was used as reference for those algorithms (Beagle and FILLIN) requiring markers positions.
+In absence of alfalfa genome, the \emph{M.
+Truncatula} genome was used as reference for those algorithms (Beagle and FILLIN) requiring markers positions.
 
@@
 The FILLIN algorithm is implemented using the Tassel CLI plugin \emph{-FILLINFindHaplotypesPlugin} followed by \emph{-FILLINImputationPlugin}.
-The latter allowed accuracy measurements through \emph{-accuracy} and \emph{-propSitesMask} options.
+The latter allowed accuracy measurements through \emph{-accuracy} and \emph{-propSitesMask} options.\\
+The Bowtie 2 tool \cite{langmead_fast_2012} was used to query the consensus sequence of each alfalfa tag pair containing a SNP against the \emph{M.
+truncatula} reference genome Version 4.0 using the \emph{-verysensitivelocal} preset.
+\textcolor{red}{TO BE WRITTEN: what about Rice alignment, what sw was used? How many marker did align? We need to report stats for alfalfa, too? (about half of the marker aligned).}

### 2016-02-03 inputs/introduction.tex

@@
 Imputation of missing alleles and genotypes is usually a preliminary step for
-a wide range of genetic analyses: in fact, most models and software for 
+a wide range of genetic analyses. In fact most models and software for 
 population genetics, genomic selection (GS) and genome-wide association 
-studies (GWAS) can not handle missing data and require complete datasets (e.g. \cite{hayes2009invited}). 
-SNP-chips represent one of the leading genotyping technologies, and produce 
-datasets that, after markers and individuals with low call-rate are edited out, 
+studies (GWAS) can not handle missing data and require complete datasets (e.g. \cite{hayes2009invited,aulchenko2007genabel,endelman2011rrblup,perez2014genome}). 
+SNP-array represent one of the leading genotyping technologies and produce 
+datasets that, after low call-rate filtering, 
 usually still contain a small proportion of uncalled genotypes (e.g., $<$5\%) 
@@
 or ``up to 80\% missing data per marker'' in wheat \cite{poland_genomic_2012}. This 
-is due to both intrinsic properties of the technology - e.g., low coverage - 
+is due to both intrinsic properties of the technology
 and to its application to species without a mature genome assembly \cite{glaubitz_tassel-gbs:_2014}.\\
@@
 genotype imputation have been applied to SNP array data, and typically 
-yield very high imputation accuracy: for instance, $>$95\% correctly imputed 
+yield very high imputation accuracy. For instance, $>$95\% correctly imputed 
 genotypes were reported in maize \cite{hickey_factors_2012} and cattle \cite{vanraden_genomic_2011}. 
-Such methods can be applied to GBS data, too (see for instance 
-\cite{huang_efficient_2014} and \cite{swarts_novel_2014} 
-for applications in rice and maize). When a reference 
+The same imputation methods can also be applied to GBS data (see \cite{huang_efficient_2014} and \cite{swarts_novel_2014} for applications in rice and maize). When a reference 
 genome is available, SNP loci can be aligned against its sequence and 
-are therefore ordered. However, GBS data may be generated 
+are therefore ordered, thus allowing the exploitation of local linkage disequilibrium (LD).
+However, GBS data may be generated 
 also for species lacking a reference genome assembly. In this case the 
 GBS output is a series of ``floating'' loci not linked to any chromosome:
-alignment is not possible and SNPs are therefore unordered. Unordered 
+alignment is not possible and SNPs are therefore considered unordered. Unordered 
 data add additional complexity to the task of imputing missing genotypes: 
-exploitation of linkage disequilibrium (LD) and haplotype reconstruction 
+exploitation of linkage disequilibrium and haplotype reconstruction 
 are less straightforward, consecutive loci may lie on different chromosomes
@@
 Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}).\\
-Compared to SNP-chip data, GBS data are more challenging: many more 
+Compared to SNP-array data, GBS is more challenging: many more 
 missing genotypes, high variability between runs, intrinsically noisy data, 
@@
 In this paper, we imputed missing genotypes from GBS data in two agronomically
-important crop species: the forage crop alfalfa (\emph{Medicago sativa}), for 
-which a reference genome is not available yet, and the 
-cereal crop rice (\emph{Oryza sativa}) whose reference genome has been 
-assembled in 2005 \cite{international_rice_genome_sequencing_project_map-based_2005}, and
-was recently updated \cite{kawahara_improvement_2013}. We compared the imputation 
-efficiency of four general data-imputation methods (Mean Value Imputation, 
-K-Nearest Neighbour Imputation, Singular Value Decomposition Imputation 
-and Random Forest Imputation) with the performance of two methods specific 
+important crop species:
+the cereal crop rice (\emph{Oryza sativa L.}) whose reference genome has been 
+assembled in 2005 \cite{international_rice_genome_sequencing_project_map-based_2005}, and updated in 2013 \cite{kawahara_improvement_2013},
+and
+the forage crop alfalfa (\emph{Medicago sativa L.}), for 
+which a reference genome is not available yet. When necessary, alfalfa genotypes
+were mapped on the close relative \emph{Medicago truncatula L.}, a model species
+for legumes, whose genome is available \cite{young_medicago_2011}.
+The species were chosen 
+as representative of very different scenarios, alfalfa being autotetraploid
+with high heterozygosity and rice being diploid with essentially 100\% homozygozity.
+The presented study thus encompasses a wide span of real application cases.\\
+We compared the imputation efficiency of four general imputation methods (Mean
+Value Imputation, K-Nearest Neighbour Imputation, Singular Value Decomposition
+Imputation and Random Forest Imputation) with the performance of two methods specific 
 for genotype imputation (localized haplotype clustering imputation, 
-implemented in the software package Beagle, %and haplotype reconstruction around the recombination site
-and the FILLIN algorithm implemented in the programming suite Tassel). The relative efficiency of 
-the different imputation methods was assessed in terms of  
-accuracy and computation time. Accuracy was measured in terms
-of gross fraction of correctly imputed genotypes, and within
-genotype classes (major/minor homozites and heterozigotes, when present).
+implemented in the software package Beagle, \textcolor{red}{and haplotype
+reconstruction around the recombination site}, implemented in the FILLIN algorithm
+as part of Tassel suite).
+The four general algorithms were chosen as well known imputation strategies implemented in several freely available software libraries and frameworks.
+The two genotype-specific methods represent the state of the art for genotypes lacking pedigree information.\\
+The relative efficiency of the different imputation methods was assessed in terms of 
+accuracy and computation time. Accuracy was measured as fraction of 
+correctly imputed genotypes, and within genotype classes 
+(major/minor homozigotes, and heterozigotes, when present).

### 2016-02-03 inputs/discussion.tex
update in results, discussion

--- 2016-01-26 inputs/discussion.tex
@@
 \label{sec:maf}
-GBS data pose a greater challenge than SNP-chip data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they usually contain.
-On our data sets we found missing rates varying from 53\% to 67\%.
-With larger amounts of missing data the complexity of the imputation problem tends to increase.
-Moreover, with such a high level of missingness, even imputation errors as low as 1\% may have negative impact on successive analysis, amplifying the importance of finding the most accurate imputation method.\\
+GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain.
+On our data sets we found missing rates varied from 53\% to 67\%.
+With larger amounts of missing data the complexity of the imputation problem increases.
+Imputation error can have a negative impact on successive analyses (see for instance \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
 The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points.
@@
 the classes appear at different frequencies in the datasets, with typically one over-represented class
-%problem with url
 (see~\cite{kotsiantis_handling_2006}
@@
 The total classification accuracy may be misleading, being ``dominated'' by the majority class
-%problem with url
 \cite{he_learning_2009}.
@@
 Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
-Some algorithms failed almost completely the imputation of the least common genotypes, while other algorithms proved to work well.
 In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI.
 In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.
-MNI performed reasonably well in the alfalfa heterozygous class, with accuracy close to 0.6, contributed to its moderately good results (total accuracy around 0.8).
+MNI performed reasonably well in the alfalfa heterozygous class, with accuracy close 60\%, contributed to its moderately good results (total accuracy around 80\%).
 
@@
 The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
-KNNI's relatively pronounced drop in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed.
+KNNI's decrease in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed.
 On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$.
@@
 Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
-%problems with this citation: url contains hashes
 \cite{the_1000_genomes_project_consortium_integrated_2012}
-for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: 
+for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
 i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
@@
 iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
-Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.
+Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.\\
+The most dramatic difference between data sets was found in FILLIN performances.
+The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms).
+This behaviour is probably due to FILLIN being designed to target inbred lines (as the name itself indicates).
 

### 2016-02-03 inputs/results.tex
update in results, discussion

@@
 The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.
-
-\textcolor{red}{Some numbers on alignment on reference genome.}
+\textcolor{red}{
+Alignment on reference genome assigned a position to 88.66\%, xxx\%, and xxx\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively.
+}
 
@@
 \label{sec:imputation_accuracy}  
-
 \textcolor{red}{
 \begin{itemize}
-  \item FILLIN on Rice: explain uninputed fraction
-  \item FILLIN on Alfalfa: FAIL
   \item all rice chromosome gave the same results, as in Suppl figures
+  \item all numbers in this section need to be checked
+  \item what about supplementary figures? still valid?
 \end{itemize}
 }
-In rice, the average imputation accuracy over all genotypic classes was close to 100\% with most methods and still above 90\% at maximum proportions of allowed (70\%) and artificial (20\%) missing genotypes (Figure~\ref{fig:accuracy_rice}).
+In rice, the average imputation accuracy over all genotypic classes was above 80\% for all methods.
+No significant difference was found between chromosome, thus results are presented averaged (Figure~\ref{fig:accuracy_rice}).
+\\
 The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\% (Figure~\ref{fig:accuracy_alfalfa}).
+Again, since no significant difference was found between the two dataset, presented results are averaged (Figure~\ref{fig:accuracy_alfalfa}).
+
 The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 90.8\% and 99.3\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 4.1\% and 73.4\% in alfalfa and rice, respectively).
+
 Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}).
@@
 Results for each individual alfalfa population, the 12 individual rice chromosomes, and the whole rice dataset are similar to those discussed here (see Supplementary Figures S2 to S16).
-\\
+
 Missing data did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate.
-KNNI showed a decreasing imputation accuracy with increasing missing data, both in alfalfa and rice.\\
+KNNI showed a decreasing imputation accuracy with increasing missing data, both in alfalfa and rice.
+
 In rice, Beagle outperformed all other imputation methods, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}).
 When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle.
+The loss of performances on reshuffled markers is exacerbated on alfalfa datasets, and in particular on heterozygous and minor homozygous classes.
+
+Similarly, FILLIN resulted in an average drop of XXX\% when marker positions are shuffled.
+FILLIN performances were in general lower that other methods due to the fraction of unimputed genotypes, here accounted as wrong imputation.
+FILLIN did not impute an average of XXX\% missing data in Rice datasets and practically the entirety of Alfalfa dataset.
+Even trying to relax FILLIN constraint parameters, it was impossible to obtain any sensible result.
+
 In general we observed for all thresholds and in both genotype classes the following ranking of methods for imputation accuracy: RFI $>$ KNNI $>$ SVDI.
-In particular, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI.
+In Rice, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI.
 As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
@@
 They did not differ much from MNI in the overall accuracy and in the major homozygous class, but gave higher imputation accuracy in the heterozygous and minor homozygous classes.
-Beagle was also applied to alfalfa, where the relative position of markers in the genome is not known.
-The overall imputation accuracy with Beagle was about 67\%, with accuracies of almost 100\% in the major homozygous class but practically 0\% in the heterozygous and minor homozygous classes.
 

### 2016-02-03 inputs/discussion.tex
computation times and consequences in the paper

@@
 All methods specifically developed for the imputation of missing genotypes have been designed for markers ordered along the genome sequence, and make explicit or implicit use of position information.
-For alfalfathere is no reference genome sequenced yet.
+
+For alfalfa there is no reference genome sequenced yet.
 The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and could in principle be used.
@@
 Truncatula} in the present analysis.\\
+
 In rice, where a reference genome is available, we could assess the effect of using ordered vs.
@@
 Interestingly, with ordered markers Beagle shows astounding resilience to high missingness, with imputation accuracy in the minority class $>$ 0.95 even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
+
 Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 0.75 (10\% allowed, 1\% artificial missing rates) to almost 0.5 (70\% allowed, 20\% artificial missing rates).
@@
 This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.\\
+
 Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provided interesting comparisons.
@@
 Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
-
-The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods.
-While problems related to the computational requirements can usually be tackled by the constant improvement in bioinformatics facilities, current trends in genotyping suggest that imputation methods and computational strategies well suited for large data sets are to be preferred.
-The non-linearity of computational efficiency generated, for most examined imputation algorithms, a total computation time for the complete rice data set which was greater than the sum of the computation times required by the individual chromosomes.
-Again, RFI provided the most blatant illustration of this behavior, with computation times for the complete rice data set going up to 937 hours (about 40 days), while the sum of the computation times for all the single chromosomes in the same experimental conditions added to 88.3 hours.

### 2016-02-03 inputs/results.tex
computation times and consequences in the paper

@@
 Only the implementation of RFI could leverage a multi-core/multi-thread environment, so that RFI computation times should be evaluated considering 10 CPUs used in parallel, while all other algorithms used only one CPU at a time.\\
-Imputation efficiency (Figure~\ref{fig:time_analysis}) was assessed with respect to the gross dimension of the data set (i.e., number of markers $\times$ number of samples) and to the total number of missing genotypes to be imputed (e.g., 341\,097 missing genotypes for the Alfalfa-Med population with 40\% missing rate and 20\% artificial missing).
-RFI required by far the longest computation times (in spite of 10-core/thread parallelization), which grew approximately exponentially ($O(e^N)$) with problem size, and easily required tens of hours for individual rice chromosomes, and hundred of hours (up to 937 hours) for the complete rice dataset.
+Imputation efficiency (Figure~\ref{fig:time_analysis}) was assessed with respect to the gross dimension of the data set (i.e., number of markers $\times$ number of samples).
+An alternative analysis relating computation times to the number of missing genotypes brought to the same results (data not shown).\\
+RFI required by far the longest computation times (in spite of parallelization), which grew approximately exponentially ($O(e^N)$) with problem size, and easily required tens of hours for individual rice chromosomes, and hundred of hours (up to 937 hours) for the complete rice dataset.
 The second slowest algorithm was KNNI, with computation times growing approximately quadratically ($O(N^2)$) with problem complexity (the chosen KNNI implementation contains no heuristic method to prune the all vs.
@@
 All other algorithms were faster, with computation times growing linearly or logarithmically with problem size.
-Beagle and SVDI resulted in similar execution times.
+Beagle, FILLIN and SVDI resulted in similar execution times.
 MNI was by far the fastest imputation algorithm, being scarcely affected by the size of the problem (computation times ranging from 0.23 to 31.29 seconds).

### 2016-02-03 inputs/discussion.tex
phrasing

@@
 We therefore decided to not use the genome of \emph{M.
-Truncatula} in the present analysis.\\
+Truncatula} in the present analysis.
 
@@
 When tested with alfalfa (with no marker position available) Beagle largerly overestimated by the majority class, performing poorly
-This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.\\
+This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.
 
 Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provided interesting comparisons.
-In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\%to 80.9\% with 1\% to 20\% missing genotypes.
+In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\% to 80.9\% with 1\% to 20\% missing genotypes.
 Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range and finding Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI.
@@
 All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios.
-Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
+Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.

### 2016-02-03 inputs/results.tex
phrasing

@@
 
-Similarly, FILLIN resulted in an average drop of XXX\% when marker positions are shuffled.
-FILLIN performances were in general lower that other methods due to the fraction of unimputed genotypes, here accounted as wrong imputation.
-FILLIN did not impute an average of XXX\% missing data in Rice datasets and practically the entirety of Alfalfa dataset.
+Similarly, FILLIN resulted in a 5-10\% accuracy drop when marker positions were shuffled.
+FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as wrong imputation.
+FILLIN did not impute an average of 14.6\% missing data in Rice and practically the entirety of Alfalfa data.
 Even trying to relax FILLIN constraint parameters, it was impossible to obtain any sensible result.

### 2016-02-03 inputs/abstract.tex
updated abstract to cite FILLIN

--- 2016-01-19 inputs/abstract.tex
@@
 Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
-and one genotype-specific method (``Beagle'') were measured 
+and two genotype-specific method (``Beagle'' and FILLIN) were measured 
 on GBS data from alfalfa (\emph{Medicago sativa}, autotetraploid, heterozygous, 
 without reference genome) and 
-rice (\emph{Oryza sativa}, with reference genome, 100\% homozygous). Benchmarks 
+rice (\emph{Oryza sativa}, diploid, 100\% homozygous, with reference genome). Benchmarks 
 consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
-proportions (up to 20\%) of known genotypes “masked” for imputation. The 
+proportions (up to 20\%) of known genotypes masked for imputation. The 
 relative performance was measured as the total proportion of correctly imputed 
@@
 Beagle was the best performing method, both accuracy- 
-and time-wise, when using a reference genome; when no reference 
+and time-wise, when using a reference genome. When no reference 
 genome was available, KNNI and RFI gave the highest accuracies, with KNNI 
@@
 (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation 
-(SVDI), Beagle, alfalfa, rice, imputation, reference genome}
+(SVDI), Beagle, FILLIN, alfalfa, rice, imputation, reference genome}
 \end{abstract}

### 2016-02-04 inputs/introduction.tex
phrasing

@@
 
-Imputation of missing alleles and genotypes is usually a preliminary step for
+Imputation of missing alleles and genotypes is a preliminary step for
 a wide range of genetic analyses. In fact most models and software for 
@@
 alignment is not possible and SNPs are therefore considered unordered. Unordered 
-data add additional complexity to the task of imputing missing genotypes: 
-exploitation of linkage disequilibrium and haplotype reconstruction 
+data add additional complexity to genotype imputation. Exploitation of linkage 
+disequilibrium and haplotype reconstruction 
 are less straightforward, consecutive loci may lie on different chromosomes
-or scaffolds, and overall data are less homogeneous. Rutkoski et al.~\cite{rutkoski_imputation_2013}
+or scaffolds, and overall data are less homogeneous.
+Rutkoski et al.~\cite{rutkoski_imputation_2013}
 considered the application of general data imputation methods 
@@
 as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).
-Imputation of missing GBS data without alignment to the reference genome has been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015}) and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}).\\
+Imputation of missing GBS data without alignment to the reference genome has
+been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015})
+and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}). These studies
+however do not focus on measuring imputation efficiency and do not compare
+alternative imputation models.\\
+	%un po' una ripetizione. Togliere?
 	Compared to SNP-array data, GBS is more challenging: many more 
@@
 the cereal crop rice (\emph{Oryza sativa L.}) whose reference genome has been 
-assembled in 2005 \cite{international_rice_genome_sequencing_project_map-based_2005}, and updated in 2013 \cite{kawahara_improvement_2013},
+assembled in 2005 \cite{international_rice_genome_sequencing_project_map-based_2005},
+and updated in 2013 \cite{kawahara_improvement_2013},
 and
@@
 reconstruction around the recombination site}, implemented in the FILLIN algorithm
-as part of Tassel suite).
-The four general algorithms were chosen as well known imputation strategies implemented in several freely available software libraries and frameworks.
-The two genotype-specific methods represent the state of the art for genotypes lacking pedigree information.\\
+as part of Tassel suite). The four general algorithms were chosen as well known 
+imputation strategies implemented in several freely available software libraries
+and frameworks. The two genotype-specific methods represent the state of the art 
+for genotypes lacking pedigree information.\\
 The relative efficiency of the different imputation methods was assessed in terms of 

### 2016-02-04 inputs/materials_and_methods.tex
update in results, discussion
phrasing

@@
 In total, 32\,706, 40\,734 and 166\,418 SNP markers were called by GBS in Alfalfa-PV, Alfalfa-Med and Rice, respectively (see Table~\ref{tab:descriptive_statistics}).
-In both datasets SNP variants were renamed so that AA represented the most common homozygote, and BB the least common homozygote.
+In all datasets SNP variants were renamed so that AA represented the most common homozygote, and BB the least common homozygote.
 
@@
 truncatula} reference genome Version 4.0 using the \emph{-verysensitivelocal} preset.
-\textcolor{red}{TO BE WRITTEN: what about Rice alignment, what sw was used? How many marker did align? We need to report stats for alfalfa, too? (about half of the marker aligned).}
+The BWA alignment tool \cite{li2009fast} was used to align rice tags on Rice Genome Annotation Project Release 7.

### 2016-02-04 inputs/discussion.tex
phrasing

@@
 With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases.
-This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighbourhood.
+This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighborhood.
 RFI - which only partially relies on local structure of the data - suffered marginally from high missingness.

### 2016-02-04 inputs/materials_and_methods.tex
phrasing

@@
 \label{sec:imputation_methods}
-We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbour Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and \textcolor{red}{haplotype reconstruction around the recombination site}
+We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbors Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and \textcolor{red}{haplotype reconstruction around the recombination site}
 (``FILLIN'' from Tassel framework).
@@
 : in K-Nearest Neighbors Imputation missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001} defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data.
-$K=4$ was used in KNNI, and neighbour values were weighted by the reciprocal of their distance from the data point to be imputed.
+$K=4$ was used in KNNI, and neighbo
+rs values were weighted by the reciprocal of their distance from the data point to be imputed.
 \paragraph{SVDI}

### 2016-02-04 inputs/materials_and_methods.tex
phrasing

@@
 We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbors Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and \textcolor{red}{haplotype reconstruction around the recombination site}
-(``FILLIN'' from Tassel framework).
+(FILLIN from Tassel framework).
 For all algorithms we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.

### 2016-02-04 inputs/materials_and_methods.tex
phrasing

@@
 The ``Beagle'' localized haplotype clustering imputation method was implemented with the Beagle software version 3.3.2 \cite{browning_rapid_2007}.
-The FILLIN algorithm is implemented using the Tassel CLI plugin \emph{-FILLINFindHaplotypesPlugin} followed by \emph{-FILLINImputationPlugin}.
+The FILLIN algorithm \cite{swarts_novel_2014} is implemented using the Tassel CLI plugin \emph{FILLINFindHaplotypesPlugin} followed by \emph{FILLINImputationPlugin}.
 The latter allowed accuracy measurements through \emph{-accuracy} and \emph{-propSitesMask} options.\\

### 2016-02-04 inputs/results.tex
new figures, updated text to show them

@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{Fig02_Rice-chrom_avg.pdf}
+\includegraphics[width=0.95\textwidth]{figure_rice_chrom.pdf}
 \caption[Rice imputation accuracies]{
 imputation accuracies overall, for the major homozygous genotype (AA), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged over the 12 rice chromosomes.
-Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black).}
+Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black), Beagle with unordered markers (dashed black), FILLIN with ordered markers (purple) and FILLIN with unordered markers (dashed purple).}
 \label{fig:accuracy_rice}
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{Fig03_Alfalfa_avg.pdf}
+\includegraphics[width=0.95\textwidth]{figure_alfalfa.pdf}
 \caption[Alfalfa imputation accuracies]{
 imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged two alfalfa populations (Alfalfa-Med and Alfalfa-PV).
-Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black).}
+Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black).
+FILLIN was unable to impute alfalfa data and is absent from figure.}
 \label{fig:accuracy_alfalfa}

### 2016-02-04 inputs/results.tex
new time figures, figure captions

@@
 imputation accuracies overall, for the major homozygous genotype (AA), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged over the 12 rice chromosomes.
-Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black), Beagle with unordered markers (dashed black), FILLIN with ordered markers (purple) and FILLIN with unordered markers (dashed purple).}
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black), Beagle with unordered markers (dashed black), FILLIN with ordered markers (purple) and FILLIN with unordered markers (dashed purple).
+Y axis scale changes to highlight differences.}
 \label{fig:accuracy_rice}
@@
 imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged two alfalfa populations (Alfalfa-Med and Alfalfa-PV).
-Lines colors represent the five imputation algorithms: MNI (orange), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black).
-FILLIN was unable to impute alfalfa data and is absent from figure.}
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black).
+FILLIN was unable to impute alfalfa data and is absent from figure.
+Y axis scale changes to highlight differences.}
 \label{fig:accuracy_alfalfa}
@@
 \begin{figure}
-\includegraphics[width=0.95\textwidth]{Fig04_time_analysis.pdf}
-\caption[Computation times]{computation time as a function of the amount of the total size of the imputed dataset (left) and of missing data (right) for the five imputation methods (MNI: orange, KNNI: red, SVDI: blue, RFI: green, Beagle:black).
-Plots include measures on Alfalfa-Med population, Alfalfa-PV population, and rice chromosomes 1 to 12.
-Complete rice datasets are omitted for readibility.}
+\includegraphics[width=0.95\textwidth]{time_vs_size.pdf}
+\caption[Computation times]{computation time as a function of the total size of the imputed dataset.
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (light blue), RFI (green), Beagle (black) and FILLIN (purple).
+Plots include measures on Alfalfa-Med population, Alfalfa-PV population, and rice chromosomes 1 to 12. Complete rice datasets are omitted for readibility.}
 \label{fig:time_analysis}

### 2016-02-09 inputs/introduction.tex
phrasing
integrating notes from Fil+N pass

@@
 population genetics, genomic selection (GS) and genome-wide association 
-studies (GWAS) can not handle missing data and require complete datasets (e.g. \cite{hayes2009invited,aulchenko2007genabel,endelman2011rrblup,perez2014genome}). 
+studies (GWAS) can not easily handle missing data and require complete datasets (e.g. \cite{hayes2009invited,aulchenko2007genabel,endelman2011rrblup,perez2014genome}). 
 SNP-array represent one of the leading genotyping technologies and produce 
@@
 and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}). These studies
-however do not focus on measuring imputation efficiency and do not compare
+however do not focus on measuring imputation accuracy and do not compare
 alternative imputation models.\\
@@
 the forage crop alfalfa (\emph{Medicago sativa L.}), for 
-which a reference genome is not available yet. When necessary, alfalfa genotypes
+which a reference genome is not available yet. Alfalfa genotypes
 were mapped on the close relative \emph{Medicago truncatula L.}, a model species
 for legumes, whose genome is available \cite{young_medicago_2011}.
-The species were chosen 
-as representative of very different scenarios, alfalfa being autotetraploid
+The species were chosen as representative of very different scenarios, alfalfa being autotetraploid
 with high heterozygosity and rice being diploid with essentially 100\% homozygozity.
-The presented study thus encompasses a wide span of real application cases.\\
-We compared the imputation efficiency of four general imputation methods (Mean
-Value Imputation, K-Nearest Neighbour Imputation, Singular Value Decomposition
+The present study thus encompasses a wide span of real application cases.\\
+We compared the imputation accuracy of four general imputation methods (Mean
+Value Imputation, K-Nearest Neighbor Imputation, Singular Value Decomposition
 Imputation and Random Forest Imputation) with the performance of two methods specific 
 for genotype imputation (localized haplotype clustering imputation, 
-implemented in the software package Beagle, \textcolor{red}{and haplotype
-reconstruction around the recombination site}, implemented in the FILLIN algorithm
+implemented in the software package Beagle, and a method of haplotype
+reconstruction around recombination sites, implemented in the FILLIN algorithm
 as part of Tassel suite). The four general algorithms were chosen as well known 
-imputation strategies implemented in several freely available software libraries
-and frameworks. The two genotype-specific methods represent the state of the art 
+imputation strategies implemented in several freely available software libraries.
+The two genotype-specific methods represent the state of the art 
 for genotypes lacking pedigree information.\\
@@
 accuracy and computation time. Accuracy was measured as fraction of 
-correctly imputed genotypes, and within genotype classes 
+correctly imputed genotypes, both as a total and within genotype classes 
 (major/minor homozigotes, and heterozigotes, when present).

### 2016-02-09 inputs/materials_and_methods.tex
integrating notes from Fil+N pass

@@
 \label{sec:imputation_methods}
-We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbors Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and \textcolor{red}{haplotype reconstruction around the recombination site}
-(FILLIN from Tassel framework).
+We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbors Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and haplotype reconstruction around the recombination sites (FILLIN from Tassel suite).
 For all algorithms we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.

### 2016-02-09 inputs/conclusions.tex
integrating notes from Fil+N pass

@@
 The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
-The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa and, most importantly, affected the relative efficiency of the methods.
-While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome, such as in alfalfa.
-In particular, RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency.
+The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa.
+While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome.
+In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency.
+The alignment of markers to closely related species can help, but comes at the price of discarding the non-aligning markers.
+Moreover, performances indicate that general methods are still the best option.\\
 The imputation accuracy tended to be relatively robust over increasing missing rates.

### 2016-02-09 inputs/discussion.tex
integrating notes from Fil+N pass

@@
 With larger amounts of missing data the complexity of the imputation problem increases.
-Imputation error can have a negative impact on successive analyses (see for instance \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
+Imputation error can have a negative impact on successive analyses (e.g.
+on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
 The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points.
@@
 In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.
-MNI performed reasonably well in the alfalfa heterozygous class, with accuracy close 60\%, contributed to its moderately good results (total accuracy around 80\%).
 
@@
 for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
-i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
-ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);\\
-iii) alfalfa datasets did not contain population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulation.
-Population structure is implicitly used by algorithms as Beagle and KNNI that create clusters of similar genotypes.
-Thus, rice dataset population structure facilitated the inputation process;\\
-iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
-Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.\\
-The most dramatic difference between data sets was found in FILLIN performances.
+i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
+ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling phase. This semplification step brought alfalfa dataset farther from biology;\\
+iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process;\\
+iv) rice markers were aligned on native genome, while alfalfa markers were aligned on a different species.
+
+Alfalfa datasets did not present population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulations.
+Population structure is implicitly used by algorithms like Beagle, FILLIN and KNNI, that create clusters of similar genotypes, and it is likely to affect imputation accuracy.\\
+The most dramatic difference between datasets was found with FILLIN.
 The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms).
@@
 
-For alfalfa there is no reference genome sequenced yet.
-The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and could in principle be used.
-However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning the SNPs on a genome of a different species (and with a different ploidity) would introduce a further level of complexity, and the effect of alignments errors would be hardly assessable.
-Moreover, the non-aligning SNPs would be discarded, not on the basis of data quality, but of local synteny.
-We therefore decided to not use the genome of \emph{M.
-Truncatula} in the present analysis.
-
 In rice, where a reference genome is available, we could assess the effect of using ordered vs.
 unordered markers for imputation.
-Among the imputation methods used in this study, Beagle is the only one to make use of marker information, and was therefore tested with ordered and randomly reshuffled markers.
-Interestingly, with ordered markers Beagle shows astounding resilience to high missingness, with imputation accuracy in the minority class $>$ 0.95 even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
+For alfalfa there is no reference genome sequenced yet.
+The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and can in principle used.
+However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning on a different species (and with different ploidity) comes at the price of discarding those markers that do not align.
+\textcolor{red}{HOW MANY?}\\
 
-Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 0.75 (10\% allowed, 1\% artificial missing rates) to almost 0.5 (70\% allowed, 20\% artificial missing rates).
-When tested with alfalfa (with no marker position available) Beagle largerly overestimated by the majority class, performing poorly
+Among the imputation methods used in this study, Beagle and FILLIN are the only ones that make use of marker information, and were therefore tested with ordered and randomly shuffled markers.
+In rice dataset with ordered markers Beagle showed astounding resilience to high missingness, with imputation accuracy in the minority class $>$95\% even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
+Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 75\% (10\% allowed, 1\% artificial missing rates) to almost 50\% (70\% allowed, 20\% artificial missing rates).
+When tested with alfalfa Beagle largerly overestimated by the majority class and had very poor performances on heterozygous class.
+Shuffling marker positions further worsened the accuracy, and almost all missing data point assigned to the majority class.\\
+FILLIN performances on rice follow the same pattern and worsen substantially when markers are shuffled.
 This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.
 
-Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provided interesting comparisons.
+Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provide interesting comparisons.
 In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\% to 80.9\% with 1\% to 20\% missing genotypes.
@@
 Finally, Swarts et al.~\cite{swarts_novel_2014} used ordered GBS maize data to compare Beagle with the FSFHap and FILLIN algorithms they developed.
-All these algorithms rely on a Hidden Markov Model (HMM) to detect recombination break-points between haplotypes.
 They mostly obtained high imputation accuracies in the three genotype classes, providing further evidence of the added value of having a reference genome assembly.

### 2016-02-09 inputs/results.tex
integrating notes from Fil+N pass

@@
 This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous loci.
+Therefore homozygous loci tended to have higher missing rates, and to
+be included progressively more with more relaxed thresholds on allowed missing level.
 The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.
@@
 In rice, the average imputation accuracy over all genotypic classes was above 80\% for all methods.
-No significant difference was found between chromosome, thus results are presented averaged (Figure~\ref{fig:accuracy_rice}).
+No significant difference was found between chromosomes.
+Thus average accuracy over all chromosomes is therefore presented (Figure~\ref{fig:accuracy_rice}).
 \\
-The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\% (Figure~\ref{fig:accuracy_alfalfa}).
-Again, since no significant difference was found between the two dataset, presented results are averaged (Figure~\ref{fig:accuracy_alfalfa}).
+The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\%.
+No significant difference were found between the two datasets.
+Thus average accuracy over the Alfalfa-Med and Alfalfa-PV is presented (Figure~\ref{fig:accuracy_alfalfa}).
 
-The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 90.8\% and 99.3\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 4.1\% and 73.4\% in alfalfa and rice, respectively).
-
+The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging \textcolor{red}{XXX}\% and \textcolor{red}{XXX}\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging \textcolor{red}{XXX}\% and \textcolor{red}{XXX}\% in alfalfa and rice, respectively).
 Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}).
 Heterozygotes had an intermediate imputation accuracy ranging from 0\% with MNI to 70\% with KNNI.
-Results for each individual alfalfa population, the 12 individual rice chromosomes, and the whole rice dataset are similar to those discussed here (see Supplementary Figures S2 to S16).
-
+Results for each individual alfalfa population and the 12 individual rice chromosomes are similar to those discussed here (see Supplementary Figures S2 to S16).
+\\
 Missing data did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate.
@@
 
-In rice, Beagle outperformed all other imputation methods, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}).
+Beagle outperformed all other imputation methods in rice, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}).
 When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle.
-The loss of performances on reshuffled markers is exacerbated on alfalfa datasets, and in particular on heterozygous and minor homozygous classes.
-
+The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular on heterozygous and minor homozygous classes.\\
 Similarly, FILLIN resulted in a 5-10\% accuracy drop when marker positions were shuffled.
@@
 FILLIN did not impute an average of 14.6\% missing data in Rice and practically the entirety of Alfalfa data.
-Even trying to relax FILLIN constraint parameters, it was impossible to obtain any sensible result.
+Even trying to relax FILLIN constraint parameters, it was impossible to obtain sensible results.
+Thus, FILLIN accuracy does not appear in Figure~\ref{fig:accuracy_alfalfa}.
 
-In general we observed for all thresholds and in both genotype classes the following ranking of methods for imputation accuracy: RFI $>$ KNNI $>$ SVDI.
+When examining general imputation methods, resulting accuracies highlighted RFI and KNNI as the best ones.
+The two methods had comparable performances, without a preferable option over all experiments.
+SVDI followed, and MNI ranked last.\\
 In Rice, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI.
 As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
-In alfalfa KNNI showed the highest imputation accuracy over all classes and in the heterozygous and minor homozygous genotype classes (Figure~\ref{fig:accuracy_alfalfa}).
+In alfalfa KNNI showed the highest imputation accuracy over all classes \textcolor{red}{(XXX\%)} and in the heterozygous \textcolor{red}{(XXX\%)} and minor homozygous \textcolor{red}{(XXX\%)} genotype classes (Figure~\ref{fig:accuracy_alfalfa}).
 RFI and SVDI were second ranking.
-They did not differ much from MNI in the overall accuracy and in the major homozygous class, but gave higher imputation accuracy in the heterozygous and minor homozygous classes.
+They did not differ much from MNI in the overall accuracy \textcolor{red}{(XXX\%)} and in the major \textcolor{red}{(XXX\%)} homozygous class, but gave higher imputation accuracy in the heterozygous \textcolor{red}{(XXX\%)} and minor homozygous \textcolor{red}{(XXX\%)} classes.
 

### 2016-02-09 reviews/answer_to_revieweres.txt
integrating notes from Fil+N pass

@@
 
-Reviewer #1: In the MS entitled "Marker imputation efficiency for Genotyping-By-Sequencing data of crop species with and without a reference genome" Authors used different tools to impute missing SNP data in Medicago sativa, and rice.
+> Reviewer #1: In the MS entitled "Marker imputation efficiency for Genotyping-By-Sequencing data of crop species with and without a reference genome" Authors used different tools to impute missing SNP data in Medicago sativa, and rice.
 The MS demonstrated very good approach to impute SNP data even independent of whole genome sequence.
@@
 General Comments:
-The four methods used for imputation are now very common.
+> The four methods used for imputation are now very common.
 Even software tool like Tassel host imputations methods like FILLIN, FSFHap, Numerical, and Beagle.
@@
 sativa) and rice (O.
-sativa); ii) the breakdown of the imputation accuracy into the different genotype classes (major and minor homozygotes, heterozygotes); iii) the effect of aligning markers to the reference genome of the species or to that of a closely related species
+sativa); ii) the breakdown of the imputation accuracy into the different genotype classes (major and minor homozygotes, heterozygotes), never done in other comparable papers; iii) the effect of aligning markers to the reference genome of the species or to that of a closely related species; iv) the effect of shuffling marker position, that allows to quantify the importance of having (good) alignments.
 
-More particularly, in species without availability of whole genome sequence methods for the imputation of SNP data have been published.
+> More particularly, in species without availability of whole genome sequence methods for the imputation of SNP data have been published.
 For instance, Ward, et al.
@@
 BMC genomics, 14(1), 2).
-
-Even in Alfaalfa (Medicago sativa) Rocher et al.
+> Even in Alfaalfa (Medicago sativa) Rocher et al.
 (Validation of Genotyping-By-Sequencing Analysis in Populations of Tetraploid Alfalfa by 454 Sequencing.
@@
 
-ANSWER: the mentioned papers are indeed relevant, and were cited in the introduction of the paper
+ANSWER: the mentioned papers are indeed relevant, and were cited in the introduction of the paper.
+However, none of the cited papers focus on algorithm comparison, nor examine the effect of different levels of call rate.
 
@@
 
-In title "crop species with and without" is miss leading since authors have used two species so better to be specific "medicago and rice".
+> In title "crop species with and without" is miss leading since authors have used two species so better to be specific "medicago and rice".
 
@@
 
-Abstract need to be more concise and informative
-
-ANSWER:
+> Abstract need to be more concise and informative
+ANSWER: we rephrased parts of the abstract.
 
@@
 -------------
-Line 38-39, GWAS can not handle missing data- need citation and reframe the sentence. Better to verify the fact.
+> Line 38-39, GWAS can not handle missing data- need citation and reframe the sentence. Better to verify the fact.
 
-ANSWER: In the paper, we were actually saying that most models and software for GWAS etc can not handle missing data, which is quite true; GWAS and genomic predictions with missing genotypes is generally more difficult.
-We added a reference to a review paper by Ben Hayes et al, 2009, in animal genomics.
+ANSWER: We added a reference to a review paper by Ben Hayes et al, 2009, in animal genomics, and cited other common GWAS and GS software libraries (BGLR, rrBLUP, GenABEL) that contain imputation functionalities or straightforward refuse input containing missing data points.
 Especially linear algebra and arithmetic methods methods suffer from missing data, whereas numerical methods like MCMC can better cope with them.
@@
 We simply modified the sentence and added a reference.
+We rephrased the sentence to highlight that imputation is often a necessary step.
 
-Line 43-44 - alternative to SNP-chip- need to improve
+> Line 43-44 - alternative to SNP-chip- need to improve
 
@@
 
-Line 48 - '52% or up to 52%' need to verify the fact
+> Line 48 - '52% or up to 52%' need to verify the fact
 
@@
 
-Line 55 -57 - imputation method has been developed for SNP-Chips. I am not sure about the accuracy of statement
+> Line 55 -57 - imputation method has been developed for SNP-Chips. I am not sure about the accuracy of statement
 
@@
 
-Line 85- rice genome recently assemble - not true
+> Line 85- rice genome recently assemble - not true
 
-ANSWER: it was assembled in 2005, and then recently updated (2013). We made this clearer in the text
+ANSWER: it was assembled in 2005, and then recently updated (2013). We made this clearer in the text.
 
@@
 -----------------
-Since the MS specifically focused on imputation of SNP data, the details like salt tolerance or high dry matter yield is unnecessary. So better to concise portion from 96-114 in single sentence.
+> Since the MS specifically focused on imputation of SNP data, the details like salt tolerance or high dry matter yield is unnecessary. So better to concise portion from 96-114 in single sentence.
 
-ANSWER:
+ANSWER: the section has been reduced.
 
-Similarly details about imputation methods can be concise (page 7 to 12) and more details can be provided as supplementary.
+> Similarly details about imputation methods can be concise (page 7 to 12) and more details can be provided as supplementary.
 
-ANSWER: details on FILLIN were added; for the other method, we tried to make it more coincise, when possible
+ANSWER: details on FILLIN were added; for the other method, we tried to make it more concise, when possible
 
@@
 --------
-Need to avoid words like close to 100% better to be very precise.
+> Need to avoid words like close to 100% better to be very precise.
 
-ANSWER:
+ANSWER: we now always report precise values.
 
-Maximum missing allowed and maximum percentage of missing actually present are two different things and it need to be very clear.
+> Maximum missing allowed and maximum percentage of missing actually present are two different things and it need to be very clear.
 
-ANSWER: [do not know exactly what to do; leave for later on, once a new draft of the paper is complete]
+ANSWER: we clarified when necessary.
 
@@
 ----------
-Subtitles are not good - for instance "Role of crop species" and "Size of the problem".
+> Subtitles are not good - for instance "Role of crop species" and "Size of the problem".
 
@@
 
-Previous publication demonstrating imputation in alfalfa, and other species without genome need to be discuss. [can be expanded]
-Also it will be better to highlight importance of percent study while discussing previous paper comparing different imputation methods. [percent study??]
+> Previous publication demonstrating imputation in alfalfa, and other species without genome need to be discuss.
+Also it will be better to highlight importance of present study while discussing previous paper comparing different imputation methods.
 
-Fig 2. and 3 - Y axis need to be zoomed to make difference visible. [when possible; contrasts with reviewer 2]
+ANSWER: we added several publications and highlighted our position when compared to the cited studies.
+
+> Fig 2. and 3 - Y axis need to be zoomed to make difference visible.
 
@@
 The other reviewer asked to zoom on the y-axis to make such differences visible.
-We decide to follow the suggestion of the latter reviewer, and zommed on the y-axis; this was specified in the text and in the Figure captions, to help the reader not get confused.
+We decide to follow the suggestion of the latter reviewer, and zoomed on the y-axis; this was specified in the text and in the Figure captions, to help the reader not get confused.
 
-I suggest to reframe the MS to make it pursuable  for readers of Molecular Breeding
+> I suggest to reframe the MS to make it pursuable  for readers of Molecular Breeding
 
@@
 
-Reviewer 2
+===========================
+Reviewer #2
 
@@
 
-1.
+> 1.
 Why these two species?
@@
 It might have been more insightful when including an outbred diploid species with a reference genome and similar to the rice to assess imputation accuracy with and without reference genome information.
-[can we include other data?]
 
-2. A missed opportunity to study the utilization of a reference genome of a related species, as is available for alfalfa
+ANSWER: we decided to use two very different species to cover a wide range of possible outcomes.
+The core idea is to follow the well known practice of having an easy benchmark and a difficult one, as it is often done in machine learning community.
+Moreover, while a general study with several species and datasets would be desirable, it was out of the scope of the current work (and effort).
+The paper results are consistent both internally and with the cited literature, and we think shine some light on the examined topic, even without the use of many different datasets.
 
-ANSWER: In alfalfa (Medicago sativa), where no reference genome is available yet, SNP markers were aligned to the genome of a closely related species, Medicago truncatula.
-Results have been added to the plots.
-[results in the text: Nelson?
-Io?]
+> 2. A missed opportunity to study the utilization of a reference genome of a related species, as is available for alfalfa
 
-3. The choice of the five imputation methods. Why these?
+ANSWER: We followed the reviewer advice and added the results aligning on Medicago truncatula genome.
 
-ANSWER:
+> 3. The choice of the five imputation methods. Why these?
 
-4. Why performs Beagle still well with the unordered SNPs in the rice dataset. How severe was the re-ordering?
+ANSWER: the (now six) imputation methods selected cover a good part of the current practice for imputation method that do not require genealogy information (like Fimpute or IMPUTE2).
+The genotype oriented ones (Beagle and FILLIN) are very well known and typically used in plant community.
+The general ones (KNNI, RFI, SVDI and MNI) are general methods easily implemented in different software platforms and libraries and hardly require any coding to be used.
+We understand that no list of algorithms could be complete, but following the steps of similar papers (e.g.
+Rutkoski et al, G3: Genes| Genomes| Genetics, 2013) we think to have embraced a good compromise between inclusiveness, representation, and feasibility.
+
+> 4. Why performs Beagle still well with the unordered SNPs in the rice dataset. How severe was the re-ordering?
 
@@
 
-5. Too much attention for computational issues while this was not the main objective of the study (there was no effort to optimize implementation of the various methods, except for RF).
+> 5. Too much attention for computational issues while this was not the main objective of the study (there was no effort to optimize implementation of the various methods, except for RF).
 
 ANSWER: This section has been re-written, shortened where possible, and a statement, in the discussion, on the limitations of the information on computation time was added.
-[Need to add results for FILLIN --> Nelson made the plot]
 
-6. No attention to results on individual rice chromosomes. In case there were no clear differences between chromosomes, then that should be stated more clearly.
+> 6. No attention to results on individual rice chromosomes. In case there were no clear differences between chromosomes, then that should be stated more clearly.
 
-ANSWER:
-
-In addition, various specific comments are provided below that may be of help to further strengthen the manuscript.­
+ANSWER: we made this aspect more clear (in short: no big differences between chromosomes).
 
 Specific comments
-Line 52: It would be nice if you could provide a list of associated packages, AND explain why Beagle was selected out of these methods.
+> Line 52: It would be nice if you could provide a list of associated packages, AND explain why Beagle was selected out of these methods.
 
-ANSWER
+ANSWER: we added several references to software sources.
 
-Line 78: There is no development of the best method in this study, nevertheless the results of this study may identify the essential properties of such a method. Was this discussed?
+> Line 78: There is no development of the best method in this study, nevertheless the results of this study may identify the essential properties of such a method. Was this discussed?
 
-ANSWER:
+ANSWER: given the very different mathematical models underlying the examined algorithms (or all algorithms, for what it's worth) we think that the essential properties of a good imputation algorithm are discussed ones, namely: accuracy (total and per class); resilience to growing missing level; computation times.
+All these aspects have been thoroughly discussed.
 
-Line 83: For alfalfa one could use the assembly of Medicago Trancatula – was this considered?
+> Line 83: For alfalfa one could use the assembly of Medicago Trancatula – was this considered?
 
-ANSWER:
+ANSWER: it is now :)
 
-Line 86: Why these four general methods?
+> Line 86: Why these four general methods?
 
-ANSWER:
+ANSWER: as answered above, those are very well known, available in many software libraries, and already proven to bear good performances (except MNI, included as reference).
 
-Line 109: How to interpret the 437 plants from 391 accessions as these accessions are supposed to be fully homozygous? Were there duplicates?
+> Line 109: How to interpret the 437 plants from 391 accessions as these accessions are supposed to be fully homozygous? Were there duplicates?
 
-ANSWER: yes, they're duplicates
+ANSWER: yes, they're duplicates. We state it more clearly now.
 
-Line 134: Why were these values (4 and 11) were chosen? I do not see the rational of taking 4 for a heterozygote and 11 for homozygote - shouldn’t these be the reverse?
+> Line 134: Why were these values (4 and 11) were chosen? I do not see the rational of taking 4 for a heterozygote and 11 for homozygote - shouldn’t these be the reverse?
 
-ANSWER:
+ANSWER: this is a common practice in GBS SNP calling.
+We updated the text, added references to other works doing the same, and added a section in supplementary material explaining a bit of the related math.
+The short answer is that no, it's in general more difficult to call a homozygote, so you want to have more reads.
 
-Line 135-139: The reduction from 5 genotypic classes to 3 genotypic classes in alfalfa is a critical simplification step.
+> Line 135-139: The reduction from 5 genotypic classes to 3 genotypic classes in alfalfa is a critical simplification step.
 There is no justification why this step was executed.
@@
 
-ANSWER:
+ANSWER: the 5-to-3 reduction is not related to algorithms (the general ones would work without a blink) but to GBS read depth of the available data.
+To be able to call the heterozygote dosage it is required a high number of reads per locus, and accepting only the markers with enough reads highly reduces their number.
+We explained the choice in the proper M&M section and added reference to papers addressing it.
 
-Line 166:  How to interpret the ‘K closest markers’ in case there is no map or assembly information (as here for alfalfa)?
+> Line 166:  How to interpret the ‘K closest markers’ in case there is no map or assembly information (as here for alfalfa)?
 
@@
 
-Line 169: How was distance specified, i.e., in bp or cM?  Did you try both and if so, where there differences?
+> Line 169: How was distance specified, i.e., in bp or cM?  Did you try both and if so, where there differences?
 
-ANSWER: again, this is not the phyisical distance between markers, but the distance between observations based on a function of the SNPs (see above)
+ANSWER: again, this is not the physical distance between markers, but the distance between observations based on a function of the SNPs (see above)
 
-Line 173-181: I wonder whether this explanation is appropriate for this audience.
+> Line 173-181: I wonder whether this explanation is appropriate for this audience.
 
@@
 
-Line 175: Were the eigenvectors of M’M and MM’ merged to arrive at the ‘first k eigenvectors’?
+> Line 175: Were the eigenvectors of M’M and MM’ merged to arrive at the ‘first k eigenvectors’?
 
@@
 
-Line 178: Which value for k was used, was there a decision criteria used?
+> Line 178: Which value for k was used, was there a decision criteria used?
 
-ANSWER: [che diciamo qui? Seguiamo Rutkoski (o che per lei)?]
+ANSWER: we used K = 4. After some trials (data not shown) we found that no big difference are found in the 3-20 range. Also, K = 4 is the same used by Rutkoski et al.
 
-Line 182: For the SVD, how critical is the initial imputation step by MNI?
-Were other methods considered for this initial imputation?
-[I think this is just an initialization prior, and the least informative the better]
+> Line 182: For the SVD, how critical is the initial imputation step by MNI?  Were other methods considered for this initial imputation?
 
@@
 
-Line 185 – 198: Explain the ‘hat’ for U in equation 2. Also, why has Y no hat while there is a hat on the Y in equation 3.
+> Line 185 – 198: Explain the ‘hat’ for U in equation 2. Also, why has Y no hat while there is a hat on the Y in equation 3.
 
@@
 
-Line 217: Elaborate on the content of this likelihood file or provide a reference (to the manual of Beagle)?
+> Line 217: Elaborate on the content of this likelihood file or provide a reference (to the manual of Beagle)?
 
@@
 
-Line 229: What was the reason to not go beyond 20% artificially induced missing values?
+> Line 229: What was the reason to not go beyond 20% artificially induced missing values?
 
@@
 
-Line 254: It is not clear whether the ‘total number of missing data’ refers to the sum of ‘natural’ and ‘artificial’ missing data or only to the latter part. Please improve description.
+> Line 254: It is not clear whether the ‘total number of missing data’ refers to the sum of ‘natural’ and ‘artificial’ missing data or only to the latter part. Please improve description.
 
@@
 
-Line 255: It may be confusing to use y in equation 4 as a discrete variable since Y is continuous in equations  [2] and [3]. It seems more intuitive to use g instead of y.
+> Line 255: It may be confusing to use y in equation 4 as a discrete variable since Y is continuous in equations  [2] and [3]. It seems more intuitive to use g instead of y.
 
@@
 
-Line 281/Table 1 & Table 2:  Why not merge the content of these two tables?
+> Line 281/Table 1 & Table 2:  Why not merge the content of these two tables?
 Transpose Table 1 to have the same columns.
@@
 
-ANSWER:
+ANSWER: we tried to follow the reviewer suggestion, but ended up with a table that is, in our opinion, harder to read. We consider the two small tables are handier and more to the point.
 
-Line 282: Why not present these rates as percentages as that is probably easier to read and more consistent (with e.g., line 286).
+> Line 282: Why not present these rates as percentages as that is probably easier to read and more consistent (with e.g., line 286).
 
@@
 
-Line 284: Figure S1 does not contain information at the ‘per individual’ level
+> Line 284: Figure S1 does not contain information at the ‘per individual’ level
 
-ANSWER
+ANSWER: updated the text.
+
+===================================================================
+===================================================================
+===================================================================
+                           FINO A QUI
+===================================================================
+===================================================================
+===================================================================
 

### 2016-02-09 inputs/abstract.tex
integrating notes from Fil+N pass

@@
 large amounts of missing marker genotypes, whose imputation becomes therefore 
-both challenging and critical to ensure reliability of later analyses.\\
+both challenging and critical for later analyses.\\
 In this work, the performance of four general imputation methods (K-Nearest 

### 2016-02-10 reviews/answer_to_revieweres.txt
alignment stat and answers to reviewers

@@
 
-Line 313/Figure 3: Suggest to maintain consistency in y-axis scale, i.e., the plots for Minor homozygous accuracy should range from 0 to 100% 
+> Line 313/Figure 3: Suggest to maintain consistency in y-axis scale, i.e., the plots for Minor homozygous accuracy should range from 0 to 100% 
 as well.
@@
 
-ANSWER : to be checked
+ANSWER :

### 2016-02-10 inputs/results.tex
alignment stat and answers to reviewers

@@
 be included progressively more with more relaxed thresholds on allowed missing level.
-The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.
-\textcolor{red}{
-Alignment on reference genome assigned a position to 88.66\%, xxx\%, and xxx\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively.
-}
+The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.\\
+Alignment on reference genome assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively.
 
@@
 \label{sec:imputation_accuracy}
-\textcolor{red}{
-\begin{itemize}
-  \item all rice chromosome gave the same results, as in Suppl figures
-  \item all numbers in this section need to be checked
-  \item what about supplementary figures? still valid?
-\end{itemize}
-}
+
 In rice, the average imputation accuracy over all genotypic classes was above 80\% for all methods.

### 2016-02-10 reviews/answer_to_revieweres.txt
completed answers to reviewers

@@
 
-===================================================================
-===================================================================
-===================================================================
-                           FINO A QUI
-===================================================================
-===================================================================
-===================================================================
+> Line 304: How sensitive are the results to the imposed quality filters?
 
-Line 304: How sensitive are the results to the imposed quality filters?
-
-ANSWER:
+ANSWER: not very sensitive.
+Quality filters have a direct effect on the elaboration steps following imputation (GWAS in particular) but little impact on imputation itself.
+Only very relaxed filtering would change imputation performances, since a lot of noisy data point would be included.
+This scenario is however very far from standard practice and of little interest for the current work.
 
@@
 
-Line 324: the flat response seems to hold for the studied range (0.01 – 0.20).
+> Line 324: the flat response seems to hold for the studied range (0.01 – 0.20).
 Were higher values of missing rates studied?
@@
 
-ANSWER: [diplomatic answer to be added]
+ANSWER: the cited range (0.01 - 0.20) is of injected missing data points, and it overlaps the other range (0.10 - 0.70) of naturally occurring missing data points.
+This brings the maximum accepted level (worst case scenario) to 0.2 + 0.7 = 0.9, which seems a good place where to stop.
+Please keep in mind that published works using GBS data usually filter on the 0.3-0.5 range, so we think that our choice abundantly covers the current practice.
 
-Line 347: Rather extensive reporting on computation time which seems not justified as the used methods were used as these were available to the authors, i.e.., there was not dedicated efforts to speed up the implemented methods.
+> Line 347: Rather extensive reporting on computation time which seems not justified as the used methods were used as these were available to the authors, i.e.., there was not dedicated efforts to speed up the implemented methods.
 
 ANSWER: This section has been re-written, shortened where possible, and a statement, in the discussion, on the limitations of the information on computation time was added.
-[Need to add results for FILLIN --> Nelson made the plot]
 
-Line 379-381: Where was this negative impact shown?  In this study or other studies – please include references.
+> Line 379-381: Where was this negative impact shown?  In this study or other studies – please include references.
 
-ANSWER: a reference to Rutkoski et al was added;we tried to explain better the difference between SNP array (~5% missing) and GBS data (~50% missing)
+ANSWER: a reference to Rutkoski et al was added. We tried to explain better the difference between SNP array (~5% missing) and GBS data (~50% missing)
 
-Line 401-407: This is still rather descriptive without any postulation whether any of the methods are less or more sensitive to MAF and/or balance.
+> Line 401-407: This is still rather descriptive without any postulation whether any of the methods are less or more sensitive to MAF and/or balance.
 The superior results of Beagle in rice are mostly driven by utilizing map/assembly information which is ignored by other methods.
 
-ANSWER
+ANSWER We updated the section, improving the description and commenting the results of an extra method that uses marker information.
 
-Line 440: The reasoning on factor ii) is not clear – would the results for rice be less accurate when analysing the genome at once? Also, why would it simplify the imputation process?
+> Line 440: The reasoning on factor ii) is not clear – would the results for rice be less accurate when analysing the genome at once? Also, why would it simplify the imputation process?
 
 ANSWER: First, the per-chromosome analysis simplifies calculations, and allow for straightforward parallelization.
-Additionally, the results with rice as a whole are less accurate when markers are reshuffle [add numbers/supplementary Table/Figure?]; this is because the LD structure is better preserved when reshuffling markers.
+Additionally, the results with rice as a whole are less accurate when markers are shuffled [add numbers/supplementary Table/Figure?]; this is because the LD structure is better preserved when reshuffling markers.
 
-Line 443: Did you explore the population structure of the three datasets? Would presence/absence of (severe) population structure affect imputation accuracy?
+> Line 443: Did you explore the population structure of the three datasets? Would presence/absence of (severe) population structure affect imputation accuracy?
 
-Answer: i) no structure in alfalfa, presence of structure in rice (indica/tropical/temperate); ii) implicitely exploited by most imputation methods (e.g.
+Answer: i) no structure in alfalfa, presence of structure in rice (indica/tropical/temperate); ii) implicitly exploited by most imputation methods (e.g.
 KNN --> distance based on genetic relatedness); iii) not likely to affect much imputation accuracy; iv) this is not GWAS: the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed;
 
-Line 450: It is too easy to state that there is additional complexity to make use of a reference genome of a related species.
+> Line 450: It is too easy to state that there is additional complexity to make use of a reference genome of a related species.
 It would be of HIGH relevance to explore this as it may be applied in many species that lack a reference genome themselves.
 
-ANSWER:
+ANSWER: we added the study using Medicago Truncatula.
 

### 2016-02-10 inputs/discussion.tex
phrasing

@@
 However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning on a different species (and with different ploidity) comes at the price of discarding those markers that do not align.
-\textcolor{red}{HOW MANY?}\\
-
+In our case only 57.54\% (Alfalfa-Med) and 57.86\% (Alfalfa-PV) aligned on the \emph{M.
+Truncatula} genome, compared to the 88.66\% of rice markers aligning on \emph{O.
+Sativa} genome, a loss of markers that could or could not be acceptable, depending on the specific application requirements.
+\\
 Among the imputation methods used in this study, Beagle and FILLIN are the only ones that make use of marker information, and were therefore tested with ordered and randomly shuffled markers.

### 2016-02-10 data/extract_averages.R
data for average stats, and text updates

@@ -1,0 +1,23 @@
+df = read.csv('medie.csv')
+
+# [1] beagle unphased           beagle unphased (shuffle) kNNI                     
+# [4] MNI                       RFI_pvariables            SVD                      
+# [7] fillin                    fillin (shuffle)   
+# 
+# alfalfa    rice       rice_chrom
+
+res = NULL
+for (p in unique(df$population)){
+for (a in unique(df$algorithm)){
+ res = rbind(res, data.frame(
+   pop = p,
+   alg = a,
+   tot = mean(subset(df, population == p & algorithm == a)$totAccuracy),
+   min = mean(subset(df, population == p & algorithm == a)$minHom),
+   maj = mean(subset(df, population == p & algorithm == a)$majHom),
+   het = mean(subset(df, population == p & algorithm == a)$hetAcc)
+   
+   
+ ))
+}}
+

### 2016-02-11 reviews/answer_to_revieweres.txt
new answers

@@
 
-ANSWER: the 5-to-3 reduction is not related to algorithms (the general ones would work without a blink) but to GBS read depth of the available data.
+ANSWER: the 5-to-3 reduction is not strictly related to algorithms.
+In fact, the general algorithms would work easily on any number of classes: MNI, SVDI and RFI would produce continuous values in the [0-4] range, and KNNI is naturally oriented toward a discrete outputs.
+Beagle and FILLIN, on the other hand, are less flexible, and using them on tetraploid genotypes is not straightforward.
+However, the main reason for the simplification of heterozygotes is the available GBS read depth.
 To be able to call the heterozygote dosage it is required a high number of reads per locus, and accepting only the markers with enough reads highly reduces their number.
-We explained the choice in the proper M&M section and added reference to papers addressing it.
+We explained the choice in the proper M&M section and added reference to papers following this (common) practice.
 

### 2016-02-11 data/extract_averages.R

@@ -21,3 +21,4 @@ for (a in unique(df$algorithm)){
  ))
 }}
 
+mean(subset(df, population == 'rice_chrom')$majHom)

### 2016-02-11 inputs/abstract.tex

@@
 and two genotype-specific method (``Beagle'' and FILLIN) were measured 
-on GBS data from alfalfa (\emph{Medicago sativa}, autotetraploid, heterozygous, 
+on GBS data from alfalfa (\emph{Medicago sativa L.}, autotetraploid, heterozygous, 
 without reference genome) and 
-rice (\emph{Oryza sativa}, diploid, 100\% homozygous, with reference genome). Benchmarks 
+rice (\emph{Oryza sativa L.}, diploid, 100\% homozygous, with reference genome). Benchmarks 
 consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 

### 2016-02-11 inputs/results.tex

@@
 
-The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging \textcolor{red}{XXX}\% and \textcolor{red}{XXX}\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging \textcolor{red}{XXX}\% and \textcolor{red}{XXX}\% in alfalfa and rice, respectively).
+The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 92.27\% and 96.47\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 5.79\% and 69.88\% in alfalfa and rice, respectively).
 Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}).
@@
 As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
-In alfalfa KNNI showed the highest imputation accuracy over all classes \textcolor{red}{(XXX\%)} and in the heterozygous \textcolor{red}{(XXX\%)} and minor homozygous \textcolor{red}{(XXX\%)} genotype classes (Figure~\ref{fig:accuracy_alfalfa}).
+In alfalfa KNNI showed the highest imputation accuracy over all classes (averaging 82.20\%) and in the heterozygous (66.02\%) and minor homozygous (14.68\%) genotype classes (Figure~\ref{fig:accuracy_alfalfa}).
 RFI and SVDI were second ranking.
-They did not differ much from MNI in the overall accuracy \textcolor{red}{(XXX\%)} and in the major \textcolor{red}{(XXX\%)} homozygous class, but gave higher imputation accuracy in the heterozygous \textcolor{red}{(XXX\%)} and minor homozygous \textcolor{red}{(XXX\%)} classes.
+They did not differ much from MNI in the overall accuracy (MNI: 78.38\%, RFI: 79.50\%, SVDI: 77.35\%) and in the major (MNI: 88.68\%, RFI: 89.05\%, SVDI 85.81\%) homozygous class, but gave higher imputation accuracy in the heterozygous (MNI: 58.87\%, RFI: 61.75\%, SVDI 61.67\%).
+On minor homozygous class only SVDI resulted in some sporadic correct imputation, while RFI and MNI did not produce any sensible result (MNI: 0\%, RFI: 0.01\%, SVDI 4.46\%) classes.
 

### 2016-02-11 inputs/supplementary.tex
phrasing

@@
 $HOM \rightarrow HET$ errors can not happen, since a homozigote will always produce 100\% of reads of the same type).
-When accounting sequencing errors and reads quality a term for $HOM \rightarrow HET$ appears, due to reads sequences as the wrong allele.
+When accounting sequencing errors and reads quality a term for $HOM \rightarrow HET$ appears, due to reads sequenced as the wrong allele.
 In the presented study we thus applied a threshold of 4 required reads per stack to call a homozigote to minimize this kind of error.
@@
 
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig02_Alfalfa-Med.pdf}\caption{
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_reforma.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for alfalfa population Alfalfa-Med.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig03_Alfalfa-PV.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_lodi.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for alfalfa population Alfalfa-PV.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig04_Rice-chrom-1.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_1.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 1 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig05_Rice-chrom-2.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_2.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 2 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig06_Rice-chrom-3.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_3.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 3 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig07_Rice-chrom-4.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_4.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 4 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig08_Rice-chrom-5.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_5.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 5 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig09_Rice-chrom-6.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_6.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 6 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig10_Rice-chrom-7.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_7.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 7 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig11_Rice-chrom-8.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_8.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 8 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig12_Rice-chrom-9.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_9.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 9 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig13_Rice-chrom-10.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_10.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 10 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig14_Rice-chrom-11.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_11.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 11 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig15_Rice-chrom-12.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice_chrom_12.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for rice chromosome 12 data.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
-\begin{figure}\includegraphics[width=0.95\textwidth]{SupplFig16_Rice.pdf}\caption{
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\begin{figure}\includegraphics[width=0.95\textwidth]{figure_rice.pdf}\caption{
 imputation accuracies overall, for the major homozygous genotype (AA) and for the minor homozygous genotype (BB) in datasets consisting of
@@
 additional missing values artificially introduced (x-axis) for whole rice data set.
-Lines colors represent the five imputation algorithms: MNI
-(orange), KNNI (red), SVDI (blue), RFI (green) and Beagle (black)}\end{figure}
+Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}

### 2016-02-11 data/extract_averages.R
new compact averages

@@ -22,3 +22,10 @@ for (a in unique(df$algorithm)){
 }}
 
 mean(subset(df, population == 'rice_chrom')$majHom)
+mean(subset(df, population == 'rice')$majHom)
+
+mean(subset(df, population == 'rice_chrom')$minHom)
+mean(subset(df, population == 'rice')$minHom)
+
+mean(subset(df, population == 'rice_chrom')$totAccuracy)
+mean(subset(df, population == 'rice')$totAccuracy)

### 2016-02-11 reviews/answer_to_revieweres.txt
answer to SVDI

@@
 
-ANSWER: the first k eigenvectors were determined based on the ordered corresponding eigenvalues. The matrix U* contained these k eigenvector
+ANSWER: the columns of matrix U contain the eigenvectors of MM' (n x n matrix), the columns of matrix V contain the eigenvectors of M'M (m x m matrix).
+The n diagonal elements of matrix Sigma (n x m) are the square root of the eigenvalues of matrix U (n x n) (which should correspond also to the non-zero eigenvalues of V' / M'M).
+The first k eigenvectors of U were then determined based on the ordered corresponding eigenvalues.
+The matrix U* (n x k) contains these k eigenvectors, which multiplied by the k regression coefficients in vector beta return the (n x 1) column vector of estimated genotypes at locus j.
 

### 2016-02-11 reviews/answer_to_revieweres.txt
pop structure

@@
 
-Answer: i) no structure in alfalfa, presence of structure in rice (indica/tropical/temperate); ii) implicitly exploited by most imputation methods (e.g.
-KNN --> distance based on genetic relatedness); iii) not likely to affect much imputation accuracy; iv) this is not GWAS: the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed;
+Answer: There was no population structure in alfalfa (see Annicchiarico et al.
+2015).
+On the other hand, there was population structure in rice (indica, tropical and temperate japonica).
+Most imputation methods implicitly exploit population structure (e.g.
+KNN --> distance based on genetic relatedness; Beagle --> haplotype reconstructed based on genetic similarities), without explicitly modelling it.
+Imputation of missing genotypes is not GWAS, where the significance and estimate of SNP effects may be inflated if population structure is not included in the model), and the impact on the accuracy of imputation is likely to be small.
+In cattle from Scandinavian countries, Brondum et al (2012) actually found higher imputation accuracy when combining populations (cattle breeds) in Beagle (without explicitly modelling the population structure), most likely as a result of the larger sample size.
+However, the influence of population structure on the accuracy of imputation was not formally tested in this work.
+Though interesting, we feel that this is outside of the scope of the present paper, which already contains a large amount of comparisons and results.
+These considerations were added to the discussion.
 

### 2016-02-11 reviews/answer_to_revieweres.txt
several adds after afternoon pass

--- 2016-02-11 reviews/answer_to_revieweres.txt
@@
 ANSWER: the mentioned papers are indeed relevant, and were cited in the introduction of the paper.
-However, none of the cited papers focus on algorithm comparison, nor examine the effect of different levels of call rate.
+However, none of the cited papers focus on algorithm comparison, nor examine the effect of different levels of call rate and missing rate.
 
@@
 
-ANSWER: We added a reference to a review paper by Ben Hayes et al, 2009, in animal genomics, and cited other common GWAS and GS software libraries (BGLR, rrBLUP, GenABEL) that contain imputation functionalities or straightforward refuse input containing missing data points.
-Especially linear algebra and arithmetic methods methods suffer from missing data, whereas numerical methods like MCMC can better cope with them.
+ANSWER: We added a reference to a review paper by Ben Hayes et al, 2009, in animal genomics, and cited other common GWAS and GS software libraries (BGLR, rrBLUP, GenABEL) that refuse input containing missing data points or that contain imputation functionalities to prepare data for analysis.
+Especially linear algebra and arithmetic methods suffer from missing data, whereas numerical methods like MCMC can better cope with them.
 In any case, using missing data, is not impossible: e.g.
@@
 The paper results are consistent both internally and with the cited literature, and we think shine some light on the examined topic, even without the use of many different datasets.
+We however understand the reasoning in the reviewer's question, and decided to adjust the declared target of the paper to be less general and focus more on the studied species.
+The change of title goes in that direction, as much as a big part of the updated text.
 
@@
 
-ANSWER: given the very different mathematical models underlying the examined algorithms (or all algorithms, for what it's worth) we think that the essential properties of a good imputation algorithm are discussed ones, namely: accuracy (total and per class); resilience to growing missing level; computation times.
+ANSWER: given the very different mathematical models underlying the examined algorithms (or all imputation algorithms, for what it's worth) we think that the essential properties of a good imputation algorithm are already discussed, namely: accuracy (total and per class); resilience to growing missing level; computation times.
 All these aspects have been thoroughly discussed.
@@
 However, the main reason for the simplification of heterozygotes is the available GBS read depth.
-To be able to call the heterozygote dosage it is required a high number of reads per locus, and accepting only the markers with enough reads highly reduces their number.
+To be able to call the heterozygote dosage a high number of reads per locus is required, and accepting only the markers with enough reads highly reduces their number.
 We explained the choice in the proper M&M section and added reference to papers following this (common) practice.
@@
 
-ANSWER: First, the per-chromosome analysis simplifies calculations, and allow for straightforward parallelization.
-Additionally, the results with rice as a whole are less accurate when markers are shuffled [add numbers/supplementary Table/Figure?]; this is because the LD structure is better preserved when reshuffling markers.
+ANSWER: First, the per-chromosome analysis simplifies calculations, and allows for straightforward parallelization.
+Additionally, the results with rice as a whole are less accurate when markers are shuffled.
+In fact, accuracy losses are:
+
+- total: -6% on total rice, -3% on chromosomes
+- major homozigote: -2% on total rice, -1% on chromosomes
+- minor homozigote: -41% on total rice, -22% on chromosomes
+
+This happens because the LD structure is better preserved when reshuffling markers within chromosomes.
 
@@
 
-Answer: i) no structure in alfalfa, presence of structure in rice (indica/tropical/temperate); ii) implicitly exploited by most imputation methods (e.g.
-KNN --> distance based on genetic relatedness); iii) not likely to affect much imputation accuracy; iv) this is not GWAS: the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed;
+ANSWER: we updated the manuscript to discuss the issue of population structure. In brief:
+
+i) there is no population structure in alfalfa, as reported by published study. Rice has structure, though, since the dataset is a collation of several subpopulations (indica/tropical/temperate)
+ii) population structure is implicitly exploited by most imputation methods, in particular those performing data clusterization (e.g. KNN --> distance based on genetic relatedness) 
+iii) it is however not likely to affect much imputation accuracy.
+This is not GWAS, and the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed.
 
@@
 
-Line 535-540: Too general/vague and thus not clear why this paragraph is included in the conclusion section (or even in the manuscript)! [
+> Line 535-540: Too general/vague and thus not clear why this paragraph is included in the conclusion section (or even in the manuscript)!
 
-ANSWER: To be checked when the new draft is ready.
+ANSWER: We updated the section to provide a more incisive conclusion.
 
-Figure 1: Be consistent in using rates versus percentages, i.e., this Figure uses rates while on Line 311 you refer to percentages.
+> Figure 1: Be consistent in using rates versus percentages, i.e., this Figure uses rates while on Line 311 you refer to percentages.
 

### 2016-02-11 inputs/abstract.tex
several adds after afternoon pass

--- 2016-02-09 inputs/abstract.tex
@@
 Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
-and two genotype-specific method (``Beagle'' and FILLIN) were measured 
-on GBS data from alfalfa (\emph{Medicago sativa}, autotetraploid, heterozygous, 
+and two genotype-specific methods (``Beagle'' and FILLIN) were measured 
+on GBS data from alfalfa (\emph{Medicago sativa L.}, autotetraploid, heterozygous, 
 without reference genome) and 
-rice (\emph{Oryza sativa}, diploid, 100\% homozygous, with reference genome). Benchmarks 
+rice (\emph{Oryza sativa L.}, diploid, 100\% homozygous, with reference genome). Benchmarks 
 consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
@@
 genotypes, globally and within each genotype class (two homozygotes in rice, 
-two homozygotes and one heterozygote in alfalfa). Imputation efficiency was 
-measured with computation times.\\
-In general imputation accuracy was robust to increasing 
+two homozygotes and one heterozygote in alfalfa).\\
+we found that imputation accuracy was robust to increasing 
 missing rates, and consistently higher in rice than in alfalfa. Accuracy 
 was as high as 90-100\% for the major (most frequent) homozygous genotype, but dropped 
-to 80-90\% (rice) and below 30\% (alfalfa) on the minor homozygous genotype. 
+to 80-90\% (rice) and below 30\% (alfalfa) in the minor homozygous genotype. 
 Beagle was the best performing method, both accuracy- 

### 2016-02-11 inputs/conclusions.tex
several adds after afternoon pass

@@
 In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency.
-The alignment of markers to closely related species can help, but comes at the price of discarding the non-aligning markers.
-Moreover, performances indicate that general methods are still the best option.\\
+The alignment of markers to closely related species can help, but comes at the price of discarding markers that do not align, and with general methods still performing better.\\
 The imputation accuracy tended to be relatively robust over increasing missing rates.
@@
 In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\
-Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.
+
+\textcolor{red}{TO BE DELETED?\\Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.}
+
+Exploring ways to improve imputation accuracy in GBS data is an important research area and can help making genotyping-by-sequencing a very cost-effective technique.
+In this paper we shown how general imputation algorithms are a valid alternative to genotype-oriented ones when reference genomes are not available or too expensive in terms of unaligned markers.
+Moreover, we highlighted the importance of examining the accuracy on minority classes, since the common practice of summarizing performances in a single index can easily mislead to undesired results.

### 2016-02-11 inputs/discussion.tex
several adds after afternoon pass

@@
 With larger amounts of missing data the complexity of the imputation problem increases.
-Imputation error can have a negative impact on successive analyses (e.g.
+Imputation errors can have a negative impact on successive analyses (e.g.
 on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
@@
 \label{sec:role_of_species}
-Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
+Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
 \cite{the_1000_genomes_project_consortium_integrated_2012}
@@
 i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
-ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling phase. This semplification step brought alfalfa dataset farther from biology;\\
-iii) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process;\\
-iv) rice markers were aligned on native genome, while alfalfa markers were aligned on a different species.
+ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling.
+This semplification step made alfalfa data less adherent to the underlying biology;\\
+iii) rice data have higher marker density (2.4~Kbp/SNP, compared to 24.5~Kbp/SNP for Alfalfa-PV and 19.6~Kbp/SNP for Alfalfa-Med) due to both a higher number of markers and a smaller genome (400~Mb for rice, 800~Mb for Alfalfa).
+This allowed for higher average values of linkage disequilibrium (LD) among SNP markers, thus facilitating the imputation process;\\
+iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on a different species.
 
@@
 The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms).
-This behaviour is probably due to FILLIN being designed to target inbred lines (as the name itself indicates).
+This behaviour is probably due to FILLIN being specifically designed to target inbred lines (as the name itself indicates).
 
@@
 \label{sec:reference_genome_ordered_vs_unordered}
-When a reference genome is available, markers positions can be used in the imputation process.
+When a reference genome is available, marker position can be used in the imputation process.
 All methods specifically developed for the imputation of missing genotypes have been designed for markers ordered along the genome sequence, and make explicit or implicit use of position information.
@@
 For alfalfa there is no reference genome sequenced yet.
-The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and can in principle used.
+The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and can in principle be used.
 However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning on a different species (and with different ploidity) comes at the price of discarding those markers that do not align.
@@
 Truncatula} genome, compared to the 88.66\% of rice markers aligning on \emph{O.
-Sativa} genome, a loss of markers that could or could not be acceptable, depending on the specific application requirements.
+Sativa} genome.
+This left with 23\,438 and 18\,923 SNPs to be used for analysis with ordered markers.
 \\
@@
 Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 75\% (10\% allowed, 1\% artificial missing rates) to almost 50\% (70\% allowed, 20\% artificial missing rates).
-When tested with alfalfa Beagle largerly overestimated by the majority class and had very poor performances on heterozygous class.
-Shuffling marker positions further worsened the accuracy, and almost all missing data point assigned to the majority class.\\
-FILLIN performances on rice follow the same pattern and worsen substantially when markers are shuffled.
+When tested with alfalfa Beagle largerly overestimated the majority class and had very poor performances on the heterozygous class.
+Shuffling marker positions further worsened Beagle performance in alfalfa, and almost all missing genotypes were assigned to the majority class.\\
+FILLIN performance in rice followed the same pattern and was substantially worse when markers were shuffled.
 This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.

### 2016-02-11 inputs/introduction.tex
several adds after afternoon pass

@@
 Imputation of missing GBS data without alignment to the reference genome has
-been reported in alfalfa (``Medicago sativa'', \cite{Rocher_validation_2015})
-and red raspberry (``Rubus idaeus'', \cite{ward_saturated_2013}). These studies
+been reported in alfalfa (\emph{Medicago sativa L.}, \cite{Rocher_validation_2015})
+and red raspberry (\emph{Rubus idaeus L.}, \cite{ward_saturated_2013}). These studies
 however do not focus on measuring imputation accuracy and do not compare
-alternative imputation models.\\
+alternative imputation methods.\\
 	%un po' una ripetizione. Togliere?
@@
 for genotype imputation (localized haplotype clustering imputation, 
-implemented in the software package Beagle, and a method of haplotype
-reconstruction around recombination sites, implemented in the FILLIN algorithm
-as part of Tassel suite). The four general algorithms were chosen as well known 
+implemented in the software package Beagle, and the method of based on haplotype
+reconstruction around recombination sites implemented in the FILLIN algorithm
+as part of the Tassel suite). The four general algorithms were chosen as well known 
 imputation strategies implemented in several freely available software libraries.

### 2016-02-11 inputs/results.tex
several adds after afternoon pass

@@
 The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets.
-Supplementary Figure S1 shows the density distribution of missing rate per markers in the complete datasets.
+Supplementary Figure S1 shows the density distribution of missing rate per marker in the complete datasets.
 The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%).
@@
 In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes.
-This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous loci.
+This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous locus.
 Therefore homozygous loci tended to have higher missing rates, and to
-be included progressively more with more relaxed thresholds on allowed missing level.
+be included progressively more frequently with more relaxed thresholds on allowed missing level.
 The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.\\
-Alignment on reference genome assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively.
+Alignment on reference genomes assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively, reflecting the difference between having a reference genome available for the exact same species (rice) or only for a close one (alfalfa).
+
+dsahe different proportion of unaligned markers directly reflect the .
 
@@
 
-In rice, the average imputation accuracy over all genotypic classes was above 80\% for all methods.
+In rice, the average imputation accuracy over all genotype classes was above 80\% for all methods.
 No significant difference was found between chromosomes.
@@
 The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\%.
-No significant difference were found between the two datasets.
+No significant differences were found between the two datasets.
 Thus average accuracy over the Alfalfa-Med and Alfalfa-PV is presented (Figure~\ref{fig:accuracy_alfalfa}).
 
-The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype class (averaging 92.27\% and 96.47\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 5.79\% and 69.88\% in alfalfa and rice, respectively).
-Alfalfa datasets included heterozygous SNP loci, accounting for an average of 30\% of individuals per locus (Table~\ref{tab:descriptive_statistics}).
+The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype (averaging 92.27\% and 96.47\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 5.79\% and 69.88\% in alfalfa and rice, respectively).
+Alfalfa datasets included heterozygous SNP loci, accounting for an average of 27\% of individuals per locus (Table~\ref{tab:descriptive_statistics}).
 Heterozygotes had an intermediate imputation accuracy ranging from 0\% with MNI to 70\% with KNNI.
-Results for each individual alfalfa population and the 12 individual rice chromosomes are similar to those discussed here (see Supplementary Figures S2 to S16).
+Results for each individual alfalfa population and the 12 individual rice chromosomes are similar to those discussed here (see Supplementary Figures S2 to S15).
 \\
-Missing data did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate.
-KNNI showed a decreasing imputation accuracy with increasing missing data, both in alfalfa and rice.
+Missing rate did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate.
+KNNI showed a decreasing imputation accuracy with increasing missing rate, both in alfalfa and rice.
 
@@
 When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle.
-The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular on heterozygous and minor homozygous classes.\\
+The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular in heterozygous and minor homozygous genotypes.\\
 Similarly, FILLIN resulted in a 5-10\% accuracy drop when marker positions were shuffled.
-FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as wrong imputation.
-FILLIN did not impute an average of 14.6\% missing data in Rice and practically the entirety of Alfalfa data.
-Even trying to relax FILLIN constraint parameters, it was impossible to obtain sensible results.
-Thus, FILLIN accuracy does not appear in Figure~\ref{fig:accuracy_alfalfa}.
+FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as imputation errors.
+On average, FILLIN did not impute 14.6\% of missing data in Rice and practically the entirety of Alfalfa data.
+Even trying to relax FILLIN parameters, it was impossible to obtain sensible results in alfalfa.
+Thus, FILLIN accuracy results are not reported in Figure~\ref{fig:accuracy_alfalfa}.
 
 When examining general imputation methods, resulting accuracies highlighted RFI and KNNI as the best ones.
-The two methods had comparable performances, without a preferable option over all experiments.
+The two methods had comparable performances over all experiments.
 SVDI followed, and MNI ranked last.\\
-In Rice, for 70\% allowed missing rate datasets, the imputation accuracy of RFI was still greater than 90\%, while it dropped towards 75\% for KNNI and SVDI.
+In Rice, for 70\% allowed missing rate, the imputation accuracy of RFI was still higher than 90\%, while it dropped towards 85\% for KNNI and SVDI.
 As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
@@
 They did not differ much from MNI in the overall accuracy (MNI: 78.38\%, RFI: 79.50\%, SVDI: 77.35\%) and in the major (MNI: 88.68\%, RFI: 89.05\%, SVDI 85.81\%) homozygous class, but gave higher imputation accuracy in the heterozygous (MNI: 58.87\%, RFI: 61.75\%, SVDI 61.67\%).
-On minor homozygous class only SVDI resulted in some sporadic correct imputation, while RFI and MNI did not produce any sensible result (MNI: 0\%, RFI: 0.01\%, SVDI 4.46\%) classes.
+In the minor homozygous class only SVDI resulted in some sporadic correct imputation, while RFI and MNI did not produce any sensible result (MNI: 0\%, RFI: 0.01\%, SVDI 4.46\%) classes.
 

### 2016-02-11 data/extract_averages.R

@@ -21,6 +21,8 @@ for (a in unique(df$algorithm)){
  ))
 }}
 
+write.csv(x = res, file='compact_res.csv', row.names = FALSE)
+
 mean(subset(df, population == 'rice_chrom')$majHom)
 mean(subset(df, population == 'rice')$majHom)
 

### 2016-02-11 reviews/answer_to_revieweres.txt

@@
 
-ANSWER: we updated the manuscript to discuss the issue of population structure. In brief:
-
-i) there is no population structure in alfalfa, as reported by published study. Rice has structure, though, since the dataset is a collation of several subpopulations (indica/tropical/temperate)
-ii) population structure is implicitly exploited by most imputation methods, in particular those performing data clusterization (e.g. KNN --> distance based on genetic relatedness) 
-iii) it is however not likely to affect much imputation accuracy.
-This is not GWAS, and the interest is not in estimating the effect of the SNP (which may be inflated if other factors are not included in the model), but in "prediction/classification" of the missing genotype to be imputed.
+Answer: There was no population structure in alfalfa (see Annicchiarico et al.
+2015).
+On the other hand, there was population structure in rice (indica, tropical and temperate japonica).
+Most imputation methods implicitly exploit population structure (e.g.
+KNN --> distance based on genetic relatedness; Beagle --> haplotype reconstructed based on genetic similarities), without explicitly modelling it.
+Imputation of missing genotypes is not GWAS, where the significance and estimate of SNP effects may be inflated if population structure is not included in the model), and the impact on the accuracy of imputation is likely to be small.
+In cattle from Scandinavian countries, Brondum et al (2012) actually found higher imputation accuracy when combining populations (cattle breeds) in Beagle (without explicitly modelling the population structure), most likely as a result of the larger sample size.
+However, the influence of population structure on the accuracy of imputation was not formally tested in this work.
+Though interesting, we feel that this is outside of the scope of the present paper, which already contains a large amount of comparisons and results.
+These considerations were added to the discussion.
 

### 2016-02-11 inputs/materials_and_methods.tex

@@
 \label{par:SVDI}
-: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (m markers, n individuals):
+: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (n individuals, m markers):
 
@@
 
-where U is a $m \times m$ unitary matrix (i.e.
-$UU^{T}=I$), $\Sigma$ is a $m \times n$ rectangular diagonal matrix of singular values and $V^{T}$ is the $n \times n$ conjugate transpose of the unitary matrix V.
-Matrix U and V contain the eigenvectors of $M^{T}M$ and $MM^{T}$, respectively, and the corresponding non-zero singular values in $\Sigma$ are equivalent to the square-root of the eigenvalues of $M^{T}M$ and $MM^{T}$.
-The first k eigenvectors - ordered by decreasing eigenvalue - capture most of the information in the marker genotype matrix, and were used to generate linear combinations (principal components) of the original m markers for the imputation of missing data points.
+where U is a $n \times n$ unitary matrix (i.e.
+$UU^{T}=I$), $\Sigma$ is a $n \times m$ rectangular diagonal matrix of singular values and $V^{T}$ is the $m \times m$ conjugate transpose of the unitary matrix V.
+The columns of matrix U and matrix V contain the eigenvectors of $MM^{T}_{(n \times n)}$ and $M^{T}M_{(m \times m)}$, respectively, and the corresponding non-zero singular values in $\Sigma$ are equivalent to the square-root of the non-zero eigenvalues of $MM^{T}$ and $M^{T}M$.
+The first k eigenvectors in U - ordered by decreasing eigenvalue (from $\Sigma$) - capture most of the information in the marker genotype matrix M, and were used to generate linear combinations (principal components) of the original m markers for the imputation of missing data points.
 The imputation procedure comprised: (i) initial imputation of missing genotypes using MNI, since SVD can only be performed on complete matrices; (ii) SVD to select the most informative k eigenvectors of the marker genotype matrix; (iii) these k eigenvectors were used as predictors in a linear regression model for marker genotypes:
@@
 
-with Y as vector of genotypes at marker j; $U^*$ matrix of the first k eigenvectors; $\beta$ vector of regression coefficients; $\varepsilon$ random errors terms; (iv) all eigenvectors (matrix U) and $\beta$ were used to estimate missing values at marker j.
+with Y as vector of n genotypes at marker j; $U^*$ the $n \times k$ matrix of the first k eigenvectors; $\beta$ the vector of $k$ regression coefficients; $\varepsilon$ the random error terms; (iv) all eigenvectors (matrix $U^*$) and $\beta$ were used to estimate missing values at marker j.
 The procedure was repeated (steps (ii) to (iv)) until convergence.

### 2016-02-11 inputs/discussion.tex
fil_last

@@
 
-Alfalfa datasets did not present population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulations.
-Population structure is implicitly used by algorithms like Beagle, FILLIN and KNNI, that create clusters of similar genotypes, and it is likely to affect imputation accuracy.\\
-The most dramatic difference between datasets was found with FILLIN.
+%
+%The most dramatic difference between datasets was found with FILLIN.
 The algorithm costantly produced 100\% uninputed results in alfalfa, while resulted in average performances in rice (albeit inferior to the other algorithms).
 This behaviour is probably due to FILLIN being specifically designed to target inbred lines (as the name itself indicates).
+%
+
+There was no population structure in alfalfa \cite{Annicchiarico2015}, while the rice dataset was intrinsically stratified ---being a collection of five subpopulations (\textit{indica}, \textit{temperate japonica} \textit{tropical japonica}, \textit{aus} and \textit{aromatic}).
+Most imputation methods implicitly exploit population structure (e.g.
+KNNI computes distances based on genetic relatedness; Beagle reconstructs haplotypes based on genetic similarities), without explicitly modelling it.
+The imputation of missing genotypes, however, is not an inferential problem such as GWAS (genome-wide association study), where the significance and estimate of SNP effects are known to be inflated if population structure is not included in the model, and the impact on the accuracy of imputation is likely to be small.
+In cattle from Scandinavian countries, Brondum et al \cite{brondum2012short} actually found higher imputation accuracy when combining populations (cattle breeds) in Beagle (without explicitly modelling the population structure), most likely as a result of the larger sample size.
+On the other hand, population stratification may help explain why FILLIN performed poorly in our rice dataset, which is a collection of subpopulations, while this algorithm is optimised for homogeneous inbred populations.
+However, the influence of population structure on the accuracy of imputation was not formally tested in this work, and this remains an interesting topic for further investigation.
 

### 2016-02-11 inputs/table_accuracies.tex
new table with accuracies

\begin{table}
\centering
\caption[Accuracy statistics]{some long caption}
\label{tab:accuracy_statistics}
\begin{tabular}{llcccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Dataset} & \textbf{Method} & \textbf{Total} & \textbf{Minor (BB)} & \textbf{Major (AA)} & \textbf{Heterozygote (AB)}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
alfalfa & Beagle & 73.14 & 14.27 & 99.37 & 20.77\\
alfalfa & Beagle (shuffle) & 66.34 & 1.31 & 99.88 & 0.50\\
alfalfa & kNNI & 82.20 & 14.68 & 90.82 & 66.03\\
alfalfa & MNI & 78.38 & 0.00 & 88.69 & 58.88\\
alfalfa & RFI  & 79.50 & 0.00 & 89.05 & 61.75\\
alfalfa & SVDI & 77.36 & 4.47 & 85.81 & 61.67\\
alfalfa & FILLIN & * & * & * & *\\
alfalfa & FILLIN (shuffle)  & * & * & * & *\\
rice (chr) & Beagle & 99.58 & 97.50 & 99.80 & -\\
rice (chr) & Beagle (shuffle) & 96.09 & 74.73 & 98.42 & -\\
rice (chr) & kNNI & 98.10 & 89.12 & 99.04 & -\\
rice (chr) & MNI & 90.19 & 1.06 & 99.89 & -\\
rice (chr) & RFI  & 98.81 & 91.51 & 99.58 & -\\
rice (chr) & SVDI & 97.81 & 85.64 & 99.11 & -\\
rice (chr) & FILLIN & 86.96 & 63.62 & 90.14 & -\\
rice (chr) & FILLIN (shuffle)  & 82.19 & 55.86 & 85.81 & -\\
rice & Beagle & 99.58 & 97.57 & 99.80 & -\\
rice & Beagle (shuffle) & 93.48 & 55.64 & 97.69 & -\\
rice & kNNI & 98.30 & 89.80 & 99.19 & -\\
rice & MNI & 90.21 & 0.61 & 99.89 & -\\
rice & RFI  & 97.86 & 84.68 & 99.28 & -\\
rice & SVDI & 95.74 & 70.96 & 98.40 & -\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

### 2016-02-11 inputs/results.tex
new table with accuracies
new table for accuracies

@@
 
+\input{inputs/table_accuracies}
+
 In rice, the average imputation accuracy over all genotype classes was above 80\% for all methods.
@@
 \\
-The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 84\%.
+The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 82\%.
 No significant differences were found between the two datasets.
@@
 Alfalfa datasets included heterozygous SNP loci, accounting for an average of 27\% of individuals per locus (Table~\ref{tab:descriptive_statistics}).
-Heterozygotes had an intermediate imputation accuracy ranging from 0\% with MNI to 70\% with KNNI.
+Heterozygotes had an intermediate imputation accuracy ranging from 0.5\% with Beagle to 66.03\% with KNNI.
 Results for each individual alfalfa population and the 12 individual rice chromosomes are similar to those discussed here (see Supplementary Figures S2 to S15).
@@
 The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular in heterozygous and minor homozygous genotypes.\\
-Similarly, FILLIN resulted in a 5-10\% accuracy drop when marker positions were shuffled.
+Similarly, FILLIN resulted in a 5-8\% accuracy drop when marker positions were shuffled.
 FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as imputation errors.
@@
 As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
-In alfalfa KNNI showed the highest imputation accuracy over all classes (averaging 82.20\%) and in the heterozygous (66.02\%) and minor homozygous (14.68\%) genotype classes (Figure~\ref{fig:accuracy_alfalfa}).
+In alfalfa KNNI showed the highest imputation accuracy over all classes (averaging 82.20\%) and in the heterozygous (66.03\%) and minor homozygous (14.68\%) genotype classes (Figure~\ref{fig:accuracy_alfalfa}).
 RFI and SVDI were second ranking.
-They did not differ much from MNI in the overall accuracy (MNI: 78.38\%, RFI: 79.50\%, SVDI: 77.35\%) and in the major (MNI: 88.68\%, RFI: 89.05\%, SVDI 85.81\%) homozygous class, but gave higher imputation accuracy in the heterozygous (MNI: 58.87\%, RFI: 61.75\%, SVDI 61.67\%).
-In the minor homozygous class only SVDI resulted in some sporadic correct imputation, while RFI and MNI did not produce any sensible result (MNI: 0\%, RFI: 0.01\%, SVDI 4.46\%) classes.
+They did not differ much from MNI in the overall accuracy (MNI: 78.38\%, RFI: 79.50\%, SVDI: 77.36\%) and in the major (MNI: 88.69\%, RFI: 89.05\%, SVDI 85.81\%) homozygous class, but gave higher imputation accuracy in the heterozygous (MNI: 58.88\%, RFI: 61.75\%, SVDI 61.67\%).
+In the minor homozygous class only SVDI resulted in some sporadic correct imputation, while RFI and MNI did not produce any sensible result (MNI: 0\%, RFI: 0.01\%, SVDI 4.47\%) classes.
 

### 2016-02-11 inputs/table_accuracies.tex
new table for accuracies

@@
 \centering
-\caption[Accuracy statistics]{some long caption}
+\caption[Accuracy statistics]{Imputation accuracies (as percentages) for the various
+datasets and methods, averaged over all allowed and injected missing thresholds.
+Rice data are reported averaged by chromosome and on the complete genotypes.
+Alfalfa data are reported averaged over the two Alfalfa-PV and Alfalfa-Med populations.
+FILLIN failed to impute alfalfa datasets.}
 \label{tab:accuracy_statistics}
@@
 alfalfa & MNI & 78.38 & 0.00 & 88.69 & 58.88\\
-alfalfa & RFI  & 79.50 & 0.00 & 89.05 & 61.75\\
+alfalfa & RFI  & 79.50 & 0.01 & 89.05 & 61.75\\
 alfalfa & SVDI & 77.36 & 4.47 & 85.81 & 61.67\\
-alfalfa & FILLIN & * & * & * & *\\
-alfalfa & FILLIN (shuffle)  & * & * & * & *\\
+alfalfa & FILLIN & * & * & * & -\\
+alfalfa & FILLIN (shuffle)  & * & * & * & -\\
 rice (chr) & Beagle & 99.58 & 97.50 & 99.80 & -\\

### 2016-02-11 inputs/conclusions.tex
conclusions+table

@@
 
-\textcolor{red}{TO BE DELETED?\\Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.}
+%Exploring ways to improve imputation accuracy in GBS data, for instance by combining predictive methods as in boosting or model ensembling, while at the same time increasing algorithm efficiency to keep computation times reasonably low, is an important research area, and can help making low-cost genotyping-by-sequencing a very cost-effective technique for genomic applications in species with and without a reference genome.}
 
-Exploring ways to improve imputation accuracy in GBS data is an important research area and can help making genotyping-by-sequencing a very cost-effective technique.
-In this paper we shown how general imputation algorithms are a valid alternative to genotype-oriented ones when reference genomes are not available or too expensive in terms of unaligned markers.
-Moreover, we highlighted the importance of examining the accuracy on minority classes, since the common practice of summarizing performances in a single index can easily mislead to undesired results.
+The results of this paper showed that high imputation accuracies can be achieved with GBS data, and that general imputation methods are a valid option when the reference genome is not available or the alignment on the genome of a closely related species leads to losing too many markers.
+Additionally, we highlighted the importance of examining imputation accuracy in the different genotype classes, since the common practice of summarizing performances in a single index can be very misleading when data are unbalanced (i.e.
+MAF is low).
+Exploring ways to improve imputation accuracy in GBS data is an important research area and can help making genotyping-by-sequencing a very attractive and cost-effective technique for genomic applications in species with and without a reference genome.

### 2016-02-11 inputs/table_accuracies.tex
conclusions+table

@@
 alfalfa & SVDI & 77.36 & 4.47 & 85.81 & 61.67\\
-alfalfa & FILLIN & * & * & * & -\\
-alfalfa & FILLIN (shuffle)  & * & * & * & -\\
+alfalfa & FILLIN & * & * & * & *\\
+alfalfa & FILLIN (shuffle)  & * & * & * & *\\
 rice (chr) & Beagle & 99.58 & 97.50 & 99.80 & -\\

### 2016-02-11 inputs/abstract.tex
abstract

@@
 without reference genome) and 
-rice (\emph{Oryza sativa L.}, diploid, 100\% homozygous, with reference genome). Benchmarks 
+rice (\emph{Oryza sativa L.}, diploid, 100\% homozygous, with reference genome). Alfalfa SNP were aligned on the genome of the closely related species \emph{Medicago truncatula L.}. Benchmarks 
 consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
@@
 two homozygotes and one heterozygote in alfalfa).\\
-we found that imputation accuracy was robust to increasing 
+We found that imputation accuracy was robust to increasing 
 missing rates, and consistently higher in rice than in alfalfa. Accuracy 
@@
 Beagle was the best performing method, both accuracy- 
-and time-wise, when using a reference genome. When no reference 
-genome was available, KNNI and RFI gave the highest accuracies, with KNNI 
-being much faster.
+and time-wise, in rice. In alfalfa, KNNI and RFI gave the highest accuracies, but KNNI 
+was much faster.
 

### 2016-02-11 reviews/answer_to_revieweres.txt
last answer to reviewer
letztes mal

@@
 
-ANSWER: sentence modified to "Most methods for genotype imputation have been applied to SNP array data, and typically yield very high imputation accuracy"
+ANSWER: sentence modified to "Most methods for genotype imputation have been applied to SNP array data, and typically yield very high imputation accuracy". [see P3, L55-56]
 
@@
 
-ANSWER: it was assembled in 2005, and then recently updated (2013). We made this clearer in the text.
+ANSWER: it was assembled in 2005, and then recently updated (2013). We made this clearer in the text. [P4, L88]
 
@@
 
-ANSWER: We followed the reviewer advice and added the results aligning on Medicago truncatula genome.
+ANSWER: We followed the reviewer advice and added the results of aligning on Medicago truncatula genome.
 
@@
 
-ANSWER: we added several references to software sources.
+ANSWER: we added several references to software sources. [e.g. reference to a review on software for (animal) genomics, + reference to all software used in the paper: e.g. Beagle, Tassel, R packages]
 
@@
 The matrix U* (n x k) contains these k eigenvectors, which multiplied by the k regression coefficients in vector beta return the (n x 1) column vector of estimated genotypes at locus j.
+This has been made clearer in the text [P8, L189-199]
 
@@
 
-ANSWER: we used K = 4. After some trials (data not shown) we found that no big difference are found in the 3-20 range. Also, K = 4 is the same used by Rutkoski et al.
+ANSWER: we used K = 4.
+After some trials (data not shown) we found that no big difference are found in the 3-20 range.
+Also, K = 4 is the same used by Rutkoski et al..
+This was added to the text [P9, L209-211]
 
@@
 
-ANSWER: more details were added to the text, plus a reference to the Beagle Manual
+ANSWER: more details were added to the text, plus a reference to the Beagle Manual [P10, L236-240]
 
@@
 
-ANSWER: it refers to the total number of artificially (introduced) missing genotypes. This was specified in the text.
+ANSWER: it refers to the total number of artificially (introduced) missing genotypes. This was specified in the text. [P12, L282]
 
@@
 
-ANSWER: updated the text.
+ANSWER: updated the text. [P13, L317]
 
@@
 
-ANSWER: a reference to Rutkoski et al was added. We tried to explain better the difference between SNP array (~5% missing) and GBS data (~50% missing)
+ANSWER: references to Rutkoski et al  and to Annicchiarico et al were added. We tried to explain better the difference between SNP array (~5% missing) and GBS data (~50% missing) [P18, L432-435]
 
@@
 These considerations were added to the discussion.
+[P20, L499-517]
 
@@
 
-ANSWER: We updated the section to provide a more incisive conclusion.
+ANSWER: We updated the section to provide a more incisive conclusion. [P24, L603-613]
 
@@
 
-ANSWER :
+ANSWER : figure and text had been updated accordingly.

### 2016-02-11 inputs/acknowledgements.tex
acknowledge

@@
 \begin{acknowledgements}
-The creation of the rice data set was financially supported by the Marie Curie European
-Reintegration Grant ``NEUTRADAPT''.
+The rice data used in this research paper were produced within the framework of the Italian national project ``RISINNOVA'' (grant n. 2010-2369), financially supported by the AGER Foundation..
 The creation of the alfalfa data sets was funded by the projects ‘Genomic selection in alfalfa (GENALFA)’ funded by the Italian Ministry of Foreign Affairs and International Cooperation in the framework of the Italy-USA scientific cooperation program, the Italian share of the FP7-ArimNet project ‘Resilient, water- and energy-efficient forage and feed crops for Mediterranean agricultural systems (REFORMA)’ funded by the Italian Ministry of Agricultural and Forestry Policies.

### 2016-05-04 Nazzicari_Imputation.tex
aggregato

\begin{filecontents*}{example.eps}
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
\RequirePackage{fix-cm}
\documentclass[smallextended]{svjour3}       \smartqed  
\usepackage{xfrac}   
\usepackage{lineno}  
\usepackage{graphicx}
\usepackage{color}
\usepackage{multirow}
\usepackage{array}
\usepackage{setspace} 
\usepackage{makecell}
\usepackage[nomarkers]{endfloat}

\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}

\linenumbers
\doublespacing 
\journalname{Molecular Breeding}
\begin{document}

\title{
Marker imputation efficiency for Genotyping-By-Sequencing data in rice (\emph{Oryza sativa}) and alfalfa (\emph{Medicago sativa})
\thanks{Nelson Nazzicari and Filippo Biscarini contributed equally to the work \\
Corresponding author: Nelson Nazzicari \email{nelson.nazzicari@crea.gov.it}}
}

\titlerunning{Imputation efficiency for GBS data with and without the reference genome}

\author{Nelson Nazzicari \and
  Filippo Biscarini  \and
  Paolo Cozzi \and
  E. Charles Brummer \and
  Paolo Annicchiarico
}

\authorrunning{Nazzicari, Biscarini et al.} 
\institute{
    N. Nazzicari, P. Annicchiarico \at Council for Agricultural Research and Economics (CREA) Research Centre for Fodder Crops and Dairy Productions, Lodi (Italy)
\and
    F. Biscarini, P. Cozzi \at Dipartimento di Bioinformatica, Fondazione Parco Tecnologico Padano, Lodi (Italy)
\and
    E.C. Brummer \at University of California, Plant Sciences Department, Davis, CA (USA)
}

\date{Received:  / Accepted:}

\maketitle

\graphicspath{{./figures/}}
\makeatletter{}\begin{abstract}
Genotyping-by-sequencing (GBS) is a rapid and cost-effective genome-wide 
genotyping technique applicable whether a reference genome is available 
or not. Due to the cost-coverage trade-off, however, GBS typically produces 
large amounts of missing marker genotypes, whose imputation becomes therefore 
both challenging and critical for later analyses.\\
In this work, the performance of four general imputation methods (K-Nearest 
Neighbors, Random Forest, Singular Value Decomposition, and Mean Value)
and two genotype-specific methods (``Beagle'' and FILLIN) were measured 
on GBS data from alfalfa (\emph{Medicago sativa L.}, autotetraploid, heterozygous, 
without reference genome) and 
rice (\emph{Oryza sativa L.}, diploid, 100\% homozygous, with reference genome). Alfalfa SNP were aligned on the genome of the closely related species \emph{Medicago truncatula L.}. Benchmarks 
consisted in progressive data filtering for marker call-rate (up to 70\%) and increasing 
proportions (up to 20\%) of known genotypes masked for imputation. The 
relative performance was measured as the total proportion of correctly imputed 
genotypes, globally and within each genotype class (two homozygotes in rice, 
two homozygotes and one heterozygote in alfalfa).\\
We found that imputation accuracy was robust to increasing 
missing rates, and consistently higher in rice than in alfalfa. Accuracy 
was as high as 90-100\% for the major (most frequent) homozygous genotype, but dropped 
to 80-90\% (rice) and below 30\% (alfalfa) in the minor homozygous genotype. 
Beagle was the best performing method, both accuracy- 
and time-wise, in rice. In alfalfa, KNNI and RFI gave the highest accuracies, but KNNI 
was much faster.

\keywords{SNP, Genotyping by Sequencing (GBS), K-Nearest Neighbors Imputation 
(KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation 
(SVDI), Beagle, FILLIN, alfalfa, rice, imputation, reference genome}
\end{abstract}

\makeatletter{}\section{Introduction}
\label{intro}

Imputation of missing alleles and genotypes is a preliminary step for
a wide range of genetic analyses. In fact most models and software for 
population genetics, genomic selection (GS) and genome-wide association 
studies (GWAS) can not easily handle missing data and require complete datasets (e.g. \cite{hayes2009invited,aulchenko2007genabel,endelman2011rrblup,perez2014genome}). 
SNP-array represent one of the leading genotyping technologies and produce 
datasets that, after low call-rate filtering, 
usually still contain a small proportion of uncalled genotypes (e.g., $<$5\%) 
randomly distributed along the genome. Genotyping-by-sequencing (GBS) is a 
relatively recent technique which is considered a viable alternative to SNP array-based genotyping to produce SNP genotype data~\cite{elshire_robust_2011}.
GBS is platform-independent and is conveniently 
used in species that lack commercial SNP-chips or even lack a reference 
genome sequence. GBS data typically present a much larger proportion of 
sporadic missing genotypes: e.g., $\sim$52\% and $\sim 58\%$ average in two maize experimental populations \cite{crossa_genomic_2013},
or ``up to 80\% missing data per marker'' in wheat \cite{poland_genomic_2012}. This 
is due to both intrinsic properties of the technology
and to its application to species without a mature genome assembly \cite{glaubitz_tassel-gbs:_2014}.\\
There are several methods that are routinely used for the imputation of 
missing genotypes. Some rely solely on genotypic information, some make 
use also of genealogies, most of them are based on reconstructing haplotypes 
to be used in predictive models \cite{nicolazzi_software_2015}. Most methods for 
genotype imputation have been applied to SNP array data, and typically 
yield very high imputation accuracy. For instance, $>$95\% correctly imputed 
genotypes were reported in maize \cite{hickey_factors_2012} and cattle \cite{vanraden_genomic_2011}. 
The same imputation methods can also be applied to GBS data (see \cite{huang_efficient_2014} and \cite{swarts_novel_2014} for applications in rice and maize). When a reference 
genome is available, SNP loci can be aligned against its sequence and 
are therefore ordered, thus allowing the exploitation of local linkage disequilibrium (LD).
However, GBS data may be generated 
also for species lacking a reference genome assembly. In this case the 
GBS output is a series of ``floating'' loci not linked to any chromosome:
alignment is not possible and SNPs are therefore considered unordered. Unordered 
data add additional complexity to genotype imputation. Exploitation of linkage 
disequilibrium and haplotype reconstruction 
are less straightforward, consecutive loci may lie on different chromosomes
or scaffolds, and overall data are less homogeneous.
Rutkoski et al.~\cite{rutkoski_imputation_2013}
considered the application of general data imputation methods 
to wheat, maize and barley populations, obtaining best case scenarios accuracies 
as high as 0.84-0.94 (measured as $R^{2}$ between true and imputed genotypes).
Imputation of missing GBS data without alignment to the reference genome has
been reported in alfalfa (\emph{Medicago sativa L.}, \cite{Rocher_validation_2015})
and red raspberry (\emph{Rubus idaeus L.}, \cite{ward_saturated_2013}). These studies
however do not focus on measuring imputation accuracy and do not compare
alternative imputation methods.\\
		Compared to SNP-array data, GBS is more challenging: many more 
	missing genotypes, high variability between runs, intrinsically noisy data, 
	reads from different loci that can overlap. For all these reasons, imputation 
	of missing genotypes with GBS data is still a relatively immature technique, 
	not yet amenable to produce highly standardised and repeatable results. 
There is therefore scientific and practical interest in gaining further 
insights into how genotype imputation with GBS data works, and in 
developing the best methods and strategies to accurately and efficiently 
impute such missing data. This would be relevant not only for the scientific 
community but also for the breeding industry, because of its direct 
consequences on genomic applications.\\
In this paper, we imputed missing genotypes from GBS data in two agronomically
important crop species:
the cereal crop rice (\emph{Oryza sativa L.}) whose reference genome has been 
assembled in 2005 \cite{international_rice_genome_sequencing_project_map-based_2005},
and updated in 2013 \cite{kawahara_improvement_2013},
and
the forage crop alfalfa (\emph{Medicago sativa L.}), for 
which a reference genome is not available yet. Alfalfa genotypes
were mapped on the close relative \emph{Medicago truncatula L.}, a model species
for legumes, whose genome is available \cite{young_medicago_2011}.
The species were chosen as representative of very different scenarios, alfalfa being autotetraploid
with high heterozygosity and rice being diploid with essentially 100\% homozygozity.
The present study thus encompasses a wide span of real application cases.\\
We compared the imputation accuracy of four general imputation methods (Mean
Value Imputation, K-Nearest Neighbor Imputation, Singular Value Decomposition
Imputation and Random Forest Imputation) with the performance of two methods specific 
for genotype imputation (localized haplotype clustering imputation, 
implemented in the software package Beagle, and the method of based on haplotype
reconstruction around recombination sites implemented in the FILLIN algorithm
as part of the Tassel suite). The four general algorithms were chosen as well known 
imputation strategies implemented in several freely available software libraries.
The two genotype-specific methods represent the state of the art 
for genotypes lacking pedigree information.\\
The relative efficiency of the different imputation methods was assessed in terms of 
accuracy and computation time. Accuracy was measured as fraction of 
correctly imputed genotypes, both as a total and within genotype classes 
(major/minor homozigotes, and heterozigotes, when present).

\makeatletter{}\section{Materials and methods}

\subsection{Plant material}
\label{sec:plant_material}

Samples from two autotetraploid alfalfa (\emph{Medicago sativa L.}) and one diploid rice (\emph{Oryza sativa L.}) populations were available (see Table~\ref{tab:descriptive_statistics}).\\
Alfalfa populations included elite germplasms from the Po Valley (Alfalfa-PV) and Mediterranean-climate environments (Alfalfa-Med). For Alfalfa-PV, 124 parent genotypes were chosen by stratified mass selection for dry matter yield over three harvests. For Alfalfa-Med, 154 parent genotypes were derived from two cycles of free intercrossing among three outstanding populations in a multi-environment study. For further details see \cite{Annicchiarico2015}. \\
The Rice dataset included 437 plants belonging to 391 accessions (46 duplicates) from the Rice Germplasm Collection maintained at CRA-Rice Research Unit (Vercelli, Italy). The sampled collection comprised accessions from the five main sub-populations of \emph{O. sativa} (274 temperate japonica, 108 tropical japonica, 28 indica, 16 aus and 11 aromatic). All accessions were purified through single seed descent, and were genotypically essentially 100\% homozygous.

\subsection{Genotyping by sequencing}
\label{sec:overview}

The 715 samples from alfalfa (Alfalfa-Med and Alfalfa-PV) and rice (Rice) populations were genotyped using the genotyping-by-sequencing (GBS) approach \cite{elshire_robust_2011}. Different GBS protocols were used to genotype the alfalfa and rice samples, since genotyping was carried out under different projects.\\
In alfalfa, DNA was isolated from fresh leaf tissues by the Wizard~\textregistered~Genomic DNA Purification Kit (Promega, A1125) and quantified with a Quant-iT PicoGreen dsDNA assay kit (Life Technologies, P7589). Two libraries were constructed for the two populations, where 100 ng of each DNA was digested with ApeKI (NEB, R0643L) and then ligated to a unique barcoded adapter and a common adapter. For the reference population Alfalfa-Med, 5 nmoles each of the primers and NEB 2X Taq Master Mix (NEB Cat \# M0270S) were included in the PCR reaction according to \cite{elshire_robust_2011} original protocol, while for Alfalfa-PV the KAPA library amplification readymix (Kapa Biosystems Cat \# KK2611) was used instead. Each library was sequenced in two lanes on Illumina HiSeq 2000 at the Genomic Sequencing and Analysis Facility at the University of Texas at Austin, Texas, USA. For both populations post sequencing analysis and SNP calling was carried out using Tassel UNEAK pipeline \cite{lu_switchgrass_2013}.

Alfalfa is an autotetraploid plant species and can therefore present three different heterozygous genotypes: AAAB, AABB and ABBB. Sequencing depth in this study was not sufficient for accurate tetraploid allelic dosage, but following \cite{li_saturated_2014} and \cite{li_genomic_2015} reliable genotype calls based on diploid allelic dosage were obtained considering diploid heterozygotes (i.e. AB), while the two homozygous genotypes (AAAA and BBBB) were considered diploid homozygotes (i.e., AA or BB). A further quality filter, implemented through ad-hoc Python scripts, removed heterozygous loci with less than 4 and homozygous loci with less than 11 aligned reads. A similar filtering was performed in \cite{Rocher_validation_2015} using less restrictive thresholds.

In rice, DNA was isolated from three-weeks old leaves using the DNeasy Plant Mini Kit (QIAGEN) with a TECAN Freedom EVO150 liquid handling robot (TECAN Group Ltd, Switzerland). DNA digestion was performed on 100 ng samples in 96-well plates using ApeKI, which was shown to cut every 1 kb on average in a in-silico digestion of the Nipponbare reference genome, and 96-plex libraries constructed according to the GBS protocol. The libraries were loaded into Genome Analyzer II (Illumina, Inc., San Diego, CA) for sequencing. Raw sequence data filtering, sequence alignment to the rice reference genome (Os-Nipponbare-Reference-IRGSP-1.0, \cite{kawahara_improvement_2013}), and SNP calling from GBS genotyping were carried out using the Tassel GBS pipeline \cite{glaubitz_tassel-gbs:_2014}.\\
In total, 32\,706, 40\,734 and 166\,418 SNP markers were called by GBS in Alfalfa-PV, Alfalfa-Med and Rice, respectively (see Table~\ref{tab:descriptive_statistics}). In all datasets SNP variants were renamed so that AA represented the most common homozygote, and BB the least common homozygote.

\makeatletter{}\begin{table}
\centering
\caption[Descriptive statistics]{
Descriptive statistics of alfalfa (\emph{Medicago sativa L.}) 
and rice (\emph{Oryza sativa L.}) genotyping. MAF is average minor allele frequency; 
p(AA), p(AB) and p(BB) are the proportions of AA, AB and BB genotypes.}
\label{tab:descriptive_statistics}
\begin{tabular}{lccccccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Dataset} & \textbf{No. samples} & \textbf{No. SNPs} & \textbf{Missing Rate} & \textbf{MAF} & \textbf{p(AA)} & \textbf{p(AB)} & \textbf{p(BB)}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
Alfalfa-PV & 124 & 32\,706 & 0.666 & 0.1724 & 0.690 & 0.274 & 0.035\\
Alfalfa-Med & 154 & 40\,734 & 0.596 & 0.1702 & 0.691 & 0.276 & 0.032\\
Rice & 437 & 166\,418 & 0.534 & 0.140 & 0.860 & - & 0.140 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\subsection{Imputation methods}
\label{sec:imputation_methods}
We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbors Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and haplotype reconstruction around the recombination sites (FILLIN from Tassel suite). For all algorithms we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively. 
\paragraph{MNI}
\label{par:MNI}
: in Mean Value Imputation each missing data point was replaced by the mean of the non-missing values for that marker, then discretized to the closest value in \{0,1,2\}.
\paragraph{KNNI}
\label{par:KNNI}
: in K-Nearest Neighbors Imputation missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001} defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data. $K=4$ was used in KNNI, and neighbo
rs values were weighted by the reciprocal of their distance from the data point to be imputed.
\paragraph{SVDI}
\label{par:SVDI}
: Singular Value Decomposition (SVD) imputation was based on the following factorization of the genotype matrix M (n individuals, m markers):

\begin{equation}
\label{eq:SVDI_general}
M = U\Sigma V^{T}
\end{equation}

where U is a $n \times n$ unitary matrix (i.e. $UU^{T}=I$), $\Sigma$ is a $n \times m$ rectangular diagonal matrix of singular values and $V^{T}$ is the $m \times m$ conjugate transpose of the unitary matrix V. The columns of matrix U and matrix V contain the eigenvectors of $MM^{T}_{(n \times n)}$ and $M^{T}M_{(m \times m)}$, respectively, and the corresponding non-zero singular values in $\Sigma$ are equivalent to the square-root of the non-zero eigenvalues of $MM^{T}$ and $M^{T}M$. The first k eigenvectors in U - ordered by decreasing eigenvalue (from $\Sigma$) - capture most of the information in the marker genotype matrix M, and were used to generate linear combinations (principal components) of the original m markers for the imputation of missing data points. The imputation procedure comprised: (i) initial imputation of missing genotypes using MNI, since SVD can only be performed on complete matrices; (ii) SVD to select the most informative k eigenvectors of the marker genotype matrix; (iii) these k eigenvectors were used as predictors in a linear regression model for marker genotypes:

\begin{equation}
\label{eq:SVDI_regression}
Y = U^*\beta +\varepsilon 
\end{equation}

with Y as vector of n genotypes at marker j; $U^*$ the $n \times k$ matrix of the first k eigenvectors; $\beta$ the vector of $k$ regression coefficients; $\varepsilon$ the random error terms; (iv) all eigenvectors (matrix $U^*$) and $\beta$ were used to estimate missing values at marker j. The procedure was repeated (steps (ii) to (iv)) until convergence. The final imputed data points were then discretized to the closer genotype in \{0,1,2\}. A value of $k=4$ for the eigenvectors to be selected in $U^*$ was used in our implementation of SVDI, based on empirical results in the range $3--20$ (data not shown). Additional details on SVDI can be found in \cite{troyanskaya_missing_2001}

\paragraph{RFI}
\label{par:RFI}
: in Random Forest (RF) imputation, missing genotypes at marker j were imputed by means of RF multiple regression trees \cite{breiman_random_2001} where all markers other than j were used for the prediction. At each marker j, 100 RF regression trees $\Theta_{1} \ldots \Theta_{100}$ were grown from a bootstrapped sample of the individuals in Y and a random subset of $\sqrt{m-1}$ markers. Missing genotypes were imputed averaging predictions over the 100 RF trees

\begin{equation}
\label{eq:RFI_regression}
\hat{Y} = \frac{1}{100}\sum_{i=1}^{100} h(x,\Theta_i)
\end{equation}

where $h(x,\Theta_i)$ is a function of the $i_{th}$ tree and subset of markers. RFI was repeated until convergence or for maximum 10 iterations. After regression, the imputed data were then discretized to the closer genotype in \{0,1,2\}.

\paragraph{Beagle}
\label{par:Beagle}
: ``localized haplotype clustering imputation'' is a method implemented in the software ``Beagle'' \cite{browning_rapid_2007}. Originally developed for human genetics, Beagle has since found wide application also in animal and plant genetics. Beagle has become so popular that the name of the software is commonly used to refer metonymically to the method it implements, making the two hardly distinguishable. Beagle infers haplotypes and imputes sporadic missing alleles both with known and unknown phase, using a localized haplotype cluster model. This is a class of directed acyclic graphs which empirically models haplotype frequencies on a local scale and therefore adapts to local structure in the data. Beagle makes use of a hidden Markov model to find the most likely haplotype pair for each individual, given the genotype data for that individual and the graphical haplotype frequency model. The method works iteratively using an expectation-maximization (EM) approach. The imputed missing data, probabilities of missing genotypes and inferred haplotypes are calculated from the model that is fitted in the last iteration. For the imputation of missing genotypes in the completely homozygous rice accessions, a likelihood file (with the prior likelihood of each of the three possible SNP genotypes) was defined to specifically exclude the imputation of non-possible heterozygous genotypes (details in \cite{beagle3_manual}).

\paragraph{FILLIN}
The ``Fast Inbred Line Library ImputatioN'' (``FILLIN'') is an imputation method optimised for inbred populations implemented in the ``Tassel'' programming suite \cite{swarts_novel_2014}. FILLIN is based on haplotype reconstruction around recombination break points. Haplotypes are clustered per genotype similarity using the Hamming distance function. This information is eventually used to impute the target locus in an iterative approach that attemps, through a Markov Chain MonteCarlo (MCMC) process, to maximise the likelihood of the observed SNP calls given the unobserved imputed genotypes. If convergence criteria are not met genotypes are left uninputed. Among the considered algorithms, FILLIN is the only one that can thus produce partially imputed results.

\subsection{Imputation procedure}
\label{sec:imputation_procedure}
In order to assess the imputation performance of the different methods, we introduced increasing proportions of artificial missing genotypes in the data sets. These were then imputed with the various algorithms, measuring the resulting imputation accuracy and computation time. We extracted four datasets from the overall data that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker. For each of these datasets, we randomly introduced an additional 1\%, 5\%, 10\% or 20\% missing genotypes, on which imputation accuracy could be measured. Table~\ref{tab:missing_rates} reports the number of markers and average missing rate (before introducing artificial missing genotypes) for the four call-rate thresholds. This procedure produced 16 combinations for each population: 4 call rate datasets $\times$ 4 levels of artificial missing markers. On each combination, we applied the 6 described imputation methods. Rice data were analysed as a whole and by each of the 12 chromosomes separately. In absence of alfalfa genome, the \emph{M. Truncatula} genome was used as reference for those algorithms (Beagle and FILLIN) requiring markers positions.

To further investigate the effect of presence (or absence) of genome information we generated ``reshuffled'' datasets where marker positions were randomly assigned inside each chromosome. This way, the linkage disequilibrium between markers is broken and the relative performance decrease can be directly assessed. The reshuffled datasets were tested only on Beagle and FILLIN.

\makeatletter{}
\begin{table}
\centering
\caption[Markers counts and missing rates]{
Total number of markers and resulting average missing rate for the four different allowed missing rate thresholds in alfalfa (two datasets) and rice.}
\label{tab:missing_rates}
\begin{tabular}{C{3cm}rccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Allowed missing rate threshold} &  & \textbf{Alfalfa-PV} & \textbf{Alfalfa-Med} & \textbf{Rice}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
   \multirow{2}{*}{70\%} & Missing rate &  0.325 & 0.29  & 0.364\\
                           & Markers      & 13\,190  & 19\,986 & 109\,372\\    
\noalign{\smallskip}\hline
   \multirow{2}{*}{40\%} & Missing rate &  0.161 & 0.142 & 0.198\\
                           & Markers      & 7\,790 & 12\,931 & 58\,553\\    
\noalign{\smallskip}\hline
   \multirow{2}{*}{20\%} & Missing rate & 0.076 & 0.074 & 0.099\\
                           & Markers      & 4\,828 & 8\,962 & 29\,872\\
\noalign{\smallskip}\hline                           
   \multirow{2}{*}{10\%} & Missing rate & 0.046 & 0.045 & 0.049\\
                           & Markers      & 3\,405 & 6\,364 & 15\,060 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\subsection{Imputation accuracy and efficiency assessment}
\label{sec:imputation_accuracy_and_efficiency_assessment}
We used the artificial missing genotypes to measure the performance of the six imputation methods. The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele). Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class. The imputation accuracy was computed as the number of missing data correctly imputed divided by the total number of artificially missing data (proportion of correct imputations):

\begin{equation}
\label{eq:accuracy}
accuracy = \frac{1}{n} \sum_{i=1}^{n}{I(g_i = \hat{g_i})}
\end{equation}

where $I()$ is an indicator function that returns 1 if the imputed ($\hat{g}$) and true ($g$) genotypes are equal, 0 otherwise. We obtained four accuracy measures for alfalfa and three for rice.\\
For each imputation method, the total computation time needed to complete the analyses was measured as an indicator of their relative performance. The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution. In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.

\subsection{Software}
\label{sec:software}
Data handling, editing and preparation, summary of results and plots were all performed using the open-source environment for statistical programming R \cite{r_core_team_r:_2014}. All imputation methods except Beagle and FILLIN were implemented in R: MNI directly in base R; KNNI using the function \emph{knncatimpute()} from the Scrime package \cite{schwender_scrime:_2013}; SVDI with the R package ``bcv'' \cite{perry_bcv:_2009}; RFI using the ``MissForest'' \cite{stekhoven_missforestnon-parametric_2012} R package (with parameters: ntree=100, maxiter=10, parallelize='variables'). The ``Beagle'' localized haplotype clustering imputation method was implemented with the Beagle software version 3.3.2 \cite{browning_rapid_2007}. The FILLIN algorithm \cite{swarts_novel_2014} is implemented using the Tassel CLI plugin \emph{FILLINFindHaplotypesPlugin} followed by \emph{FILLINImputationPlugin}. The latter allowed accuracy measurements through \emph{-accuracy} and \emph{-propSitesMask} options.\\
The Bowtie 2 tool \cite{langmead_fast_2012} was used to query the consensus sequence of each alfalfa tag pair containing a SNP against the \emph{M. truncatula} reference genome Version 4.0 using the \emph{-verysensitivelocal} preset. The BWA alignment tool \cite{li2009fast} was used to align rice tags on Rice Genome Annotation Project Release 7.

\makeatletter{}\section{Results}
\label{sec:results}

\subsection{Genotypes}
\label{sec:genotypes}  
GBS detected 32\,706, 40\,734 and 166\,418 SNP loci in Alfalfa-PV, Alfalfa-Med and Rice, respectively (Table~\ref{tab:descriptive_statistics}). The overall missing rate was $66.6\%$, $59.6\%$ and $53.4\%$ in the three datasets. Supplementary Figure S1 shows the density distribution of missing rate per marker in the complete datasets. The amount of missing marker genotypes varied with the threshold of maximum per-marker missing-rate allowed in the data (10\%, 20\%, 40\% and 70\%) and the proportion of artificial missing genotypes that were introduced (1\%, 5\%, 10\% and 20\%). Missing data point counts ranged from a minimum of 3\,878 in Alfalfa-PV with 10\% allowed and 1\% artificial missing genotypes, to a maximum of 5\,580\,616 in Rice with 70\% allowed and 20\% artificial missing genotypes. Over all datasets and allowed/artificial missing genotypes thresholds, the amount of missing data points to be imputed averaged 249\,405.\\
Minor Allele Frequency (MAF) was 0.172, 0.170 and 0.140 in Alfalfa-PV, Alfalfa-Med and Rice, respectively. The three datasets were therefore unbalanced with respect to genotype classes: Rice had the lowest MAF, while Alfalfa-PV and Alfalfa-Med had the smallest minor classes (minor homozygotes 3.5\% and 3.2\% of the total genotypes). Figure~\ref{fig:genotype_classes} reports the proportion of genotypes in each class in the complete datasets (100\%) and as a function of the maximum allowed missing rate per marker. The relative proportion of genotype classes remained approximately stable in rice. In contrast, the heterozygous class in alfalfa tended to become smaller with increasing proportion of allowed genotypes. This reflected the different GBS calling criteria for homozygous and heterozygous SNP: heterozygous SNPs required fewer overlapping reads to be called, while the implemented quality filters on alfalfa required a larger number of reads to call a homozygous locus. Therefore homozygous loci tended to have higher missing rates, and to
be included progressively more frequently with more relaxed thresholds on allowed missing level.
The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.\\
Alignment on reference genomes assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively, reflecting the difference between having a reference genome available for the exact same species (rice) or only for a close one (alfalfa).

dsahe different proportion of unaligned markers directly reflect the .

\begin{figure}
\includegraphics[width=0.95\textwidth]{Fig01_class_breakdown_all.pdf}
\caption[Proportions of genotype classes]{​Proportion of genotype classes for the different missing rate thresholds (0.1, 0.2, 0.4, 0.7 and 1 -the complete dataset), in Alfalfa-PV, Alfalfa-Med and Rice (all chromosomes and chromosome 1 only). Red bars represents the most common homozygous (AA), blue bars the least common homozygous (BB) and green bars the heterozygous (AB).}
\label{fig:genotype_classes}
\end{figure}

\subsection{Imputation accuracy}
\label{sec:imputation_accuracy}

\makeatletter{}\begin{table}
\centering
\caption[Accuracy statistics]{Imputation accuracies (as percentages) for the various
datasets and methods, averaged over all allowed and injected missing thresholds. Rice data are reported averaged by chromosome and on the complete genotypes. Alfalfa data are reported averaged over the two Alfalfa-PV and Alfalfa-Med populations. FILLIN failed to impute alfalfa datasets.}
\label{tab:accuracy_statistics}
\begin{tabular}{llcccc}
\hline\noalign{\smallskip}
\noalign{\smallskip}\hline\noalign{\smallskip}
\textbf{Dataset} & \textbf{Method} & \textbf{Total} & \textbf{Minor (BB)} & \textbf{Major (AA)} & \textbf{Heterozygote (AB)}\\
\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
alfalfa & Beagle & 73.14 & 14.27 & 99.37 & 20.77\\
alfalfa & Beagle (shuffle) & 66.34 & 1.31 & 99.88 & 0.50\\
alfalfa & kNNI & 82.20 & 14.68 & 90.82 & 66.03\\
alfalfa & MNI & 78.38 & 0.00 & 88.69 & 58.88\\
alfalfa & RFI  & 79.50 & 0.01 & 89.05 & 61.75\\
alfalfa & SVDI & 77.36 & 4.47 & 85.81 & 61.67\\
alfalfa & FILLIN & * & * & * & *\\
alfalfa & FILLIN (shuffle)  & * & * & * & *\\
rice (chr) & Beagle & 99.58 & 97.50 & 99.80 & -\\
rice (chr) & Beagle (shuffle) & 96.09 & 74.73 & 98.42 & -\\
rice (chr) & kNNI & 98.10 & 89.12 & 99.04 & -\\
rice (chr) & MNI & 90.19 & 1.06 & 99.89 & -\\
rice (chr) & RFI  & 98.81 & 91.51 & 99.58 & -\\
rice (chr) & SVDI & 97.81 & 85.64 & 99.11 & -\\
rice (chr) & FILLIN & 86.96 & 63.62 & 90.14 & -\\
rice (chr) & FILLIN (shuffle)  & 82.19 & 55.86 & 85.81 & -\\
rice & Beagle & 99.58 & 97.57 & 99.80 & -\\
rice & Beagle (shuffle) & 93.48 & 55.64 & 97.69 & -\\
rice & kNNI & 98.30 & 89.80 & 99.19 & -\\
rice & MNI & 90.21 & 0.61 & 99.89 & -\\
rice & RFI  & 97.86 & 84.68 & 99.28 & -\\
rice & SVDI & 95.74 & 70.96 & 98.40 & -\\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

Accuracies are reported, averaged, in Table~\ref{tab:accuracy_statistics}. In rice, the average imputation accuracy over all genotype classes was above 80\% for all methods. No significant difference was found between chromosomes. Thus average accuracy over all chromosomes is therefore presented (Figure~\ref{fig:accuracy_rice}). \\
The overall accuracy for alfalfa, averaged across the two datasets, was 12-25\% lower than rice, ranging between 66\% and 82\%. No significant differences were found between the two datasets. Thus average accuracy over the Alfalfa-Med and Alfalfa-PV is presented (Figure~\ref{fig:accuracy_alfalfa}).

The imputation accuracy varied across the different genotype classes, with the highest accuracy in the most common genotype (averaging 92.27\% and 96.47\% in alfalfa and rice, respectively), while the least common genotype class showed much lower accuracy (averaging 5.79\% and 69.88\% in alfalfa and rice, respectively).
Alfalfa datasets included heterozygous SNP loci, accounting for an average of 27\% of individuals per locus (Table~\ref{tab:descriptive_statistics}). Heterozygotes had an intermediate imputation accuracy ranging from 0.5\% with Beagle to 66.03\% with KNNI. Results for each individual alfalfa population and the 12 individual rice chromosomes are similar to those discussed here (see Supplementary Figures S2 to S15). \\
Missing rate did not affect the imputation accuracy substantially, with most algorithms showing a flat or almost flat response to increased missing rate. KNNI showed a decreasing imputation accuracy with increasing missing rate, both in alfalfa and rice.

Beagle outperformed all other imputation methods in rice, making efficient use of marker position (Figure~\ref{fig:accuracy_rice}). When markers were randomly reshuffled (thus offsetting position information), though, general imputation methods (other than MNI) showed a higher imputation accuracy than Beagle. The accuracy loss with shuffled markers is exacerbated in alfalfa datasets, and in particular in heterozygous and minor homozygous genotypes.\\
Similarly, FILLIN resulted in a 5-8\% accuracy drop when marker positions were shuffled. FILLIN performances were in general lower than other methods due to the fraction of unimputed genotypes, here accounted as imputation errors. On average, FILLIN did not impute 14.6\% of missing data in Rice and practically the entirety of Alfalfa data. Even trying to relax FILLIN parameters, it was impossible to obtain sensible results in alfalfa. Thus, FILLIN accuracy results are not reported in Figure~\ref{fig:accuracy_alfalfa}.

When examining general imputation methods, resulting accuracies highlighted RFI and KNNI as the best ones. The two methods had comparable performances over all experiments. SVDI followed, and MNI ranked last.\\
In Rice, for 70\% allowed missing rate, the imputation accuracy of RFI was still higher than 90\%, while it dropped towards 85\% for KNNI and SVDI. As expected, MNI had an imputation accuracy of 100\% in the major homozygous class, but dropped to 0\% in the minor homozygous class.\\
In alfalfa KNNI showed the highest imputation accuracy over all classes (averaging 82.20\%) and in the heterozygous (66.03\%) and minor homozygous (14.68\%) genotype classes (Figure~\ref{fig:accuracy_alfalfa}). RFI and SVDI were second ranking. They did not differ much from MNI in the overall accuracy (MNI: 78.38\%, RFI: 79.50\%, SVDI: 77.36\%) and in the major (MNI: 88.69\%, RFI: 89.05\%, SVDI 85.81\%) homozygous class, but gave higher imputation accuracy in the heterozygous (MNI: 58.88\%, RFI: 61.75\%, SVDI 61.67\%). In the minor homozygous class only SVDI resulted in some sporadic correct imputation, while RFI and MNI did not produce any sensible result (MNI: 0\%, RFI: 0.01\%, SVDI 4.47\%) classes.

\begin{figure}
\includegraphics[width=0.95\textwidth]{figure_rice_chrom.pdf}
\caption[Rice imputation accuracies]{
imputation accuracies overall, for the major homozygous genotype (AA), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged over the 12 rice chromosomes. Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black), Beagle with unordered markers (dashed black), FILLIN with ordered markers (purple) and FILLIN with unordered markers (dashed purple). Y axis scale changes to highlight differences.}
\label{fig:accuracy_rice}
\end{figure}

\begin{figure}
\includegraphics[width=0.95\textwidth]{figure_alfalfa.pdf}
\caption[Alfalfa imputation accuracies]{
imputation accuracies overall, for the major homozygous genotype (AA), for heterozygotes (AB), and for the minor homozygous genotype (BB) in datasets consisting of 10\%, 20\%, 40\% and 70\% allowed missing data per locus (boxes) with 1\%, 5\%, 10\% and 20\% additional missing values artificially introduced (x-axis) averaged two alfalfa populations (Alfalfa-Med and Alfalfa-PV). Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black). FILLIN was unable to impute alfalfa data and is absent from figure. Y axis scale changes to highlight differences.}
\label{fig:accuracy_alfalfa}
\end{figure}

\subsection{Computation time}
\label{sec:computation_time}  
The amount of time required to complete the imputation process was recorded for each method. Only the implementation of RFI could leverage a multi-core/multi-thread environment, so that RFI computation times should be evaluated considering 10 CPUs used in parallel, while all other algorithms used only one CPU at a time.\\
Imputation efficiency (Figure~\ref{fig:time_analysis}) was assessed with respect to the gross dimension of the data set (i.e., number of markers $\times$ number of samples). An alternative analysis relating computation times to the number of missing genotypes brought to the same results (data not shown).\\
RFI required by far the longest computation times (in spite of parallelization), which grew approximately exponentially ($O(e^N)$) with problem size, and easily required tens of hours for individual rice chromosomes, and hundred of hours (up to 937 hours) for the complete rice dataset. The second slowest algorithm was KNNI, with computation times growing approximately quadratically ($O(N^2)$) with problem complexity (the chosen KNNI implementation contains no heuristic method to prune the all vs. all distance calculation). KNNI was however significantly faster than RFI (on average 16.7 times faster), requiring a time ranging from 20.68 seconds (rice chromosome 7, 10\% allowed, 1\% artificial) to about 28 hours (rice complete dataset, 70\% allowed, 20\% artificial) to complete the imputation task.\\
All other algorithms were faster, with computation times growing linearly or logarithmically with problem size. Beagle, FILLIN and SVDI resulted in similar execution times. MNI was by far the fastest imputation algorithm, being scarcely affected by the size of the problem (computation times ranging from 0.23 to 31.29 seconds).

\begin{figure}
\includegraphics[width=0.95\textwidth]{time_vs_size.pdf}
\caption[Computation times]{computation time as a function of the total size of the imputed dataset.
Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (light blue), RFI (green), Beagle (black) and FILLIN (purple).
Plots include measures on Alfalfa-Med population, Alfalfa-PV population, and rice chromosomes 1 to 12. Complete rice datasets are omitted for readibility.}
\label{fig:time_analysis}
\end{figure}

\makeatletter{}\section{Discussion}
\label{sec:discussion}

\subsection{Minor Allele Frequency and data (un)balancedness}
\label{sec:maf}
GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain. On our data sets we found missing rates varied from 53\% to 67\%. With larger amounts of missing data the complexity of the imputation problem increases. Imputation errors can have a negative impact on successive analyses (e.g. on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points. Classification problems are known to be harder when data are unbalanced, i.e. the classes appear at different frequencies in the datasets, with typically one over-represented class
(see~\cite{kotsiantis_handling_2006}
and~\cite{sun_classification_2009} for a review). In SNP genotype imputation, data balancedness is directly related to the minor allele frequency (MAF). In the classification of unbalanced observations, it is important to look not only at the total classification accuracy, but also at the per-class accuracy. The total classification accuracy may be misleading, being ``dominated'' by the majority class
\cite{he_learning_2009}.
Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans). Our per-class dissection of results allowed a deeper insight into the imputation process. Indeed, all imputation algorithms performed considerably better in the majority class rather than in the less frequent classes.
In alfalfa, KNNI was easily the best imputation method in the heterozygous and minor homozygous classes, while SVDI and RFI performed only slightly better than MNI. In rice, Beagle gave the best imputation results, with accuracy close to 100\% both in the major and minor homozygous classes.

\subsection{Missing rate and the ``curse of dimensionality''}
\label{sec:missing_rate_curse_dimensionality}
In general, imputation methods were relatively robust to increasing missing rates. Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates. The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
KNNI's decrease in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed. 
On average, the side $l$ of the hypercube to include $k$ neighbors is a function of $k$, $n$ (sample size) and $p$ (number of parameters): $l = \left ( \frac{k}{n} \right )^{1/p}$. With $n$ constant, the hypercube in which the $k$ neighbors lie gets bigger as $p$ increases. This holds especially for local learning methods (e.g.,~K-Nearest Neighbors methods, local regression) which rely heavily on the the information content of the neighborhood. RFI - which only partially relies on local structure of the data - suffered marginally from high missingness. SVD and MNI, which build on more general properties of the data, and Beagle - whose learning process is fundamentally different and specific for genotype imputation - were scarcely affected by missing rates.

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling. This semplification step made alfalfa data less adherent to the underlying biology;\\
iii) rice data have higher marker density (2.4~Kbp/SNP, compared to 24.5~Kbp/SNP for Alfalfa-PV and 19.6~Kbp/SNP for Alfalfa-Med) due to both a higher number of markers and a smaller genome (400~Mb for rice, 800~Mb for Alfalfa). This allowed for higher average values of linkage disequilibrium (LD) among SNP markers, thus facilitating the imputation process;\\
iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on a different species.

There was no population structure in alfalfa \cite{Annicchiarico2015}, while the rice dataset was intrinsically stratified ---being a collection of five subpopulations (\textit{indica}, \textit{temperate japonica} \textit{tropical japonica}, \textit{aus} and \textit{aromatic}). Most imputation methods implicitly exploit population structure (e.g. KNNI computes distances based on genetic relatedness; Beagle reconstructs haplotypes based on genetic similarities), without explicitly modelling it. The imputation of missing genotypes, however, is not an inferential problem such as GWAS (genome-wide association study), where the significance and estimate of SNP effects are known to be inflated if population structure is not included in the model, and the impact on the accuracy of imputation is likely to be small. In cattle from Scandinavian countries, Brondum et al \cite{brondum2012short} actually found higher imputation accuracy when combining populations (cattle breeds) in Beagle (without explicitly modelling the population structure), most likely as a result of the larger sample size. On the other hand, population stratification may help explain why FILLIN performed poorly in our rice dataset, which is a collection of subpopulations, while this algorithm is optimised for homogeneous inbred populations. 
However, the influence of population structure on the accuracy of imputation was not formally tested in this work, and this remains an interesting topic for further investigation.

\subsection{Reference genome assembly and ordered vs unordered markers}
\label{sec:reference_genome_ordered_vs_unordered}
When a reference genome is available, marker position can be used in the imputation process. All methods specifically developed for the imputation of missing genotypes have been designed for markers ordered along the genome sequence, and make explicit or implicit use of position information.

In rice, where a reference genome is available, we could assess the effect of using ordered vs. unordered markers for imputation. For alfalfa there is no reference genome sequenced yet. The genome of the close relative diploid \emph{Medicago Truncatula} is available \cite{young_medicago_2011} and can in principle be used. However, while the two genomes show high synteny \cite{li_saturated_2014}, aligning on a different species (and with different ploidity) comes at the price of discarding those markers that do not align. In our case only 57.54\% (Alfalfa-Med) and 57.86\% (Alfalfa-PV) aligned on the \emph{M. Truncatula} genome, compared to the 88.66\% of rice markers aligning on \emph{O. Sativa} genome. This left with 23\,438 and 18\,923 SNPs to be used for analysis with ordered markers. \\
Among the imputation methods used in this study, Beagle and FILLIN are the only ones that make use of marker information, and were therefore tested with ordered and randomly shuffled markers. In rice dataset with ordered markers Beagle showed astounding resilience to high missingness, with imputation accuracy in the minority class $>$95\% even in the most extreme scenarios (70\% allowed, 20\% artificial missing genotypes).
Beagle worked substantially worse when the marker position was randomly reshuffled, with imputation accuracy in the minority class ranging from 75\% (10\% allowed, 1\% artificial missing rates) to almost 50\% (70\% allowed, 20\% artificial missing rates). When tested with alfalfa Beagle largerly overestimated the majority class and had very poor performances on the heterozygous class. Shuffling marker positions further worsened Beagle performance in alfalfa, and almost all missing genotypes were assigned to the majority class.\\
FILLIN performance in rice followed the same pattern and was substantially worse when markers were shuffled. This result confirmed that imputation methods specifically designed for ordered markers are not an option for species lacking a reference genome.

Previous studies, though not detailing a per-class breakdown of the imputation accuracy, provide interesting comparisons. In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\% to 80.9\% with 1\% to 20\% missing genotypes. Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range and finding Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI. 
Finally, Swarts et al.~\cite{swarts_novel_2014} used ordered GBS maize data to compare Beagle with the FSFHap and FILLIN algorithms they developed. They mostly obtained high imputation accuracies in the three genotype classes, providing further evidence of the added value of having a reference genome assembly.

\subsection{Size of the imputation problem}
\label{sec:size_of_problem}
Albeit this was not the main objective of the paper, and no formal effort to optimise the implementation of the various imputation methods was done, still the recorded computation times give us some interesting information.
The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.

\makeatletter{}\section{Conclusions}
\label{sec:conclusions}
Imputation of missing genotypes can be an effective technique also for GBS data.
The most accurate imputation methods resulted in a total amount of wrongly imputed genotypes near to zero in rice and slightly over 10\% in alfalfa. 
The proportion of imputation errors, however, varied dramatically among genotype classes, approaching very high levels with the worst methods on the minor homozygous class.
The structure of the genome and the maturity of the reference assembly play a role in the imputation efficiency, as indicated by the greater efficiency obtained in rice compared with alfalfa. While Beagle was preferable for rice, general data imputation methods performed significantly better in the absence of a reference genome. In particular, in alfalfa RFI and KNNI showed the lowest imputation errors in all classes, with KNNI to be preferred for computational efficiency. The alignment of markers to closely related species can help, but comes at the price of discarding markers that do not align, and with general methods still performing better.\\
The imputation accuracy tended to be relatively robust over increasing missing rates. However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''. In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\

The results of this paper showed that high imputation accuracies can be achieved with GBS data, and that general imputation methods are a valid option when the reference genome is not available or the alignment on the genome of a closely related species leads to losing too many markers. Additionally, we highlighted the importance of examining imputation accuracy in the different genotype classes, since the common practice of summarizing performances in a single index can be very misleading when data are unbalanced (i.e. MAF is low).
Exploring ways to improve imputation accuracy in GBS data is an important research area and can help making genotyping-by-sequencing a very attractive and cost-effective technique for genomic applications in species with and without a reference genome.

\makeatletter{}\begin{acknowledgements}
The rice data used in this research paper were produced within the framework of the Italian national project ``RISINNOVA'' (grant n. 2010-2369), financially supported by the AGER Foundation..
The creation of the alfalfa data sets was funded by the projects ‘Genomic selection in alfalfa (GENALFA)’ funded by the Italian Ministry of Foreign Affairs and International Cooperation in the framework of the Italy-USA scientific cooperation program, the Italian share of the FP7-ArimNet project ‘Resilient, water- and energy-efficient forage and feed crops for Mediterranean agricultural systems (REFORMA)’ funded by the Italian Ministry of Agricultural and Forestry Policies.
\end{acknowledgements}

\bibliographystyle{spmpsci}      
\bibliography{inputs/nazzicari.bib}

\end{document}

### 2016-05-04 Nazzicari_Imputation.tex
aggregato

@@
 correctly imputed genotypes, both as a total and within genotype classes 
-(major/minor homozigotes, and heterozigotes, when present).
+(major/minor homozygotes, and heterozygotes, when present).
 
@@
 
-The 715 samples from alfalfa (Alfalfa-Med and Alfalfa-PV) and rice (Rice) populations were genotyped using the genotyping-by-sequencing (GBS) approach \cite{elshire_robust_2011}.
-Different GBS protocols were used to genotype the alfalfa and rice samples, since genotyping was carried out under different projects.\\
+The 715 alfalfa (Alfalfa-Med and Alfalfa-PV) and rice (Rice) samples were genotyped using the genotyping-by-sequencing (GBS) approach \cite{elshire_robust_2011}.
+Different GBS protocols were used to genotype alfalfa and rice, since genotyping was carried out under different projects.\\
 In alfalfa, DNA was isolated from fresh leaf tissues by the Wizard~\textregistered~Genomic DNA Purification Kit (Promega, A1125) and quantified with a Quant-iT PicoGreen dsDNA assay kit (Life Technologies, P7589).
@@
 We considered six imputation methods: Mean Value Imputation (MNI), K-Nearest Neighbors Imputation (KNNI), Random Forest Imputation (RFI), Singular Value Decomposition Imputation (SVDI), Localized Haplotype Clustering Imputation (``Beagle'') and haplotype reconstruction around the recombination sites (FILLIN from Tassel suite).
-For all algorithms we imputed a $m\times n$ matrix of m individuals and n markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.
+For all algorithms we imputed a $n\times m$ matrix of n individuals and m markers whose data points, defined in the set \{0,1,2,NA\}, represented the three possible genotypes (AA, AB, and BB) and the missing value, respectively.
 \paragraph{MNI}
@@
 : in K-Nearest Neighbors Imputation missing data points were imputed based on the weighted average of the K closest markers \cite{troyanskaya_missing_2001} defined by the Simple Matching Coefficient distance function \cite{schwender_statistical_2007}, specifically designed for categorical data.
-$K=4$ was used in KNNI, and neighbo
-rs values were weighted by the reciprocal of their distance from the data point to be imputed.
+$K=4$ was used in KNNI, and neighbors values were weighted by the reciprocal of their distance from the data point to be imputed.
 \paragraph{SVDI}
@@
 The final imputed data points were then discretized to the closer genotype in \{0,1,2\}.
-A value of $k=4$ for the eigenvectors to be selected in $U^*$ was used in our implementation of SVDI, based on empirical results in the range $3--20$ (data not shown).
+A value of $k=4$ for the eigenvectors to be selected in $U^*$ was used in our implementation of SVDI, based on empirical results in the range 3--20 (data not shown).
 Additional details on SVDI can be found in \cite{troyanskaya_missing_2001}
@@
 : in Random Forest (RF) imputation, missing genotypes at marker j were imputed by means of RF multiple regression trees \cite{breiman_random_2001} where all markers other than j were used for the prediction.
-At each marker j, 100 RF regression trees $\Theta_{1} \ldots \Theta_{100}$ were grown from a bootstrapped sample of the individuals in Y and a random subset of $\sqrt{m-1}$ markers.
-Missing genotypes were imputed averaging predictions over the 100 RF trees
+At each marker j, 100 RF regression trees $\Theta_{1} \ldots \Theta_{100}$ were grown from a bootstrapped sample of the individuals in Y and a random subset $x$ of $\sqrt{m-1}$ markers.
+Missing genotypes were imputed averaging predictions over the 100 RF trees:
 
@@
 These were then imputed with the various algorithms, measuring the resulting imputation accuracy and computation time.
-We extracted four datasets from the overall data that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker.
+From the overall data, we extracted four datasets that had a maximum of 10\%, 20\%, 40\% or 70\% missing data per marker.
 For each of these datasets, we randomly introduced an additional 1\%, 5\%, 10\% or 20\% missing genotypes, on which imputation accuracy could be measured.
 Table~\ref{tab:missing_rates} reports the number of markers and average missing rate (before introducing artificial missing genotypes) for the four call-rate thresholds.
-This procedure produced 16 combinations for each population: 4 call rate datasets $\times$ 4 levels of artificial missing markers.
+This procedure produced 16 combinations for each population: 4 call rates $\times$ 4 levels of artificial missing markers.
 On each combination, we applied the 6 described imputation methods.
 Rice data were analysed as a whole and by each of the 12 chromosomes separately.
-In absence of alfalfa genome, the \emph{M.
-Truncatula} genome was used as reference for those algorithms (Beagle and FILLIN) requiring markers positions.
+In absence of the alfalfa genome, the \emph{M.
+Truncatula} genome was used as reference for those algorithms (Beagle and FILLIN) requiring marker position.
 
@@
 The rice dataset contained only two genotype classes: AA (homozygous for the major allele) and BB (homozygous for the minor allele).
-Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).\\
+Both alfalfa datasets contained also the third genotype classes AB (heterozygous, pooling the three possible heterozygotes AAAB, AABB, ABBB).
+
 For each experiment we measured the overall imputation accuracy and the imputation accuracy within each genotype class.
@@
 The measured time equals the total dedicated CPU time in case of single thread execution, and corresponds to a fraction of it in case of algorithm supporting multi thread execution.
-In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs, thus all the reported times are to be considered as time elapsed while employing ten times more computational resources than the other algorithms.\\
-To ensure consistency in the experimental conditions, all analysis were run on the same computing platform: a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.
+In our experiments only RFI implemented parallel execution, and it was configured to use 10 CPUs: thus, all RFI reported times are to be considered as time elapsed while employing ten times more computational resources compared to the other algorithms.
+
+To ensure consistency in the experimental conditions, all analysis were run on the same computing platform at PTP Science Park (www.ptp.it): a multi-node cluster with 64 CPUs AMD Opteron(tm) 2400 MHz and 256GB RAM for each node.
 

### 2016-05-04 redo_pdf.sh
update to compiled joint tex source

@@ -22,3 +22,16 @@ pdflatex $FILE
 bibtex $FILE
 pdflatex $FILE
 pdflatex $FILE
+
+#DOING MAIN WITH ALL INPUTS IN ONE FILE
+FILE=Nazzicari_Imputation
+
+#cleaning all LaTeX clutter
+rm -f $FILE.aux $FILE.dvi $FILE.log $FILE.idx $FILE.ind $FILE.toc $FILE.ilg $FILE.out $FILE.bbl $FILE.blg $FILE.nav $FILE.snm $FILE.ps $FILE.pdf
+
+#doing the usual latex routine
+pdflatex $FILE
+bibtex $FILE
+pdflatex $FILE
+pdflatex $FILE
+

### 2016-05-04 redo_pdf.sh
updates to have all supplementary in a single joint file

@@ -35,3 +35,15 @@ bibtex $FILE
 pdflatex $FILE
 pdflatex $FILE
 
+#DOING SUPPL WITH ALL INPUTS IN ONE FILE
+FILE=Nazzicari_Imputation_suppl
+
+#cleaning all LaTeX clutter
+rm -f $FILE.aux $FILE.dvi $FILE.log $FILE.idx $FILE.ind $FILE.toc $FILE.ilg $FILE.out $FILE.bbl $FILE.blg $FILE.nav $FILE.snm $FILE.ps $FILE.pdf
+
+#doing the usual latex routine
+pdflatex $FILE
+bibtex $FILE
+pdflatex $FILE
+pdflatex $FILE
+

### 2016-05-04 Nazzicari_Imputation_suppl.tex
supplementary in joint file

--- 2016-02-11 inputs/supplementary.tex
@@
-\section{Supplementary material}
+\begin{filecontents*}{example.eps}
+gsave
+newpath
+  20 20 moveto
+  20 220 lineto
+  220 220 lineto
+  220 20 lineto
+closepath
+2 setlinewidth
+gsave
+  .4 setgray fill
+grestore
+stroke
+grestore
+\end{filecontents*}
+\RequirePackage{fix-cm}
+\documentclass[smallextended]{svjour3}       \smartqed  
+\usepackage{xfrac}   
+\usepackage{lineno}  
+\usepackage{graphicx}
+\usepackage{color}
+\usepackage{multirow}
+\usepackage{array}
+\usepackage{setspace} 
+\usepackage{makecell}
+\usepackage{amsmath}
+
+\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+
+\renewcommand{\figurename}{Suppl. Fig.}
+\renewcommand{\tablename}{Suppl. Table}
+
+\linenumbers
+\doublespacing 
+\journalname{Molecular Breeding}
+\begin{document}
+
+\title{
+Marker imputation efficiency for Genotyping-By-Sequencing data of crop species with and without a reference genome 
+\thanks{Nelson Nazzicari and Filippo Biscarini contributed equally to the work \\
+Corresponding author: Nelson Nazzicari \email{nelson.nazzicari@crea.gov.it}}
+}
+
+\titlerunning{Imputation efficiency for GBS data with and without the reference genome}
+
+\author{Nelson Nazzicari \and
+  Filippo Biscarini  \and
+  Paolo Cozzi \and
+  E. Charles Brummer \and
+  Paolo Annicchiarico
+}
+
+\authorrunning{Nazzicari, Biscarini et al.} 
+\institute{
+    N. Nazzicari, P. Annicchiarico \at Council for Agricultural Research and Economics (CREA) Research Centre for Fodder Crops and Dairy Productions, Lodi (Italy)
+\and
+    F. Biscarini, P. Cozzi \at Dipartimento di Bioinformatica, Fondazione Parco Tecnologico Padano, Lodi (Italy)
+\and
+    E.C. Brummer \at University of California, Plant Sciences Department, Davis, CA (USA)
+}
+
+\date{Received:  / Accepted:}
+
+\maketitle
+
+\graphicspath{{./figures/}}
+\makeatletter{}\section{Supplementary material}
 \label{sec:supplementary_material}
@@
 In the presented study we thus applied a threshold of 4 required reads per stack to call a homozigote to minimize this kind of error.
+
+\subsection{Supplementary tables}
+\label{sec:supplementary_tables}
+\makeatletter{}\begin{table}
+\centering
+\caption[Accuracy statistics]{Imputation accuracies (as percentages) for the various
+datasets and methods, averaged over all allowed and injected missing thresholds.
+Rice data are reported averaged by chromosome and on the complete genotypes.
+Alfalfa data are reported averaged over the two Alfalfa-PV and Alfalfa-Med populations.
+FILLIN failed to impute alfalfa datasets.}
+\label{tab:accuracy_statistics}
+\begin{tabular}{llcccc}
+\hline\noalign{\smallskip}
+\noalign{\smallskip}\hline\noalign{\smallskip}
+\textbf{Dataset} & \textbf{Method} & \textbf{Total} & \textbf{Minor (BB)} & \textbf{Major (AA)} & \textbf{Heterozygote (AB)}\\
+\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
+alfalfa & Beagle & 73.14 & 14.27 & 99.37 & 20.77\\
+alfalfa & Beagle (shuffle) & 66.34 & 1.31 & 99.88 & 0.50\\
+alfalfa & kNNI & 82.20 & 14.68 & 90.82 & 66.03\\
+alfalfa & MNI & 78.38 & 0.00 & 88.69 & 58.88\\
+alfalfa & RFI  & 79.50 & 0.01 & 89.05 & 61.75\\
+alfalfa & SVDI & 77.36 & 4.47 & 85.81 & 61.67\\
+alfalfa & FILLIN & * & * & * & *\\
+alfalfa & FILLIN (shuffle)  & * & * & * & *\\
+rice (chr) & Beagle & 99.58 & 97.50 & 99.80 & -\\
+rice (chr) & Beagle (shuffle) & 96.09 & 74.73 & 98.42 & -\\
+rice (chr) & kNNI & 98.10 & 89.12 & 99.04 & -\\
+rice (chr) & MNI & 90.19 & 1.06 & 99.89 & -\\
+rice (chr) & RFI  & 98.81 & 91.51 & 99.58 & -\\
+rice (chr) & SVDI & 97.81 & 85.64 & 99.11 & -\\
+rice (chr) & FILLIN & 86.96 & 63.62 & 90.14 & -\\
+rice (chr) & FILLIN (shuffle)  & 82.19 & 55.86 & 85.81 & -\\
+rice & Beagle & 99.58 & 97.57 & 99.80 & -\\
+rice & Beagle (shuffle) & 93.48 & 55.64 & 97.69 & -\\
+rice & kNNI & 98.30 & 89.80 & 99.19 & -\\
+rice & MNI & 90.21 & 0.61 & 99.89 & -\\
+rice & RFI  & 97.86 & 84.68 & 99.28 & -\\
+rice & SVDI & 95.74 & 70.96 & 98.40 & -\\
+\noalign{\smallskip}\hline
+\end{tabular}
+\end{table}
 
@@
 Lines colors represent the five imputation algorithms: MNI (salmon), KNNI (red), SVDI (blue), RFI (green), Beagle with ordered markers (solid black) and Beagle with unordered markers (dashed black)}\end{figure}
+
+\bibliographystyle{spmpsci}
+
+\end{document}

### 2016-05-04 Nazzicari_Imputation.tex
aggregato

@@
 The average missing-rate and MAF in the 12 rice chromosomes were comparable to the whole-rice dataset.\\
-Alignment on reference genomes assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively, reflecting the difference between having a reference genome available for the exact same species (rice) or only for a close one (alfalfa).
-
-dsahe different proportion of unaligned markers directly reflect the .
+Alignment on reference genomes assigned a position to 88.66\%, 57.54\% and 57.86\% of Rice, Alfalfa-Med and Alfalfa-PV markers, respectively, reflecting the difference between having a reference genome available (rice) or using that of a closely related species (alfalfa).
 
@@
 On average, FILLIN did not impute 14.6\% of missing data in Rice and practically the entirety of Alfalfa data.
-Even trying to relax FILLIN parameters, it was impossible to obtain sensible results in alfalfa.
+Even trying to relax FILLIN parameters, we couldn't obtain meaningful results in alfalfa.
 Thus, FILLIN accuracy results are not reported in Figure~\ref{fig:accuracy_alfalfa}.
@@
 GBS data pose a greater challenge than SNP-array data to imputation algorithms, mainly as a consequence of the much larger quantity of missing genotypes they contain.
-On our data sets we found missing rates varied from 53\% to 67\%.
+In our data sets we found that missing rates varied from 53\% to 67\%.
 With larger amounts of missing data the complexity of the imputation problem increases.
 Imputation errors can have a negative impact on successive analyses (e.g.
-on genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).\\
+genomic predictions \cite{rutkoski_imputation_2013,Annicchiarico2015}).
+
 The imputation of missing SNP genotypes is a special case of the broader family of classification problems: three-class missing genotypes (AA, AB and BB) at any given SNP locus are classified based on known genotypes at all remaining data points.
@@
 \cite{he_learning_2009}.
-Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present on minority classes. 
+Indeed, we found that for most methods, even when the total classification accuracy was very high, relatively large error rates were present in the minority classes. 
 The dependency of imputation results on MAF has already been acknowledged  (e.g.,~\cite{hickey_factors_2012} in maize; \cite{ma_comparison_2013} in cattle; \cite{pei_analyses_2008} in humans).
@@
 Only KNNI, and to a lesser extent RFI, showed slowly degrading performances at very high missing rates.
-The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the data set.\\
+The divergent response of the different imputation methods to missing rate became apparent only in the most challenging scenarios, when markers with up to 40-70\% missing genotypes were allowed in the dataset.\\
 KNNI's decrease in imputation accuracy with increasing missing rates can be interpreted in relation to the phenomenon known as the ``curse of dimensionality'' \cite{bellman_dynamic_1957,marimont_nearest_1979}: with the same amount of data, the increasing number of parameters - the ``dimensions'' of the problem - increases and complicates the identification of k neighbors which are close enough to the data point to be classified/imputed.
@@
 \cite{the_1000_genomes_project_consortium_integrated_2012}
-for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: \\
+for human data). On the other hand, imputation accuracy in alfalfa was much lower. This can be ascribed to four main factors: \\
 i) the imputation problem was simpler in rice than in alfalfa, since the rice dataset comprised only two genotype classes (AA and BB) instead of alfalfa's three (AA, AB, BB); \\
 ii) rice is natively diploid while alfalfa, autotetraploid, has been rendered diploid ``in silico'' during SNP calling.
-This semplification step made alfalfa data less adherent to the underlying biology;\\
+This simplification step made alfalfa data less adherent to the underlying biology;\\
 iii) rice data have higher marker density (2.4~Kbp/SNP, compared to 24.5~Kbp/SNP for Alfalfa-PV and 19.6~Kbp/SNP for Alfalfa-Med) due to both a higher number of markers and a smaller genome (400~Mb for rice, 800~Mb for Alfalfa).
 This allowed for higher average values of linkage disequilibrium (LD) among SNP markers, thus facilitating the imputation process;\\
-iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on a different species.
+iv) rice markers were aligned on their native genome, while alfalfa markers were aligned on the genome of a different species.
 

### 2016-05-04 Nazzicari_Imputation.tex
aggregato

@@
 \newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
+\newcommand{\specialcell}[2][c]{%
+  \begin{tabular}[#1]{@{}c@{}}#2\end{tabular}}
 
@@
 
-\makeatletter{}\begin{table}
+\begin{table}
 \centering
-\caption[Descriptive statistics]{
+\caption[Descriptive statistics, markers counts, and missing rates]{
 Descriptive statistics of alfalfa (\emph{Medicago sativa L.}) 
 and rice (\emph{Oryza sativa L.}) genotyping. MAF is average minor allele frequency; 
-p(AA), p(AB) and p(BB) are the proportions of AA, AB and BB genotypes.}
-\label{tab:descriptive_statistics}
-\begin{tabular}{lccccccc}
+p(AA), p(AB) and p(BB) are the proportions of AA, AB and BB genotypes.
+Total number of markers and resulting average missing rate for all markers
+and for four allowed thresholds of missing rate per markers.}
+\label{tab:missing_rates}
+\begin{tabular}{C{3cm}rrrr}
 \hline\noalign{\smallskip}
+\specialcell{\textbf{Allowed missing}\\ \textbf{rate per marker}}  & & \textbf{Alfalfa-PV} & \textbf{Alfalfa-Med} & \textbf{Rice}\\
+\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
+\multirow{7}{*}{100\%}  & \textbf{No. samples}  & 124     & 154       & 437\\
+                        & \textbf{MAF}          & 0.1724  & 0.1702    & 0.140 \\
+                        & \textbf{p(AA)}        & 0.690   & 0.691     & 0.860\\
+                        & \textbf{p(AB)}        & 0.274   & 0.276     & -\\
+                        & \textbf{p(BB)}        & 0.035   & 0.032     & 0.140\\
+                        & \textbf{Missing rate} & 0.666   & 0.596     & 0.534\\
+                        & \textbf{Markers}      & 32\,706 & 40\,734   & 166\,418\\    
 \noalign{\smallskip}\hline\noalign{\smallskip}
-\textbf{Dataset} & \textbf{No. samples} & \textbf{No. SNPs} & \textbf{Missing Rate} & \textbf{MAF} & \textbf{p(AA)} & \textbf{p(AB)} & \textbf{p(BB)}\\
-\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
-Alfalfa-PV & 124 & 32\,706 & 0.666 & 0.1724 & 0.690 & 0.274 & 0.035\\
-Alfalfa-Med & 154 & 40\,734 & 0.596 & 0.1702 & 0.691 & 0.276 & 0.032\\
-Rice & 437 & 166\,418 & 0.534 & 0.140 & 0.860 & - & 0.140 \\
+   \multirow{2}{*}{70\%}   & \textbf{Missing rate} &  0.325 & 0.29  & 0.364\\
+                           & \textbf{Markers}      & 13\,190  & 19\,986 & 109\,372\\    
+\noalign{\smallskip}\hline\noalign{\smallskip}
+   \multirow{2}{*}{40\%}   & \textbf{Missing rate} &  0.161 & 0.142 & 0.198\\
+                           & \textbf{Markers}      & 7\,790 & 12\,931 & 58\,553\\    
+\noalign{\smallskip}\hline\noalign{\smallskip}
+   \multirow{2}{*}{20\%}   & \textbf{Missing rate} & 0.076 & 0.074 & 0.099\\
+                           & \textbf{Markers}      & 4\,828 & 8\,962 & 29\,872\\
+\noalign{\smallskip}\hline\noalign{\smallskip}
+   \multirow{2}{*}{10\%}   & \textbf{Missing rate} & 0.046 & 0.045 & 0.049\\
+                           & \textbf{Markers}      & 3\,405 & 6\,364 & 15\,060 \\
 \noalign{\smallskip}\hline
@@
 The reshuffled datasets were tested only on Beagle and FILLIN.
-
-\makeatletter{}
-\begin{table}
-\centering
-\caption[Markers counts and missing rates]{
-Total number of markers and resulting average missing rate for the four different allowed missing rate thresholds in alfalfa (two datasets) and rice.}
-\label{tab:missing_rates}
-\begin{tabular}{C{3cm}rccc}
-\hline\noalign{\smallskip}
-\noalign{\smallskip}\hline\noalign{\smallskip}
-\textbf{Allowed missing rate threshold} &  & \textbf{Alfalfa-PV} & \textbf{Alfalfa-Med} & \textbf{Rice}\\
-\noalign{\smallskip}\Xhline{3\arrayrulewidth}\noalign{\smallskip}
-   \multirow{2}{*}{70\%} & Missing rate &  0.325 & 0.29  & 0.364\\
-                           & Markers      & 13\,190  & 19\,986 & 109\,372\\    
-\noalign{\smallskip}\hline
-   \multirow{2}{*}{40\%} & Missing rate &  0.161 & 0.142 & 0.198\\
-                           & Markers      & 7\,790 & 12\,931 & 58\,553\\    
-\noalign{\smallskip}\hline
-   \multirow{2}{*}{20\%} & Missing rate & 0.076 & 0.074 & 0.099\\
-                           & Markers      & 4\,828 & 8\,962 & 29\,872\\
-\noalign{\smallskip}\hline                           
-   \multirow{2}{*}{10\%} & Missing rate & 0.046 & 0.045 & 0.049\\
-                           & Markers      & 3\,405 & 6\,364 & 15\,060 \\
-\noalign{\smallskip}\hline
-\end{tabular}
-\end{table}
 
@@
 In sugar beet, Biscarini et al.~\cite{biscarini_genome-enabled_2014} used Beagle to impute SNP-chip markers with partial within-chromosome/scaffold alignment, finding global accuracies ranging from 84\% to 80.9\% with 1\% to 20\% missing genotypes.
-Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range and finding Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI.
+Huang et al.~\cite{huang_efficient_2014} tested several algorithms, including Beagle, on simulated ordered GBS rice data with missing rates in the 30-60\% range, and found that Beagle gave imputation accuracy consistently higher (+15-20\%) than KNNI.
 Finally, Swarts et al.~\cite{swarts_novel_2014} used ordered GBS maize data to compare Beagle with the FSFHap and FILLIN algorithms they developed.
@@
 However, KNNI showed a somewhat lower accuracy on high missingness scenarios, probably as a consequence of the ``curse of dimensionality''.
-In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).\\
+In terms of computation requirements, all methods proved to be tractable over all problem complexities with a standard bioinformatics-lab computation infrastructure, except RFI (whose computational requirements increased exponentially with problem size and took up to 40 days with the complete rice dataset and maximum missing rate thresholds).
 
@@
 \makeatletter{}\begin{acknowledgements}
-The rice data used in this research paper were produced within the framework of the Italian national project ``RISINNOVA'' (grant n. 2010-2369), financially supported by the AGER Foundation..
+The rice data used in this research paper were produced within the framework of the Italian national project ``RISINNOVA'' (grant n. 2010-2369), financially supported by the AGER Foundation.
 The creation of the alfalfa data sets was funded by the projects ‘Genomic selection in alfalfa (GENALFA)’ funded by the Italian Ministry of Foreign Affairs and International Cooperation in the framework of the Italy-USA scientific cooperation program, the Italian share of the FP7-ArimNet project ‘Resilient, water- and energy-efficient forage and feed crops for Mediterranean agricultural systems (REFORMA)’ funded by the Italian Ministry of Agricultural and Forestry Policies.
